---
title: "Manscript draft"
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Sara H Knox^4^, Jaclyn
  Hatala Matthes^5^, Joseph G Verfaillie^1^, Daphne J Szutu^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  pdf_document: default
  word_document: default
fontsize: 11pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
csl: global-change-biology.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, Hayward, California, USA

^3^National Ecological Observatory Network, Boulder, Colorado, USA

^4^U.S. Geological Survey, Menlo Park, California, USA

^5^Department of Biological Sciences, Wellesley College, Wellesley, Massachusetts, USA

**Running Head**:

**Keywords**:

**Type of Paper**: Primary research article

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
```

\newpage
# Abstract

# Introduction
Wetlands plays a fundamental role in regulating global climate by storing ~18% of Earth's soil carbon (C) [@lal_carbon_2008], emitting ~30% of global methane (CH~4~) emissions [@kirschke_three_2013; @saunois_global_2016], while only occupying 5-8% of the terrestrial land surface [@mitsch_wetlands_2013]. Most wetlands are greenhouse gas (GHG) sinks over multi-century time scales due to carbon dioxide (CO~2~) sequestration within anoxic soils, though they are commonly greenhouse gas (GHG) sources over shorter time scales due to the high global warming potential of emitted CH~4~ [@whiting_greenhouse_2001; @mitsch_wetlands_2013; @petrescu_uncertain_2015]. Wetland C sequestration and CH~4~ emissions are inextricably linked, as anoxic soil conditions inhibiting ecosystem respiration (*ER*) also activates archaeal CH~4~ production. The balance between these opposing fluxes primarily determines wetland GHG balances [@petrescu_uncertain_2015], and small differences in ecosystem CH~4~ emissions can induce large changes in wetland GHG exchange due to the > 28-fold radiative forcing of CH~4~ compared to CO~2~ over decadal to centennial time horizons [@pachauri_climate_2015]. Recent rises in atmospheric CH~4~ have been attributed to a global increase in wetland CH~4~ fluxes [@nisbet_rising_2016], and wetlands are currently the most uncertain component of the global CH~4~ budget [@kirschke_three_2013]. A key driver of this uncertainty is a lack of ecosystem-scale flux measurements to better constrain models and improve our understanding of how process-level controls scale to whole ecosystem CH~4~ emissions [@bridgham_methane_2013; @saunois_global_2016]. 
<!-- It is therefore important to target ecosystem flux measurement campaigns across environmental gradients known to influence CH~4~ fluxes from small-scale field and laboratory studies. -->

Methane fluxes from wetlands are governed by a number of interacting biophysical drivers including temperature, C inputs, alternative electron acceptor pools, and ecosystem inundation [@bridgham_methane_2013]. It is uncertain how these governing factors interact and scale to whole ecosystem fluxes due to sparse CH~4~ flux measurements across time and space [@petrescu_uncertain_2015], and current ecosystem measurement campaigns may under-sample relevant environmental gradients. Chamber flux syntheses have demonstrated that water table depth, temperature, vegetation, disturbance, and wetland type are important modulators of wetland CH~4~ flux [@turetsky_synthesis_2014]. A similar understanding of CH~4~ flux controls is emerging from eddy covariance studies, where temperature [@rinne_annual_2007; @olson_interannual_2013; @chu_net_2014], recent C inputs [@hatala_gross_2012; @morin_environmental_2014], wetland structure [@matthes_parsing_2014], and water table depth [@brown_evidence_2014; @goodrich_overriding_2015; @chamberlain_underlying_2015; @chamberlain_influence_2016; @sturtevant_identifying_2016; @chamberlain_impact_2017] have been identified as dominant controls across many wetland ecosystems. Small-scale field and laboratory studies have also demonstrated the influence of soil C content [@bridgham_methane_2013; @ye_soil_2016] and alternative electron acceptor pools, such as sulfate and ferric iron (Fe^3+^), on wetland CH~4~ fluxes [@laanbroek_methane_2010; @poffenbarger_salinity_2011; @bridgham_methane_2013]; however, the importance of these factors has not been consistently demonstrated at ecosystem scale. Recently, @holm_ecosystem_2016 and @krauss_component_2016 observed attenuation of ecosystem CH~4~ fluxes with increasing salinity (i.e. sulfate concentrations), verifying the ecosystem-scale relevance of previous chamber-based observations [@poffenbarger_salinity_2011]. Iron (Fe) presence is also expected to inhibit ecosystem CH~4~ fluxes [@laanbroek_methane_2010], as observed in many laboratory studies [@jackel_suppression_2000; @van_bodegom_direct_2004; @jerman_wetland_2009], but Fe inhibition to ecosystem fluxes has not been documented. Understanding the relevance of such factors may be particularly important to constraining wetland CH~4~ emissions from the tropics, the largest source of the global wetland CH~4~ emissions [@kirschke_three_2013], where high soil iron concentrations influence rates of organic matter decomposition and CH~4~ production in forest soils [@teh_suppression_2008; @dubinsky_tropical_2010; @hall_iron_2013].

Wetland restoration and management programs have increasingly been proposed and implemented as a means to mitigate climate change, and restoration strategies that minimize CH~4~ emissions and maximize CO~2~ uptake will provide the optimum climate benefit. These programs are particularly appealing in coastal ecosystems where CO~2~ sequestration rates are high and CH~4~ flux rates are low [@crooks_mitigating_2011; @conservation_international_blue; @poffenbarger_salinity_2011], and in drained peatlands where large CO~2~ emissions from oxidizing peat can be attenuated by re-flooding the landscape [@hatala_greenhouse_2012; @knox_agricultural_2015; @wilson_multiyear_2016]. A strong knowledge of CH~4~ flux drivers is essential to these programs because the magnitude of CH~4~ emissions may determine if a restored wetland is a GHG source or sink [@knox_agricultural_2015], and these emissions could affect funding of wetland restoration programs if financed through carbon markets [@oikawa_evaluation_2017]. 

The Sacramento-San Joaquin Delta region of California (hereafter referred to as the Delta) is an area of extensive wetland restoration where financing through California's Cap and Trade program is being considered. The Delta is comprised of a network of artificially drained islands reclaimed for agriculture, where the water table is maintained below sea level by active pumping and levees surrounding each island. Drainage began in the 1850's to facilitate agricultural development, and exposure of the highly organic soils has led to substantial land surface subsidence [@deverel_historic_2010] and CO~2~ emissions [@hatala_greenhouse_2012; @knox_agricultural_2015]. Prior to drainage, this region was a 5 billion m^3^ tidal marsh where soil mineral and C content varied substantially across space. A major driver of this soil variation was fluvial deposition from major rivers, which carried mineral alluvium from the Sierra Nevada mountain range to the Delta marshes. Alluvium marshes frequently developed adjacent to peat marsh soils that were spatially segregated from the main river channels. Alluvium soils are most common on islands adjacent to the Sacramento River compared to the central and eastern Delta region where fluvial deposition was less pronounced [@deverel_historic_2010]. These differences in pre-drainage wetland geomophology give rise to a contemporary agricultural landscape where large changes in soil carbon and mineral content occur over small spatial scales [@soilsurvey], but how this edaphic variation influences current GHG fluxes from restored wetlands is unknown. Understanding the importance of these legacy effects is highly desirable, as it would allow for more informed wetland restoration strategies where restoration sites could be chosen to minimize GHG emissions.

The objectives of this study were to (1) determine how wetland CH~4~ and GHG budgets varied between peat and alluvium restored wetlands, (2) identify how relationships between biophysical drivers and CH~4~ fluxes varied across wetland types, (3) determine how soil properties varied across wetland types and which properties most likely caused observed flux differences, and (4) determine whether observed flux differences were economically relevant in a carbon market. We measured ecosystem-scale CH~4~ and CO~2~ fluxes by eddy covariance from three restored wetlands for 3+ years, where two wetlands were restored on peat soils and one wetland was restored on alluvium soil. We then used a combination of wavelet analysis and information theory to identify scale-emergent biophysical drivers of wetland CH~4~ fluxes and how they varied across the three wetlands, a technique previously applied by @sturtevant_identifying_2016 to identify scale-emergent CH~4~ flux controls for the two peat wetlands. We then measured multiple soil properties within the tower footprints, including soil C, nitrogen (N), ferrous iron (Fe^2+^), Fe^3+^, and reducible Fe oxides (Fe~ca~), to identify edaphic drivers of observed flux differences. These tower sites are an ideal network to isolate these controlling factors, as all wetlands sites have a similarly managed hydroperiod, are within 13 km of one another, experience the same climate, and have similar plant community compositions. To further control potential covariates, we used two temporary flux towers to simultaneously measure CH~4~ fluxes from additional locations within a restored wetland complex where age is held constant but soil type varies. Finally, we calculated potential economic benefits of restoring wetlands on peat versus alluvium soils using the observed differences in our eddy covariance fluxes and current carbon market costs used in California's Cap and Trade Program [@carbon_dash].

# Materials and Methods
## Site Description
We measured wetland ecosystem CH~4~ and CO~2~ fluxes using a network of eddy covariance towers in the Sacramento-San Joaquin Delta region of California, USA.  The Delta is located within a Mediterranean climate that experiences hot, dry summers and cooler, rainy winters. The mean annual temperature is 15.1$^\circ$C (1998-2015 average), and the region receives 326 mm of rainfall annually [@knox_biophysical_2016]. All measurement sites were located on Sherman and Twitchell Islands in the northwest Delta. Both islands are adjacent to the Sacramento River and have typically lower carbon soils than central and eastern Delta islands within the San Joaquin River drainage [@deverel_historic_2010]. Sherman and Twitchell Islands are a mosaic of alluvium mollisols and deep histosols [@soilsurvey]. The alluvium mollisols formed via fluvial deposition from major rivers, and histosols formed densely organic soils in areas less disturbed by fluvial input [@deverel_historic_2010]. The dominant mollisol series on these islands are Gazwell and Scribner (Fig. 1), both of which are formed in mixed alluvium, belong to the soil class Cumulic Endoaquolls, and are only found on Delta islands within the Sacramento River drainage [@soilsurvey]. Rindge is the major histosol series found on both islands (Fig. 1) and is widely distributed across the Delta region. Rindge soils belong to the  Typic Haplosaprists class, and are characterized by deep, poorly drained marsh soils formed from decomposed plant organic matter [@soilsurvey]. 

Wetland ecosystem fluxes have been measured for over three years from three core wetland sites, where one site is restored on Scribner series and two sites are restored on Rindge series (Fig. 1). The Rindge series site on Sherman Island (Ameriflux site Us-Myb; N 38.0498, W 121.7650), hereafter referred to as the young peat site, was restored from a former pasture in October 2010, and ecosystem flux measurement began at this time. The Rindge series site on Twitchell Island (Ameriflux site Us-Tw1; N 38.1074, W 121.6469), hereafter referred to as the old peat site, was restored from a degraded agricultural site in 1997 and ecosystem flux measurements began in July 2012. The Scribner series site on Twitchell Island (Ameriflux site US-Tw4; N 38.10275, W 121.64125), hereafter referred to as the alluvium site, was restored in January 2014 on a former corn field and ecosystem flux measurements began at this time. All sites are vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*) and are actively managed to remain inundated year-round. Both the young peat and alluvium sites were constructed to have heterogeneous bathymetry, providing channels of open water and areas of closed vegetation. The old peat site is entirely closed vegetation and the ground surface is saturated tule and cattail detritus, whereas the recently restored young peat and alluvium sites have standing water under the plant canopy as less peat has accumulated compared to the old peat wetland. For a more detailed description the young and old peat sites see @miller_carbon_2011, @matthes_parsing_2014, and @knox_agricultural_2015.

Two temporary towers were installed within the same restored wetland complex as the core alluvium site, where wetland age is held constant and soil type varies. One temporary tower was installed in a Rindge soil (N 38.10607, W 121.64172), and the second tower was installed within the same Scribner soil region (N 38.09876, W 121.64217) as the core alluvium site (Fig. 1). We verified that soil properties at these additional sites were characteristic of peat versus alluvium soils using a soil C dataset measured across 117 locations in this wetland complex from December 2013, prior to re-flooding (data not shown). The temporary towers and the core alluvium tower are installed within the same wetland complex that was restored in January 2014. Temporary sites are also vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*), are actively managed to remain inundated year-round, and are similar in structure to the young peat and core alluvium sites.

![Soil type and tower flux footprints for (a) Sherman and (b) Twitchell Islands in the Sacramento-San Joaquin Delta of California. Tower footprints include wetland area only, and footprint rings, from largest to smallest, correspond to the 90, 85, 80, 70, and 50% flux footprints at each site. Tower footprints were calculated using a two-dimensional analytical model described in @detto_soil_2006. The white footprint rings represent the core tower sites, where the footprint on (a) in Rindge series is the young peat site, the footprint on (b) in Rindge series is the old peat site, and the footprint on (b) in Scribner series is the alluvium site. The yellow footprints on (b) represent the temporary tower sites within the same wetland complex as the core alluvium site.](fig1.png)

## Ecosystem Flux and Environmental Measurements
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat were measured from the three core and two temporary wetland sites by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA). High-frequency data from the alluvium, old peat, young peat, and temporary peat site were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval, and high-frequency data from the temporary alluvium site were logged to a CR1000 datalogger (Campbell Scientific, Logan, Utah, USA) at 10 Hz interval. Measurements took place at a height of 5.44m, 4.64m, 5.36m, 4.70m, and 3.88m for the young peat, old peat, alluvium, temporary alluvium, and temporary peat site, respectively. Corrections were applied to vertical wind speed data prior to flux processing for WindMaster Pro's affected by the 'w bug' that causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto_scaling_2010; @hatala_greenhouse_2012; @knox_agricultural_2015]. Flux corrections and quality control were applied as described in detail in @knox_agricultural_2015, @chamberlain_evaluation_2017, and Eichelmann et al. (*in review*), and included high-frequency data despiking, 2-D coordinate rotations, density corrections, and site-specific friction velocity (*u$^\ast$*) filtering. At the old peat site, we reject fluxes from wind directions 290$^\circ$-240$^\circ$ because fluxes from these directions were from other wetland types; however, we did not apply wind direction filtering to the other sites where flux footprints were more homogeneous. Footprints at all sites were calculated using a two-dimensional analytical model [@detto_soil_2006; @hsieh_approximate_2000]. We then filled all gaps in the half-hourly flux time series using artificial neural networks (ANNs) as described in detail in @knox_biophysical_2016 and Eichelmann et al. (*in review*). Separate ANNs were trained for daytime and nighttime CO~2~ fluxes, and the nighttime CO~2~ flux ANN was used to model *ER* at all times. Gross ecosystem photosynthesis (*GEP*) was then estimated as the difference between the gap-filled CO~2~ flux and modeled *ER* [@baldocchi_does_2015]. More detail on this partitioning strategy can be found in @baldocchi_does_2015 and @knox_biophysical_2016. We did not gap fill or partition fluxes for the temporary tower sites and only present their observed fluxes.

We also measured a number of auxiliary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using shielded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*PA*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component and double-sided net-radiometers (CNR1 or NR Lite; Kipp and Zonen, Delft, Netherlands). We measured soil (*T~s~*) and water (*T~w~*) temperature at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). An in depth description of the measurements made and instruments used at each core tower site can be found in Eichelmann et al. (*in review*).

## Soil Analyses
We measured soil C, N, and Fe content at the young peat, old peat, and alluvium sites in the recently accreted wetland O horizon, hereafter referred to as accreted horizon, and the underlying pre-restoration parent soils, hereafter referred to as the parent horizon. Fifteen samples of both accreted and parent soil were collected at each site across three transects with each sampling location at least 3m apart. The transects were within each tower's flux footprint (Fig. 1), all samples were collected from fully inundated locations, and the accreted sample was collected directly above the underlying parent soil sample. At the young peat and alluvium sites (both < 8 years old), the differentiation between these horizons was clear, as the recently accreted horizon was loose, mucky, and heavily comprised of poorly-decomposed plant matter, whereas the underlying parent horizon was hardpack agricultural soil. At these sites, we collected the accreted horizon by hand and the top 15cm of parent horizon using a sediment core. Our sampling strategy was different at the old peat site because more than 0.5m of O horizon has accumulated since initial restoration. Here, the accreted horizon has accumulated to the water surface, and we collected the top 2 cm of this horizon to represent the most recently accreted peat. We then bore holes through the 0.5-0.7m saturated peat layer to collect the underlying parent soil horizon. Differentiation between these layers was also clear, as the top 0.5-0.7m was comprised of poorly-decomposed plant matter and the underlying horizon was silty soil. 

All soil samples were immediately bagged, and sub-samples were field extracted in 0.5M HCl and 0.2M sodium citrate/0.05M sodium ascorbate solutions to measure HCl-extractable Fe (Fe^2+^ and Fe^3+^) and Fe~ca~ pools, respectively. The HCl extraction solubilizes reactive Fe^3+^ minerals and absorbed/solid Fe^2+^, and is used to quantify oxidized (Fe^3+^) and reduced (Fe^2+^) Fe pools. The low pH of HCl prevents any Fe^2+^ oxidation after the samples are collected [@hall_reducing_2015]. Roughly 3g of sample (dry mass) were immediately placed into the HCl solution, and once back in the lab, samples were vortexed, shaken overnight, and centrifuged for 20 minutes at 1000 rcf. Concentrations of Fe^2+^ and Fe^3+^ were measured colorimetrically using the ferrozine assay [@viollier_ferrozine_2000]. Citrate-ascorbate extractions were used to quantify Fe oxides that are reducible by soil microbes [@hyacinthe_reactive_2006]. Roughly 1.5g of sample (dry mass) were immediately placed into the citrate-ascorbate solutions, and in the lab, samples were vortexed, shaken overnight, and then centrifuged for 20 minutes at 1000 rcf. Fe~ca~ concentrations were then quantified using an inductively coupled plasma optical emission spectrometer (ICP-OPS; Perkin Elmer Optima 5300 DV, Waltham, Massachusetts, USA). We then air dried soils samples at room temperature for the percent C (%C) analysis. Dried subsamples were sieved to 2mm and all major visible roots were removed by hand. Sieved, root-picked samples were then ground to a fine powder and analyzed in duplicate for total C using an elemental analyzer (CE Elantech, Lakewood, New Jersey, USA).

## Data Analysis
### Wavelet and Information Theory Analysis
We used a combination of wavelet decomposition and information theory to (1) isolate major time scales of variation within the continuous CH~4~ flux time series, and (2) identify scale-emergent interactions between CH~4~ fluxes and a number of biophysical drivers. Here, flux and environmental variable time series' were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [@percival_estimation_1995], where we isolate variation occurring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales corresponding to 2^n^ measurement intervals. The hourly time scale was reconstructed from dyadic scales 1-2 (1-2h), diel scale from dyadic scales 3-6 (4h to 1.3 days), and multiday scale from dyadic scales 7-10 (2.7-21.3 days). This approach is explained in greater detail in @sturtevant_identifying_2016, and our analysis used the same approach and processing software described therein.

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information is derived from Shannon entropy (*H*), a measure of uncertainty [@shannon1998mathematical], and *I* quantifies the amount of information shared between two variables. Mutual information and other information theory metrics, such as transfer entropy, are particularly useful for identifying relationships in complex systems because they do not assume linearity or other functional relationships and are capable of identifying asynchronous relationships [@ruddell_applying_2013]. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies of each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables as follows:
$$H_x=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{x,y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Here, we bin our continuous data into discrete states to calculate *H~x~*, *H~y~*, and *H~x,y~*. We follow the same processing methodology in @sturtevant_identifying_2016 for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. Analyses for each year were confined to the growing season (DOY 100-300), and time series border periods were removed to avoid wavelet-induced distortions. We then calculated joint and marginal Shannon entropies using 10 fixed-interval bins for the probability distributions of each variable, and calculated *I* for a number of time lags of *X*. We calculated *I* at lags up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases, CH~4~ flux (*F~CH4~*) was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps described above. A more detailed description of this processing methodology can be found in @sturtevant_identifying_2016, and all wavelet decomposition and entropy calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R^*) where *I* is divided by the Shannon entropy of the Y variable (*F~CH4~*). 

### Statistical Analyses and Data Processing

# Results

```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat19_processing.R") #19yr old peat site
source("R/peat6_processing.R") #6yr old peat site
source("R/alluvium_processing.R") #alluvium site

#merge daily sums to single dataset for core sites
all_daily <- rbind(peat19_daily, peat6_daily); all_daily <- rbind(all_daily, alluvium_daily)
all_daily$date <- as.Date(all_daily$datetime) #data plotting variable

#load temporary tower sites
load("data/mobile_alluvium.Rdata")
load("data/mobile_peat.Rdata")
```

## Weather and Hydrology
The three core wetland sites on Sherman and Twitchell Islands experienced similar meteorologic and hydrologic conditions (Fig. 2). Incoming *PAR* and *T~a~* exhibited strong seasonality, where peak values were observed in summer months, and daily mean values were similar across all three wetland sites (Fig. 2a,b). Water temperature exhibited a similar seasonality across all three sites; however, *T~w~* was lower at the old peat compared to the two young wetland sites (Fig. 2c). Wetlands were similarly managed to maintain year-round inundation, where wetland *WTD* was maintained above the land surface across all years (Fig. 2d). In some cases, wetland water control structures would fail causing periods when the *WTD* would drop below the land surface. These events were common at the old peat wetland in summer 2013, when *WTD* dropped below the land surface 24 total days (Fig. 2d). A more detailed summary of meteorologic conditions across sites can be found in Eichelmann et al. (in review).

```{r fig2, fig.cap="Daily mean (a) photosynthetically active radiation (*PAR*), (b) air temperature (*T~a~*), (c) water temperature (*T~w~*), and (d) water table depth (*WTD*) for the alluvium, young peat, and old peat wetland sites. The dashed horizontal line on panel d represents the ground surface elevation at water table well locations.", warning=FALSE, fig.width=7, fig.height=5}
#themes
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#create plot objects
a <- ggplot(all_daily, aes(x=date, y=PAR, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(PAR~" ("~mu~mol~m^{-2}~s^{-1}~")")) + 
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

b <- ggplot(all_daily, aes(x=date, y=Tair, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[a]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")


c <- ggplot(all_daily, aes(x=date, y=Tw, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[w]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

d <- ggplot(all_daily, aes(x=date, y=WTD, color=site)) +
  geom_line(size=0.5) + theme_bw() + 
  geom_hline(yintercept=0, linetype="dashed") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab("WTD (cm)") +
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  time_series + theme(legend.title=element_blank(), legend.position=c(.09, .45),
                      legend.margin = margin(0, 0, 0, 0), 
                      legend.background = element_rect(fill = "transparent"))

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

## CO~2~ and CH~4~ Flux Magnitudes
```{r combine_annual, echo=FALSE}
#Add age since restoration variable
peat19_yearly$age <- peat19_yearly$year - 1997
peat6_yearly$age <- peat6_yearly$year - 2011
alluvium_yearly$age <- alluvium_yearly$year - 2014

peat19_s <- subset(peat19_yearly, year != 2013) # remove year with draw down disturbance at WP

#merge to single dataset
all_sites <- rbind(peat19_s, peat6_yearly); all_sites <- rbind(all_sites, alluvium_yearly)

#remove incomplete years
all_sites <- subset(all_sites, Days >= 365)
```

```{r annual_stats, echo=FALSE, include=FALSE}
#Computation of significant differences among sites. These output are reported by hand in the following section, where significance is reported at 0.05, 0.001, and 0.0001 levels for all values where P < 0.05. In all cases, the maximum P value is reported when multiple Tukey HSD comparisons are cited. 

NEE <- aov(all_sites$tNEE ~ all_sites$site); summary(NEE) #NEE - ANOVA

GPP <- aov(all_sites$tGEP ~ all_sites$site); summary(GPP) #GPP - ANOVA

ER <- aov(all_sites$tER ~ all_sites$site); summary(ER) #ER - ANOVA

CH4 <- aov(all_sites$tCH4 ~ all_sites$site); summary(CH4) #FCH4 ANOVA
TukeyHSD(CH4) #Tukey HSD comparisons

GHG <- aov(all_sites$GHGbalance ~ all_sites$site); summary(GHG) #CO2-eq budget ANOVA
```

```{r fig3, fig.cap="Average annual fluxes of net ecosystem exchange (*NEE*), gross ecosystem photosynthesis (*GEP*), ecosystem respiration (*ER*), and methane flux (*F~CH4~*) for the three wetland sites. Solid horizontal lines are medians, boxes are the interquartile range (IQR), whiskers are the 95% confidence interval, and points denote measured outliers. Asterisks mark significant differences between sites at the *P* < 0.05 level.", echo=FALSE, fig.width=3, fig.height=8}
#theme
box_theme <- theme(axis.title.x = element_blank(), 
                   axis.title.y = element_text(size=9),
                   axis.text.x = element_text(size=8))

a <- ggplot(all_sites, aes(x=site, y=tNEE)) +
  geom_boxplot() + 
  ylab(expression(NEE~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) +
  #annotate("text",x=0.6,y=270,label="(a)") +
  box_theme

b <- ggplot(all_sites, aes(x=site, y=abs(tGEP))) +
  geom_boxplot() + 
  ylab(expression(GEP~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) + 
  #annotate("text",x=0.6,y=2250,label="(b)") +
  theme(axis.title.x = element_blank(), axis.title.y = element_text(size=10)) + box_theme

c <- ggplot(all_sites, aes(x=site, y=tER)) +
  geom_boxplot() + 
  ylab(expression(ER~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) + 
  #annotate("text",x=0.6,y=1850,label="(c)") +
  box_theme

sig_comparisons <- list( c("Alluvium", "Old Peat"), c("Alluvium", "Young Peat"))

d <- ggplot(all_sites, aes(x=site, y=tCH4)) +
  geom_boxplot() + 
  ylab(expression(F[CH4]~" ("~g~CH[4]-C~m^{-2}~yr^{-1}~")")) + 
  scale_y_continuous(limits=c(20,75)) +
  stat_compare_means(comparisons=sig_comparisons, label="p.signif", method="t.test") +
  #annotate("text",x=0.6,y=75,label="(d)") +
  box_theme


grid.arrange(a, b, c, d, ncol=1)

```

Mean annual net ecosystem exchange (*NEE*) budgets were not significantly different across the three wetland sites (ANOVA, *P* = 0.7), though all sites were CO~2~ sinks on average (Fig. 3). Mean annual *NEE* was `r mean(alluvium_yearly$tNEE)` $\pm$ `r sd(alluvium_yearly$tNEE)` (*n* = `r length(alluvium_yearly$tNEE)`), `r mean(peat19_s$tNEE)` $\pm$ `r sd(peat19_s$tNEE)` (*n* = `r length(peat19_yearly$tNEE)`), and `r mean(peat6_yearly$tNEE)` $\pm$ `r sd(peat6_yearly$tNEE)` (*n* = `r length(peat6_yearly$tNEE)`) g CO~2~-C m^-2^ yr^-1^ for the alluvium, old peat, and young peat wetlands, respectively (mean $\pm$ SD). Similarly, mean annual *GEP* (ANOVA, *P* = 0.97) and *ER* (ANOVA, *P* = 0.9) did not vary across the three wetland sites (Fig. 3). These averages comprise three complete years for the alluvium wetland, four years for the old peat wetland, and six years for the young peat wetland (Fig. 4). Ecosystem CO~2~ fluxes (*NEE*, *GEP*, and *ER*) followed similar seasonal patterns across all three wetlands, where peak daily fluxes were observed during summer and the lowest fluxes were observed January to February of each year (Fig. 4). Daily *NEE* for the three wetlands overlapped in range, particularly in 2015, though the young peat wetland had highly variable fluxes across years (Fig. 4a). At this site, large *GEP* and *ER* fluxes were observed one year post-restoration when the vegetation was completely established, and fluxes were attenuated mid-growing season in 2014 and 2016 due to insect outbreaks and salinization, respectively (Fig. 4b,c). Young peat CO~2~ fluxes were similarly reduced in the 2017 growing season, coincident with an insect outbreak and large-scale flushing event to reduce salinity (Fig. 4a,b,c). CO~2~ fluxes from the old peat wetland were much more consistent across time (Fig. 4a,b,c), and once the alluvium wetland was fully vegetated (2015 onward), *NEE*, *GEP*, and *ER* fluxes from this system were very similar to the old peat wetland site (Fig. 4a,b,c). Additionally, CO~2~ flux patterns from the young peat and alluvium wetland were similar during the initial flooding phase when wetland vegetation cover was not fully established. Here, peak CO~2~ fluxes were observed late in the growing season when plant cover became established within the flux footprints (Fig. 4a,b,c).

Mean annual CH~4~ fluxes varied significantly across sites (ANOVA, *P* < 0.01). Annual CH~4~ fluxes from the alluvium wetland were significantly lower than either peat site (Tukey's HSD, *P* < 0.05), and did not vary between the peat wetlands (Tukey's HSD, *P* = 0.92) (Fig. 3). Annual CH~4~ fluxes were `r mean(alluvium_yearly$tCH4)` $\pm$ `r sd(alluvium_yearly$tCH4)` (*n* = `r length(alluvium_yearly$tCH4)`), `r mean(peat19_s$tCH4)` $\pm$ `r sd(peat19_s$tCH4)` (*n* = `r length(peat19_yearly$tCH4)`), and `r mean(peat6_yearly$tCH4)` $\pm$ `r sd(peat6_yearly$tCH4)` (*n* = `r length(peat6_yearly$tCH4)`) g CH~4~-C m^-2^ yr^-1^ for the alluvium, old peat, and young peat wetlands, respectively. Methane fluxes from the old peat wetland were highly variable in 2013 due to multiple periods when *WTD* dropped below the land surface for 24 total days (Fig. 4d). Old peat wetland fluxes from 2013 were therefore omitted from the annual flux analysis because they were not representative of stable state wetland fluxes. One similar draining event occurred in 2014, but CH~4~ fluxes recovered from this event and late season fluxes were similar to the young peat wetland. Methane fluxes from the two peat sites were very similar for 2015 and 2016; however, CH~4~ fluxes from the alluvium wetland were much lower than either peat site (Fig. 4d). Consequently, annual GHG equivalent budgets were higher from the old (`r mean(peat19_s$GHGbalance)` $\pm$ `r sd(peat19_s$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^) and young peat sites (`r mean(peat6_yearly$GHGbalance)` $\pm$ `r sd(peat6_yearly$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^) compared to the alluvium site (`r mean(alluvium_yearly$GHGbalance)` $\pm$ `r sd(alluvium_yearly$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^); however, these differences were not statistically significant (ANOVA, *P* = 0.15).

```{r fig4, fig.cap="Daily (a) net ecosystem exchange (*NEE*), (b) gross ecosystem photosynthesis (*GEP*), (c) ecosystem respiration (*ER*), and (d) methane flux (*F~CH4~*) from the alluvium, old peat, and young peat wetlands. Flux measurements from the young peat and alluvium wetlands begin at the point of restoration.", warning=FALSE, fig.width=7, fig.height=5.2}
#themes
times_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#plot objects, complete years only
a <- ggplot(all_daily, aes(x=date, y=gCO2, color=site)) +
  geom_line() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(NEE~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.position = "none")

b <- ggplot(all_daily, aes(x=date, y=gGEP, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(GEP~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.position = "none")

c <- ggplot(all_daily, aes(x=date, y=gER, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  scale_y_continuous(limits=c(0,15)) +
  ylab(expression(ER~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.title=element_blank(),
                       legend.position=c(.08, .68), 
                       legend.margin = margin(0, 0, 0, 0),
                       legend.background = element_rect(fill = "transparent"))

d <- ggplot(all_daily, aes(x=date, y=mgCH4, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2011-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  times_series + theme(legend.position = "none")

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

## Flux patterns at recently restored sites
Methane flux trajectories from the young peat and alluvium wetlands were quite different following initial restoration (Fig. 5). General flux patterns were similar across sites, where CH~4~ fluxes peaked late-season during year zero (days 1-365), and peak CH~4~ fluxes occurred mid-summer in the following growing seasons (Fig. 5). Flux magnitudes were considerably lower from the alluvium site for the first three years following restoration (Fig. 5). Alluvium wetland fluxes increased with each year, whereas CH~4~ fluxes from the young peat wetland peaked during year one and decreased with subsequent years (Figs. 4d and 5). Differences in daily fluxes across wetlands were largest during year one (days 366-730) when alluvium wetland fluxes were increasing and young peat wetland emissions were peaking (Fig. 5). Annual CH~4~ budgets for the young peat site were over two times larger than the alluvium site during year zero and one. Year zero budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2014]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2011]` g CH~4~ m^-2^ yr^-1^ for the alluvium and young peat wetland, respectively, and year one budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2015]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2012]` g CH~4~ m^-2^ yr^-1^ for the alluvium and young peat wetland, respectively. This gap began to close during year two (days 731-1095), when annual fluxes were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2016]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2013]` g CH~4~ m^-2^ yr^-1^ from the alluvium and young peat wetland, respectively. During year three (days > 1095), fluxes from the young peat and alluvium wetlands were roughly equivalent, and differences observed in previous years were no longer apparent (Fig. 5). Similarly, CH~4~ fluxes from the old peat and alluvium wetlands were similar during the 2017 growing season (Fig. 4d). Methane fluxes from the young peat wetland were lower and unstable for most of 2017, but by mid-summer reached a similar magnitude as the other sites (Fig. 4d). Reduced CH~4~ fluxes at the young peat site were likely due to multiple sources of disturbance, namely insect outbreak and freshwater flushing, which also impacted CO~2~ fluxes in 2017 (Fig. 4).

```{r fig5, fig.cap="Methane fluxes (*F~CH4~*) at the young peat and alluvium wetland from time since initial restoration. Day zero is 2011-01-01 for the young peat wetland and 2014-01-01 for the alluvium wetland.", fig.width=3.5, fig.height=2, warning=FALSE}
#subset and index data
trajectory <- subset(all_daily, site != "Old Peat") #remove old peat site

#create a year since restoration column
trajectory$rest_year <- ifelse(trajectory$site == "Alluvium",  
                         trajectory$year - 2013, trajectory$year - 2010)

#compare first three years only (we only have 3.5 yrs of alluvium site)
trajectory <- subset(trajectory, rest_year < 5)

#Early restoration time series
ggplot(trajectory, aes(x=(dday-365), y=mgCH4, color=site)) +
  geom_point(aes(fill=site), colour="black", pch=21, size=0.9) +
  ylab(expression(F[CH4]~" "~"("~mg~C~m^{-2}~day^{-1}~")")) + 
  xlab("Day since restoration") +  theme_bw() + 
  scale_x_continuous(breaks=c(0, 365, 730, 1095)) +
  theme(legend.title=element_blank(), legend.position=c(.18, .82),
        legend.margin = margin(0, 0, 0, 0), legend.background = element_rect(fill = "transparent"),
        axis.title = element_text(size=10), panel.grid.minor=element_blank(), 
        panel.grid.major.y = element_blank())

```

```{r mobile_towers, include=FALSE}
#subset all Twitchell Island four year old sites and combine to single data set
alluvium4 <- select(alluvium_all, wm, datetime) #core alluvium site
alluvium4$site <- "Alluvium (core site)"
alluvium4 <- subset(alluvium4, datetime > "2017-06-30" & datetime <= "2017-08-15")

peat4 <- select(mobile_peat, wm, datetime) #mobile peat site
peat4$site <- "Peat (temporary)"
peat4 <- subset(peat4, datetime > "2017-06-30" & datetime <= "2017-08-15")

alluvium2_4 <- select(mobile_alluvium, wm, datetime)#mobile alluvium site
alluvium2_4$site <- "Alluvium (temporary)"
alluvium2_4 <- subset(alluvium2_4, datetime > "2017-06-30" & datetime <= "2017-08-15")

#combine to single dataframe
twitchell <- rbind(alluvium2_4, peat4); twitchell <- rbind(twitchell, alluvium4)

## summary statistics for comparisons across site. P reported at the 0.0001, 0.001, 0.01, 0.05 in text
#differences in flux magnitudes
t.test(alluvium4$wm, alluvium2_4$wm, alternative="less")
t.test(alluvium4$wm, peat4$wm, alternative = "less")

ks.test(alluvium4$wm, alluvium2_4$wm)
ks.test(alluvium4$wm, peat4$wm)

```

We also measured ecosystem CH~4~ fluxes from two additional locations within the wetland complex containing the alluvium site, where age was held constant (four years old) but underlying soil properties vary (Fig. 1). Simultaneous measurements across all three sites occurred between July 1 and August 15, 2017. Methane fluxes from the two alluvium sites generally overlapped in range, and mean fluxes were `r mean(alluvium4$wm, na.rm=T)` $\pm$ `r sd(alluvium4$wm, na.rm=T)` and `r mean(alluvium2_4$wm, na.rm=T)` $\pm$ `r sd(alluvium2_4$wm, na.rm=T)` nmol CH~4~ m^-2^ s^-1^ (mean $\pm$ SD) from the core and temporary alluvium site, respectively (Fig. 6). Despite overlapping ranges, CH~4~ fluxes magnitudes were larger (Welch's t, *P* < 0.0001) and more broadly distributed (Kolmogorov-Smirnov, *P* < 0.0001) at the temporary alluvium site compared to the core alluvium site (Fig. 6). Mean CH~4~ fluxes from the temporary peat site were `r mean(peat4$wm, na.rm=T)` $\pm$ `r sd(peat4$wm, na.rm=T)` nmol CH~4~ m^-2^ s^-1^, much larger than fluxes from the core alluvium site (Welch's t, *P* < 0.0001), and their distributions exhibited little overlap in range (Kolmogorov-Smirnov, *P* < 0.0001) (Fig. 6). In general, CH~4~ fluxes from the core alluvium site were more similar to the temporary alluvium site than the temporary peat site (Fig. 6).

```{r fig6, fig.cap="Distribution of methane fluxes (*F~CH4~*) from the three sites on Twitchell Island where wetland age is held constant and parent soil type varies. All measurements were made during the 2017 growing season between July 1 and August 15. All sites were restored in 2014. The temporary peat tower measured fluxes from a Rindge peat soil series, and the temporary alluvium tower measured fluxes from the same Scribner alluvium soil as the core alluvium site. Both temporary towers were located within 500m of the core alluvium tower.", fig.height=3, fig.width=3.5}
ggplot(data=twitchell, aes(wm, color=site, fill=site)) +
  geom_density(alpha=0.1) +
  xlab(expression(F[CH4]~" "~"("~nmol~CH[4]~m^{-2}~s^{-1}~")")) + 
  theme_bw() + 
  theme(legend.title=element_blank(), legend.position=c(.75, .88),
        legend.margin = margin(0, 0, 0, 0), legend.background = element_rect(fill = "transparent"),
        axis.title = element_text(size=10), 
        panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8))
```

## Biophysical drivers of CH~4~ flux
During the first year post-restoration, when differences in CH~4~ flux magnitudes between the young peat and alluvium wetland were largest (Fig. 5), CH~4~ flux patterns and their relations to biophysical drivers were notably different across sites. At the young peat wetland in 2012, CH~4~ fluxes peaked between 10:00 to 12:00 local time (LT) coinciding with the period of peak *NEE* (Fig. S1). Similar diel CH~4~ flux patterns were also observed from the old peat wetland, and these mid-day CH~4~ flux peaks were consistent across years at the old peat site (Fig. S1). Methane fluxes from the young peat wetland were strongly coupled to plant processes (*NEE*, *GEP*, and *ET*) across multiple time scales during year one (Fig. 7a,c,e). At the hourly time scale, CH~4~ flux shared most information with *ET* (Fig. 7a), and at the diel scale, dominant interactions were synchronous and related to plant processes, such as *ET*, *GEP*, and *NEE*, as well as wind direction (*WD*) (Fig. 7c). At the multiday time scale, CH~4~ flux only shared significant levels of information with *ET* and *WD* (Fig. 7e). In total, 51.5% of total CH~4~ flux variability occurred at the diel scale, 33.4% occurred the multiday scale, and 15.1% occurred at the hourly time scale (Fig. 7a,c,e).

![Mutual information (*I^R^*) shared between methane flux (*F~CH4~*) and biophysical variables at multiple time scales for the young peat and alluvium wetland (a-f) one year post-restoration and (g-l) during the 2016 growing season. Black bars indicate *I^R^* at zero time lag, and colored extension indicates maximum *I^R^* observed at the time lag illustrated by the color scale. Vertical grey bars represent 95% confidence intervals generated from 1000 Monte Carlo random walks. Italicized percentage inset on each panel indicates the percent of total wavelet variance occurring at each time scale. Biophysical variables include net ecosystem exchange (*NEE*), gross ecosystem photosynthesis (*GEP*), evapotranspiration (*ET*), air temperature (*T~a~*), air pressure (*PA*), friction velocity (*u$^\ast$*), wind direction (*WD*), soil temperature (*T~s~*), and water temperature (*T~w~*).](fig8.png)

<!-- # ```{r fig8, fig.height = 5, fig.width = 3.5, fig.cap="Diel patterns of (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), and (c) evapotranspiration (*ET*) for the first year following restoration and 2016 for both recently restored wetlands. One year post-restoration was 2012 and 2015 for the young peat and alluvium wetland, respectively, when the wetlands were fully colonized by plants. By 2016 the young peat wetland was six years old and the alluvium wetland was three years old. Solid lines and shaded regions represent means $\\pm$ 1 standard deviation.", echo=FALSE} -->
<!-- # source("R/Diel_patterns.R") #process the data -->
<!-- #  -->
<!-- # #Compare first year fluxes and 2016. Indexing and subsetting (young peat and alluvium only) -->
<!-- # diel_s <- subset(diel_all, site == "Young Peat" | site == "Alluvium") -->
<!-- # diel_s$index <- ifelse(diel_s$site == "Young Peat" & diel_s$year == 2012, "Year 1", -->
<!-- #                        ifelse(diel_s$site == "Alluvium" & diel_s$year == 2015, "Year 1", -->
<!-- #                        diel_s$year)) -->
<!-- #  -->
<!-- # diel_s <- subset(diel_s, index == "Year 1" | index== "2016") # comparing most recent and 1st year -->
<!-- #  -->
<!-- # #turn index into factor for plotting -->
<!-- # diel_s$index = factor(diel_s$index, levels=c('Year 1', '2016')) -->
<!-- #  -->
<!-- # ##plotting -->
<!-- # #themes -->
<!-- # time_series <- theme_bw() + -->
<!-- #   theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(), -->
<!-- #         axis.title = element_text(size=10)) -->
<!-- #  -->
<!-- # #plot objects -->
<!-- # a <- ggplot(diel_s, aes(x=time, y=CH4, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   geom_ribbon(aes(x=time, ymin=CH4-sdCH4, ymax=CH4+sdCH4, group=site, fill=site), alpha=0.1) + -->
<!-- #   ylab(expression(F[CH4]~" ("~nmol^{-2}~s^{-1}~")")) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(legend.position = "none", axis.title.x = element_blank()) -->
<!-- #  -->
<!-- # b <- ggplot(diel_s, aes(x=time, y=NEE, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   ylab(expression(NEE~" ("~mu~mol^{-2}~s^{-1}~")")) +  -->
<!-- #   geom_ribbon(aes(x=time, ymin=NEE-sdNEE, ymax=NEE+sdNEE, group=site, fill=site), alpha=0.1) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(legend.position = "none", axis.title.x = element_blank(), -->
<!-- #                       strip.background = element_blank(), strip.text = element_blank()) -->
<!-- #  -->
<!-- # c <- ggplot(diel_s, aes(x=time, y=ET, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   ylab(expression(ET~" ("~mmol^{-2}~s^{-1}~")")) + xlab("Time of day") + -->
<!-- #   geom_ribbon(aes(x=time, ymin=ET-sdET, ymax=ET+sdET, group=site, fill=site), alpha=0.1) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(strip.background = element_blank(), strip.text = element_blank(), -->
<!-- #                       legend.position = c(0.65,.85), legend.title = element_blank(), -->
<!-- #                       legend.margin = margin(0, 0, 0, 0), legend.key.size =  unit(0.15, "in"), -->
<!-- #                        legend.background = element_rect(fill = "transparent"),  -->
<!-- #                       legend.text = element_text(size=7)) -->
<!-- #  -->
<!-- # grid.arrange(a, b, c, ncol=1) -->
<!-- # ``` -->

Alluvium wetland CH~4~ flux patterns and interactions were notably different from to the young peat wetland one year post-restoration. Alluvium wetland CH~4~ fluxes were dominantly coupled to physical processes, such as temperature and atmospheric pressure fluctuations (Fig. 7b,d,f), in contrast to the young peat site where fluxes were primarily coupled to plant processes (Fig. 7a,c,e). At the hourly scale, CH~4~ fluxes also shared most information with *ET* (Fig. 7b). At the diel scale, dominant interactions were observed with *PA* and *T~w~* at no time lag, and with *T~a~* at a three hour lag (Fig. 7d). This dominant diel *PA* interaction was not observed at the peat site (Fig. 7c), and at the alluvium wetland maximum CH~4~ fluxes were observed in the late afternoon (17:00 to 19:00 LT) during *PA* lows (Fig. S2). Strong interactions were also observed between CH~4~ flux and *GEP*/*NEE*/*ET*, but these interactions occurred at a four to five hour time lag (Fig. 7d). These interactions can be intuitively seen through the timing of fluxes, where peak *NEE* and *ET* were observed mid-day, while peak CH~4~ flux occurred many hours later in the afternoon (Fig. S1). At the multiday scale, alluvium CH~4~ fluxes only shared significant information with temperature variables (*T~a~* and *T~w~*) (Fig. 7f). Hourly scale CH~4~ flux variability was more dominant at the alluvium wetland in year one, as 57% of total CH~4~ flux signal variability occurred at the diel scale, 30.5% occurred at the hourly scale, and only 12.5% occurred at the multiday time scale (Fig. 7).

By the 2016 growing season, CH~4~ flux patterns and interactions had converged across the recently restored wetland sites. Peak CH~4~ fluxes were observed at roughly 15:00 LT at both sites by 2016. This shift to an afternoon peak flux occurred in 2013 at the young peat wetland, and patterns in flux timing were similar for each year thereafter (Fig. S1). Both alluvium and young peat sites were quite different in terms of flux timing compared to the old peat site, where peak CH~4~ flux occurred around noon most years (Fig. S1). 

Scale-emergent controls of CH~4~ flux were also very similar between the alluvium and young peat site during the 2016 growing season (Fig. 7). For both sites, hourly CH~4~ fluxes exhibited dominant synchronous couplings to *ET* and *u$^\ast$* (Fig. 7g,h), and at the diel scale, dominant interactions with CH~4~ flux included *T~a~*, *ET*, *GEP*, and *NEE* at both wetlands (Fig. 7i,j). By 2016, dominant synchronous *PA* interactions observed at the alluvium site (Fig. 7d) were no longer apparent, though *PA* shared similar levels of information with CH~4~ flux at a 5-6 hour time lag (Fig. 7j). Multiday scale interactions were also similar, as significant interactions with temperature variables (*T~a~*, *T~w~*, and *T~s~*), as well as *ET*, were observed across both recently restored wetlands (Fig. 7k,l). There were some notable differences between sites at the multiday scale, as CH~4~ flux also shared information with *NEE* at the young peat site (Fig. 7k), and CH~4~ flux shared information with *PA* at the alluvium site (Fig. 7l). At the young peat wetland, most variability still occurred at the diel scale (48.1%), followed by the multiday (29.2%), and hourly (22.7%) time scales. Most variability at the alluvium wetland still occurred at the diel scale (44.6%), followed closely by the hourly scale (41.4%), and much less variability occurred at the multiday scale (14%) (Fig. 7).

## Soil Properties
```{r soils, echo=FALSE, include=FALSE}
source("R/soils.R") #process soils data into summary dataframe

#Computation of signficant differences among sites. These output are reported by hand in the following section, where significance is reported at 0.05, 0.001, and 0.0001 levels for all values where P < 0.05. In all cases, the maximum P value is reported when multiple Tukey HSD comparisons are cited.  Data were normally distrubuted within each site/horizon combination, so ANOVA and Tukey's were used.

#Accreted comparisons
accreted <- subset(soils_df, horizon == "Accreted")

hclfe_A <- aov(accreted$FeT_mg.g ~ accreted$site) #hcl fe
summary(hclfe_A); TukeyHSD(hclfe_A)

feII_A <- aov(accreted$FeII_mg.g ~ accreted$site) #feII
summary(feII_A); TukeyHSD(feII_A)

pcfe_A <- aov(accreted$FePC_mg.g ~ accreted$site) #fe PC
summary(pcfe_A); TukeyHSD(pcfe_A)

oxfe_A <- aov(accreted$percent_ox ~ accreted$site) #percent oxidized
summary(oxfe_A); TukeyHSD(oxfe_A)

pH_A <- aov(accreted$pH ~ accreted$site) #pH
summary(pH_A); TukeyHSD(pH_A)

percentC_A <- aov(accreted$C ~ accreted$site) #percent C
summary(percentC_A); TukeyHSD(percentC_A)

## Parent comparisons
parent <- subset(soils_df, horizon == "Parent")

hclfe_P <- aov(parent$FeT_mg.g ~ parent$site) #hcl fe
summary(hclfe_P); TukeyHSD(hclfe_P)

feII_P <- aov(parent$FeII_mg.g ~ parent$site) #feII
summary(feII_P); TukeyHSD(feII_P)

pcfe_P <- aov(parent$FePC_mg.g ~ parent$site) #fe PC
summary(pcfe_P); TukeyHSD(pcfe_P)

oxfe_P <- aov(parent$percent_ox ~ parent$site) #percent oxidized
summary(oxfe_P); TukeyHSD(oxfe_P)

pH_P <- aov(parent$pH ~ parent$site) #pH
summary(pH_P); TukeyHSD(pH_P)

percentC_P <- aov(parent$C ~ parent$site) #percent C
summary(percentC_P); TukeyHSD(percentC_P)
```

Extractable Fe concentrations were significantly higher at the alluvium site compared to both peat sites in the recently accreted (Tukey's HSD, *P* < 0.0001) and parent soil horizons (Tukey's HSD, *P* < 0.001) (Fig. 8a). Extractable Fe concentrations did not vary between parent horizons of the two peat sites (Tukey's HSD, *P* = 0.14), but were significantly lower in the recently accreted horizon of the old peat site compared to the young peat site (Tukey's HSD, *P* = 0.0001) (Fig. 8a). At the alluvium site, extractable Fe concentrations did not vary with depth; however, extractable Fe concentrations were lower in the accreted horizon at both old peat (Welch's t, *P* < 0.0001) and young peat sites (Welch's t, *P* < 0.01) (Fig. 8a). Mean Fe concentrations were `r mean(soils_df$FeT_mg.g[which(soils_df$site == "Young Peat" & soils_df$horizon == "Parent")]) - mean(soils_df$FeT_mg.g[which(soils_df$site == "Young Peat" & soils_df$horizon == "Accreted")])` and `r mean(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Parent")]) - mean(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Accreted")])` mg Fe g^-1^ soil lower in accreted compared to parent horizons at the young and old peat sites, respectively (Fig. 8a). Extractable Fe concentrations were close to zero in the accreted horizon at the old peat site (median = `r median(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Accreted")])` mg Fe g^-1^), where the accreted layer was predominantly poorly decomposed organic matter. Appreciable quantities of Fe^3+^ were present in all other site horizons, and `r mean(soils_df$percent_ox[which(soils_df$site == "Alluvium")])`%, `r mean(soils_df$percent_ox[which(soils_df$site == "Young Peat")])`%, and `r mean(soils_df$percent_ox[which(soils_df$site == "Old Peat" & soils_df$horizon == "Parent")], na.rm=T)`% of extractable Fe was in Fe^3+^ form at the alluvium site, young peat site, and parent horizon of the old peat site, respectively. Most Fe was in the reduced Fe^2+^ form across all sites, and statistical trends in Fe^2+^ concentrations were similar to those of total extractable Fe (Fig. 8b). While total extractable Fe concentrations did not vary between horizons at the alluvium site, Fe^2+^ concentrations were significantly higher in the accreted horizon at the alluvium site compared to its underlying parent soil (Welch's t, *P* < 0.05).

Reducible Fe oxide (Fe~ca~) concentrations were roughly two orders of magnitude lower than HCl-extractable Fe concentrations across all three sites (Fig. 8c), and trends were similar to extractable Fe. Soil Fe~ca~ concentrations were significantly higher at the alluvium site compared to both peat sites in the recently accreted (Tukey's HSD, *P* < 0.0001) and parent soil horizons (Tukey's HSD, *P* < 0.0001), and concentrations did not vary between the two peat sites in either the accreted (Tukey's HSD, *P* = 0.31) or parent soil horizons (Tukey's HSD, *P* = 0.74) (Fig. 8c). Soil Fe~ca~ concentrations trended lower in accreted horizons of all sites, but these differences were only significant at the old peat wetland (Welch's t, *P* < 0.001). Mean Fe~ca~ concentrations were over an order of magnitude higher in alluvium soils (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Alluvium")])` $\pm$ `r sd(soils_df$FePC_mg.g[which(soils_df$site == "Alluvium")])` mg Fe g^-1^ soil; mean $\pm$ SD) compared to the young (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Young Peat")])` $\pm$ `r sd(soils_df$FePC_mg.g[which(soils_df$site == "Young Peat")])` mg Fe g^-1^ soil) and old (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Old Peat")])` $\pm$ `r sd(soils_df$FePC_mg.g[which(soils_df$site == "Old Peat")])` mg Fe g^-1^ soil) peat soils (Fig. 8c).

Soil pH also varied across wetland sites (ANOVA, *P* < 0.05), where pH was lowest at the alluvium site (`r mean(soils_df$pH[which(soils_df$site == "Alluvium")])` $\pm$ `r sd(soils_df$pH[which(soils_df$site == "Alluvium")])`), intermediate at the old peat site (`r mean(soils_df$pH[which(soils_df$site == "Old Peat")])` $\pm$ `r sd(soils_df$pH[which(soils_df$site == "Old Peat")])`), and near-neutral at the young peat site (`r mean(soils_df$pH[which(soils_df$site == "Young Peat")], na.rm=T)` $\pm$ `r sd(soils_df$pH[which(soils_df$site == "Young Peat")], na.rm=T)`) (Fig. 8d). pH did not vary by depth at either the old or young peat wetlands; however, pH varied significantly with depth at the alluvium wetland (Welch's t, *P* < 0.0001), where the parent horizon was more acidic (Fig. 8d).

Parent horizon %C did not vary between the young peat and alluvium wetlands (Tukey's HSD, *P* = 0.50), but was higher at the old peat wetland compared to either other site (Tukey's HSD, *P* < 0.01) (Fig. 8e). Significant differences in %C were observed across all sites in the accreted horizon (Tukey's HSD, *P* < 0.0001), where %C was highest at the old peat site (`r mean(soils_df$C[which(soils_df$site == "Old Peat")])` $\pm$ `r sd(soils_df$C[which(soils_df$site == "Old Peat")])`%), intermediate at the young peat site (`r mean(soils_df$C[which(soils_df$site == "Young Peat")], na.rm=T)` $\pm$ `r sd(soils_df$C[which(soils_df$site == "Young Peat")], na.rm=T)`%), and lowest at the alluvium site (`r mean(soils_df$C[which(soils_df$site == "Alluvium")])` $\pm$ `r sd(soils_df$C[which(soils_df$site == "Alluvium")])`%). Soil %C was higher in the recently accreted horizon at both peat sites (Welch's t, *P* < 0.001), but no differences in C content were observed between horizons at the alluvium wetland (Fig. 8e).

As shown in figures 4d and 5, growing season CH~4~ fluxes from the alluvium wetland increased with time since restoration until they were the same magnitude as either peat site in 2017. If we consider growing season fluxes from year one as indicative of CH~4~ emissions from the underlying parent soil only and fluxes from 2017 as indicative of CH~4~ emissions from the combined accreted and parent soil profile, we observe a significant increase in daily CH~4~ flux between these time periods (Welch's t, *P* < 0.0001) (Fig. 8f). Similar comparisons are not appropriate for the young and old peat sites because the young peat site experienced multiple disturbances that reduced ecosystem CH~4~ flux in 2016-2017 and we have no CH~4~ flux measurements from the old peat site at the time of restoration. Significant increases in alluvium CH~4~ flux from the parent to accreted soil phase (Fig. 8f) also correspond to significant increases in Fe^2+^ (Welch's t, *P* < 0.05) and decreases in acidity (Welch's t, *P* < 0.0001) within the accreted alluvium soil horizon (Fig. 8).

```{r fig8, fig.cap="Soil (a) HCl-extractable iron (Fe), (b) ferrous iron (Fe^2+^), (c) reducible Fe oxides (Fe~ca~), (d) pH, and (e) percent carbon (%C) in accreted and parent horizons across the three core wetland sites (*n* $\\geq$ 14 per depth). Panel (f) shows daily methane fluxes (*F~CH4~*) for the early growing season (April-July) first year post-restoration and 2017 to represent *F~CH4~* when only the parent horizon was present and after accretted soil had developed, respectively. Solid horizontal lines are medians, boxes are the interquartile range (IQR), whiskers are the 95% confidence interval, and points denote measured outliers. Significance levels are shown for pairwise comparisons between soil horizons at each site (Welch's t-test), where ns denotes no significant differences and increasing asterisks denotes significance levels from *P* < 0.05, 0.01, 0.001, and 0.0001.", warning=FALSE, fig.width=7.0, fig.height=4}
source("R/soils.R") #process soils data into summary dataframe
#theme
boxplot_theme <- theme(axis.text.x = element_text(size=8),
                 axis.title.y = element_text(size=9),
                 axis.title.x = element_blank())

#add panel showing change from 1st year to 2017 alluvium fluxes
alluvium_daily$horizon <- ifelse(alluvium_daily$year == 2015, "Parent",
                            ifelse(alluvium_daily$year == 2017, "Accreted",
                            NA))

alluvium_s <- alluvium_daily[!is.na(alluvium_daily$horizon),]
all_s <- subset(all_daily, site != "Alluvium"); all_s$horizon <- NA
all_s <- rbind(alluvium_s, all_s)

all_s <- subset(all_s, month < 8 & month > 3)

a <- ggplot(soils_df, aes(x=site, y=FeT_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Extractable Fe (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,32)) +
  annotate("text",x=0.6,y=32,label="(a)") +
  boxplot_theme + theme(legend.position="none")

b <- ggplot(soils_df, aes(x=site, y=FeII_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Fe2+ (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,25)) +
  annotate("text",x=0.6,y=25,label="(b)") +
  boxplot_theme + theme(legend.position="none")

c <- ggplot(soils_df, aes(x=site, y=FePC_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression(Fe[ca]~" (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,0.5)) +
  annotate("text",x=0.6,y=0.5,label="(c)") +
  boxplot_theme + theme(legend.position="none")

d <- ggplot(soils_df, aes(x=site, y=pH, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(4.5,7.6)) +
  ylab(expression("pH")) +
  annotate("text",x=0.6,y=7.6,label="(d)") +
  boxplot_theme +
  theme(legend.position=c(.75,.2), legend.title = element_blank(), 
        legend.margin = margin(0, 0, 0, 0), 
        legend.background = element_rect(fill = "white"))

e <- ggplot(soils_df, aes(x=site, y=C, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression("%C")) +
  scale_y_continuous(limits=c(10,50)) +
  annotate("text",x=0.6,y=50,label="(e)") +
  boxplot_theme + theme(legend.position="none")

f <- ggplot(alluvium_s, aes(x=site, y=mgCH4, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression(F[CH4]~" ("~mg~CH[4]-C~m^{-2}~d^{-1}~")")) + 
  scale_y_continuous(limits=c(0,400)) +
  annotate("text",x=0.6,y=400,label="(f)") +
  boxplot_theme + theme(legend.position="none")

grid.arrange(a, b, c, d, e, f, ncol=3)
```

# Discussion
Our results suggest CH~4~ fluxes from restored wetlands are influenced by legacy soil properties from the pre-restoration landscape, causing a greater than twofold difference in annual wetland CH~4~ fluxes. We can rule out the influence of variable meterological conditions, as all the wetlands observed in this study were within 13 km of on another and were fully inundated across the measurement period (Fig. 2). Methane fluxes from wetlands restored on alluvium soils were significantly lower than those from wetlands restored on peat soils; however, with time CH~4~ flux magnitudes and controls converged across peat and alluvium wetlands. Differences in soil Fe content (Fe^2+^, Fe^3+^, and Fe~ca~) between peat and alluvium wetlands most likely explains observed flux differences. Concentrations of multiple Fe species were higher in alluvium soils compared to soils at either peat site, and Fe concentrations did not vary across parent soils of the peat sites (Fig. 8). Similarly, alluvium mean annual CH~4~ fluxes were significantly lower than either peat site, and CH~4~ fluxes did not vary between peat wetlands, mirroring the patterns in soil Fe content. High soil Fe concentrations are known to inhibit CH~4~ production in tropical forest soils [@teh_suppression_2008], but this is the first study tho demonstrate the importance of soil Fe content in attentuating ecosystem-scale CH~4~ fluxes. 

Soil C is also known to vary between Delta peat and alluvium soils [@soilsurvey], but we can rule out the importance of soil %C in driving observed CH~4~ fluxes differences across soil types for the following reasons. First, soil C did not vary between parent soil horizons at the young peat and alluvium wetlands (Fig. 8), yet we observed greater than two-fold higher CH~4~ fluxes the first years following restoration at the young peat compared to alluvium site (Fig. 5). During these early years accreted soil horizons had not yet developed, and if parent soil %C did not vary across sites it rules out the hypothesis that C substrates might limit methanogenesis at the alluvium wetland. Secondly, we might also expect to see smaller *ER* fluxes via reduced heterotrophic respiration if C substrates were limited at the alluvium wetland, but instead, *ER* fluxes from the alluvium site were often similar in magnitude or larger than those from either peat wetland (Figs 3 and 4). @mcnicol_separate_2014 demonstrated that heterotrophic respiration from these soils only reduced by 50% under flooded, anaerobic conditions compared to oxic conditions, therefore we expect that we could observe changes in heterotrophic respiration at the ecosystem-scale even under anoxic conditions where aerobic respiration is limited. Similarly, *GEP* from the alluvium wetland was similar or larger than either peat site (Figs. 3 and 4), demonstrating that the input of recent photosynthates is not limiting methanogenesis at the alluvium wetland. Finally, CH~4~ fluxes did not scale with %C at the peat sites, where soil %C was higher in both horizons of the old peat site but daily and annual CH~4~ fluxes were more or less equivalent (Figs. 3 and 4). This combination of observations demonstrates that soil %C or recent photosynthates neither clearly limits or enhances ecosystem CH~4~ fluxes across these three wetland sites.

We are also able to rule out the importance of wetland age in confounding our findings because CH~4~ fluxes were no different between the young and old peat wetlands (Figs. 3 and 4). Our temporary tower campaign further controlled for age, and we observed that CH~4~ fluxes from the core alluvium site were more similar to the temporary alluvium site than the temporary peat site (Fig. 6). It is worth noting that the flux magnitudes from this temporary peat site were notably larger than the old peat site over the same time period; however, this was a relatively short measurement campaign that does not capture seasonal to interannual variability at the temporary sites. Our measurements from the young peat site clearly demonstrate the importance of interannual variability across these sites (Fig. 4d), so interpretation of the temporary tower campaign should be restrained. Regardless, the temporary towers validate the trend that CH~4~ fluxes are larger from wetlands restored on peat soils when all else is held constant. Wetland age is an important factor in that the differences between peat and alluvium wetlands appears to be transient, and observed differences become less appartent multiple years post-restoration after new wetland soil has accumulated (Fig. 5). This convergence in fluxes is also observed in the biophsyical drivers of CH~4~ flux, as the two sites had notably different controls at diel to multiday time scales one year post-restoration but had become more similar by the 2016 growing season (Fig. 7).

Differences in biophysical CH~4~ drivers across sites further suggests that CH~4~ production was inhibited at the alluvium wetland in the first years following restoration, in line with our hypothesis that Fe presence inhibited CH~4~ production in the parent soils. Alluvium CH~4~ fluxes were decoupled from plant processes across multiple scales compared to the young peat site, where CH~4~ fluxes were more dominantly couples to plant processes, such as *GEP* and *ET* (Fig. 7). Instead, alluvium CH~4~ fluxes were more dominantely coupled to physical transport processes (*PA* fluctuations) and factors such as temperature that influence methanogenesis in bulk soil when differences in flux magnitudes between sites were largest. A strong disconnect of CH~4~ flux from plant-derived substrates (*GEP*) and transport pathways (*ET*) is expected when Fe is present in the rhizosphere. Oxygen enters soil the soil through plant roots, and therefore, Fe is more likely to be reoxidized in the plant root zone where methanogenesis could be suppressed by iron-reducing bacteria [@laanbroek_methane_2010; @bridgham_methane_2013]. It follows that CH~4~ fluxes from wetlands where rhizosphere methanogenesis is inhibited would be more closely coupled to physical processes, while a site with active rhizopshere methanogenesis would be more coupled to plant subtrate production and transport processes. @hatala_gross_2012 demonstrated that CH~4~ fluxes form rice paddies in this region were closely coupled to plant photosynthates, and peak CH~4~ fluxes were coincident with peak *GEP*. @knox_biophysical_2016 further demonstrated temporal links between rice CH~4~ flux and *GEP*, where peak CH~4~ flux occurred earlier than peak soil temperature suggesting stronger couplings to plant processes. We see a very different dynamic at the alluvium wetland site one year post-restoration, where peak CH~4~ flux occurs late in the afternoon, hours later than peak *GEP*, *ET*, and even temperature, and is instead strongly coupled to *PA* fluctuations (Fig. 7). This behavior suggests strong decoupling from plant processes and transport pathways, and a much stronger coupling to physical transport and drivers bulk soil methanogenesis. These dynamics are expected in a system where CH~4~ production is inhibited in the rhizosphere, further suggesting Fe inhibition of CH~4~ emissions from these alluvium wetlands.

Suprisingly, much more flux variability occurred at the multiday scale for the peat wetland than the alluvium wetland, and this pattern was consistent over time. While the driver of this variation is not clear, it demonstrates why eddy covariance measurements are particularly suited to ecosystem CH~4~ flux measurements. Chamber-based measurement campaigns could easily undersample CH~4~ flux hot moments, particuarly if the dominant modes of variation change over relatively small spatial scales.

# References

