---
title: "Soil iron content and sediment accretion drive methane fluxes in restored wetlands"
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Daphne J Szutu^1^, Joseph
  G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document: default
  html_document: default
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{hyperref}
link-citations: yes
csl: global-change-biology.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, California, USA

^3^National Ecological Observatory Network, Battelle, Boulder, Colorado, USA

**Running Head**: Wetland CH~4~ emission and iron interactions

**Keywords**: eddy covariance, greenhouse gas balance, carbon flux, peatland, redox, alternative electron acceptor, Sacramento-San Joaquin Delta, information theory

**Type of Paper**: Primary research article

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
library(caret)
library(modelr)
```

\newpage

# Abstract
Wetlands are the largest source of methane (CH~4~) globally, yet our understanding of how process-level controls scale to ecosystem fluxes remains limited. It is particularly uncertain how variable soil properties, such as soil iron content, influence ecosystem-scale CH~4~ emissions. We measured ecosystem carbon dioxide (CO~2~) and CH~4~ fluxes by eddy covariance from three wetlands restored on peat and alluvium soils within the Sacramento-San Joaquin Delta of California, where iron (Fe) and carbon (C) content varies across soil types. Annual CH~4~ fluxes from the alluvium wetland were significantly lower than those from either peat site for multiple years, and CH~4~ budgets did not vary across peat wetlands. We observed similar trends in soil Fe, where Fe concentrations were significantly higher in alluvium soils and did not vary across peat soils. Peat-alluvium CH~4~ fluxes were decoupled from plant processes compared to the recently restored peat site, as expected when Fe reduction inhibits CH~4~ production in the rhizosphere. Peat-alluvium wetland CO~2~ uptake rates were similar to either peat site, suggesting that C inputs were not limiting alluvium CH~4~ fluxes. Carbon content did not vary between peat and alluvium parent soils at the two recently restored wetlands and thus could also be ruled out as a driver of CH~4~ flux differences. Differences in CH~4~ fluxes between soil types were transient; alluvium CH~4~ fluxes were similar to peat wetlands three years post-restoration. Recently accreted alluvium soils were less acidic and contained more reduced Fe compared to the pre-restoration parent soils, suggesting that CH~4~ emissions increased as conditions became more favorable to methanogenesis within accreted sediments. This study demonstrates that natural variation in soil Fe is capable of inhibiting ecosystem-scale wetland CH~4~ flux, but these effects appear to be transient without continued input of alluvium to wetland sediments. (295/300 words)

\newpage

# Introduction
Wetlands play a fundamental role in regulating global climate by storing ~18% of Earth's soil carbon (C) [@lal_carbon_2008], emitting ~30% of global methane (CH~4~) emissions [@kirschke_three_2013; @saunois_global_2016], while only occupying 5-8% of the terrestrial land surface [@mitsch_wetlands_2013]. Most wetlands are greenhouse gas (GHG) sinks over multi-century time scales due to carbon dioxide (CO~2~) sequestration within anoxic soils, though they are commonly GHG sources over shorter time scales due to the high global warming potential of emitted CH~4~ [@whiting_greenhouse_2001; @mitsch_wetlands_2013; @neubauer_moving_2015; @petrescu_uncertain_2015]. Wetland C sequestration and CH~4~ emissions are inextricably linked, as anoxic soil conditions inhibiting ecosystem respiration (*ER*) also activates archaeal CH~4~ production [@whalen_biogeochemistry_2005; @conrad_global_2009]. The balance between these opposing fluxes primarily determines wetland GHG balances [@bubier_ecological_1994; @hendriks_full_2007; @petrescu_uncertain_2015], and small differences in sustained ecosystem CH~4~ emissions can induce large changes in wetland GHG exchange due to the > 45-fold radiative forcing of CH~4~ compared to CO~2~ over decadal to centennial time horizons [@neubauer_moving_2015]. Recent rises in atmospheric CH~4~ have been attributed to a global increase in wetland CH~4~ fluxes [@nisbet_rising_2016], and wetlands are currently the most uncertain component of the global CH~4~ budget [@kirschke_three_2013]. A key driver of this uncertainty is a lack of ecosystem-scale flux measurements to better constrain models and improve our understanding of how process-level controls scale to whole ecosystem CH~4~ emissions [@bridgham_methane_2013; @saunois_global_2016].

Methane fluxes from wetlands are governed by many interacting biophysical drivers including temperature, C inputs, alternative electron acceptor pools, and ecosystem inundation [@bridgham_methane_2013]. It is uncertain how these governing factors interact and scale to whole ecosystem fluxes due to sparse CH~4~ flux measurements across time and space [@petrescu_uncertain_2015], and current ecosystem measurement campaigns may under-sample relevant environmental gradients. Chamber flux syntheses have demonstrated that water table depth, temperature, vegetation, disturbance, and wetland type are important modulators of wetland CH~4~ flux [@turetsky_synthesis_2014]. A similar understanding of the controls on CH~4~ fluxes is emerging from eddy covariance studies, where temperature [@rinne_annual_2007; @wille_methane_2008; @hendriks_multi-technique_2010; @olson_interannual_2013; @chu_net_2014], recent C inputs [@hatala_gross_2012; @morin_environmental_2014], wetland structure [@matthes_parsing_2014; @mcnicol_effects_2017], and water table depth [@hendriks_full_2007; @hendriks_multi-technique_2010; @brown_evidence_2014; @goodrich_overriding_2015; @chamberlain_underlying_2015; @chamberlain_influence_2016; @sturtevant_identifying_2016; @chamberlain_impact_2017] have been identified as dominant controls across many wetland ecosystems. Small-scale field and laboratory studies have also demonstrated the influence of soil C content [@levy_methane_2012; @bridgham_methane_2013; @ye_soil_2016] and alternative electron acceptor pools, such as sulfate and ferric iron, on wetland CH~4~ fluxes [@laanbroek_methane_2010; @poffenbarger_salinity_2011; @bridgham_methane_2013; @miller_methane_2015]; however, the importance of these factors has not been consistently demonstrated at ecosystem scale. Recently, @holm_ecosystem_2016 and @krauss_component_2016 observed attenuation of ecosystem CH~4~ fluxes with increasing salinity (i.e. sulfate concentrations), verifying the ecosystem-scale relevance of previous chamber-based observations [@poffenbarger_salinity_2011]. Iron (Fe) minerals can also inhibit ecosystem CH~4~ fluxes [@laanbroek_methane_2010], and while this has been observed in many laboratory studies [@jackel_suppression_2000; @teh_suppression_2008; @zhou_methanogenesis_2014], Fe inhibition to ecosystem fluxes has not been documented. Understanding the relevance of such factors may be particularly important to constraining wetland CH~4~ emissions from the tropics, the largest source of the global wetland CH~4~ emissions [@kirschke_three_2013], where high Fe concentrations influence rates of organic matter decomposition and CH~4~ production in forest soils [@teh_suppression_2008; @dubinsky_tropical_2010; @hall_iron_2013].

Wetland restoration and management programs have increasingly been proposed and implemented to mitigate climate change [@mitsch_creating_1998; @mitsch_wetlands_2013], and restoration strategies that minimize CH~4~ emissions and maximize CO~2~ uptake will provide the optimum climate benefit. These programs are particularly appealing in coastal ecosystems where CO~2~ sequestration rates are high and CH~4~ flux rates are low [@conservation_international_blue; @poffenbarger_salinity_2011], and in drained peatlands where large CO~2~ emissions from oxidizing peat can be attenuated by re-flooding the landscape [@hatala_greenhouse_2012; @knox_agricultural_2015; @wilson_multiyear_2016]. A thorough understanding of CH~4~ flux drivers is essential to these programs because the magnitude of CH~4~ emissions may determine if restored wetlands are a GHG source or sink [@knox_agricultural_2015]. These emissions could affect funding of wetland restoration programs if financed through carbon markets alone [@oikawa_evaluation_2017], though wetlands provide a number of relevant ecosystem services, such as nutrient retention, protection from storms and sea-level rise, and habitat preservation for wildlife [@hansson_conflicting_2005; @zedler2005wetland; @swain_trade-offs_2013].

The Sacramento-San Joaquin Delta region of California (hereafter referred to as the Delta) is an area where financing of wetland restoration through California's Cap and Trade program is being considered. The Delta is comprised of a network of artificially drained islands reclaimed for agriculture, where the water table is maintained below sea level by active pumping and levees surrounding each island. Drainage began in the 1850's to facilitate agricultural development. Exposure of the highly organic soils has led to substantial land surface subsidence [@deverel_historic_2010] and CO~2~ emissions [@hatala_greenhouse_2012; @knox_agricultural_2015], and wetland restoration efforts aim to accrete new sediments and reduce GHG emissions [@miller_subsidence_2008; @miller_carbon_2011; @hatala_greenhouse_2012]. Prior to drainage, this region was a 5 billion m^3^ tidal marsh where soil mineral and C content varied substantially across space. A major driver of this soil variation was fluvial deposition from major rivers, which carried mineral alluvium from the Sierra Nevada mountain range to the Delta marshes. Alluvium marsh soils frequently developed adjacent to predominant peat soils that were spatially segregated from the main river channels [@atwater1979history; @atwater1980tidal]. Mixed alluvium soils are most common on islands adjacent to the Sacramento River compared to the central and eastern Delta region where fluvial deposition was less pronounced [@deverel_historic_2010]. These differences in pre-drainage wetland geomorphology give rise to a contemporary agricultural landscape where large changes in soil C and mineral content occur over small spatial scales [@soilsurvey], but how this edaphic variation influences current GHG fluxes from restored wetlands is unknown. Understanding the importance of these legacy effects is highly desirable, as it would allow for more informed wetland restoration strategies where sites could be chosen to minimize GHG emissions.

The objectives of this study were to (1) determine how soil properties and wetland GHG fluxes varied across wetlands restored on peat versus alluvium soils, (2) identify how relationships between biophysical drivers and CH~4~ fluxes varied across wetland types, (3) determine most likely edaphic or climatic drivers of observed flux variation. We measured ecosystem-scale CH~4~ and CO~2~ fluxes by eddy covariance for two wetlands from the point of initial restoration (inundation), where one wetland was restored on peat-dominated soils and the other was restored on a mixed alluvium soil, and compared fluxes from these sites to understand the effect of soil type on fluxes following restoration. We then measured multiple soil properties within the tower footprints, including soil C, nitrogen (N), ferrous iron (Fe^2+^), ferric iron (Fe^3+^), and poorly crystalline Fe oxides, to identify edaphic drivers of observed flux differences. These tower sites are an ideal network to isolate these controlling edaphic factors, as all wetlands sites have a similarly managed hydroperiod, are within 13 km of one another, experience the same climate, and have similar plant community compositions.

# Materials and Methods
## Site Description
We measured wetland ecosystem CH~4~ and CO~2~ fluxes using a network of eddy covariance towers in the Sacramento-San Joaquin Delta region of California, USA. The Delta is located within a Mediterranean climate that experiences hot, dry summers and cooler, rainy winters. The mean annual temperature is 15.1 $^\circ$C (1998-2015 average), and the region receives 326 mm of rainfall annually [@knox_biophysical_2016]. All measurement sites were located on Sherman and Twitchell Islands in the northwest Delta. These islands are a mosaic of mixed alluvium mollisols and highly organic histosols [@soilsurvey]. The mixed alluvium mollisols formed via fluvial deposition from major rivers, and histosols formed densely organic soils in areas less disturbed by fluvial input [@deverel_historic_2010]. The dominant mollisol series on these islands are Gazwell and Scribner (Fig. 1), both of which are formed in mixed alluvium, belong to the soil class Cumulic Endoaquolls, and are only found on Delta islands within the Sacramento River drainage [@soilsurvey]. Rindge is the major histosol series found on both islands (Fig. 1) and is widely distributed across the Delta region. Rindge soils belong to the  Typic Haplosaprists class, and are characterized by deep, poorly drained marsh soils formed from decomposed plant organic matter [@soilsurvey]. 

Wetland ecosystem fluxes have been measured by eddy covariance for over three years from three core sites. Fluxes have been measured from the time of initial restoration at two sites, where one wetland, hereafter referred to as the 'peat' site, was restored in 2010 on peat Rindge soils (Fig. 1a) and the other, hereafter referred to as the 'alluvium' site, was restored in 2014 on alluvium Scribner soils (Fig. 1b). We have also been measuring fluxes since 2013 from an older, mature wetland, hereafter referred to as the 'old peat' site, that was restored in 1999 on peat Rindge soils (Fig. 1b); however, we do not have measurements from the time of restoration. A more detailed description of all tower sites and instrumentation can be found in the Supplemental Materials. We followed a similar flux quality control procedure described in @knox_agricultural_2015, @chamberlain_evaluation_2017, and Eichelmann et al. (*in review*), gap filled and partitioned fluxes using artificial neural networks [@baldocchi_does_2015; @knox_agricultural_2015; @oikawa_revisiting_2017], and estimated flux footprints using a 2-D analytic model [@detto_soil_2006; @hsieh_approximate_2000]. A more detailed description of these approaches can be found in the Supplemental Materials.

![Soil type and tower flux footprints for (a) Sherman and (b) Twitchell Islands in the Sacramento-San Joaquin Delta of California. Tower footprints include wetland area only, and footprint rings, from largest to smallest, correspond to the 90, 85, 80, 70, and 50% cumulative flux footprints at each site. Tower footprints were calculated using a two-dimensional analytical model described in @detto_soil_2006. The white footprint rings represent the core tower sites, where the footprint on (a) in Rindge series is the peat site, the footprint on (b) in Rindge series is the old peat site, and the footprint on (b) in Scribner series is the alluvium site. The yellow footprints on (b) represent the temporary tower sites within the same wetland complex as the core alluvium site.](fig1.png)

## Soil Analyses
We measured soil C, N, and Fe content at the three wetland sites in the recently accreted wetland O horizon, hereafter referred to as accreted horizon, and the underlying pre-restoration parent soils, hereafter referred to as the parent horizon. Fifteen samples of both accreted and parent horizon soils were collected at each site across three transects, with each sampling location at least 3 m apart. The transects were within each tower's flux footprint, all samples were collected from fully inundated locations, and the accreted sample was collected directly above the underlying parent soil sample. At the peat and alluvium sites (both < 8 years old), the differentiation between these horizons was clear, as the recently accreted horizon was loose, mucky, and heavily comprised of poorly-decomposed plant matter, whereas the underlying parent horizon was compacted agricultural soil. At these sites, we collected the accreted horizon by hand (grab samples) and the top 15 cm of parent horizon using a sediment core. Our sampling strategy was different at the old peat site because more than 0.5 m of O horizon had accumulated since initial restoration. Here, the accreted horizon reached the water surface, and we collected the top 2 cm of this horizon to represent the most recently accreted peat. We then bore holes through the 0.5-0.7 m saturated peat layer to collect the underlying parent soil horizon. Differentiation between these layers was also clear, as the top 0.5-0.7 m was comprised of poorly-decomposed plant matter and the underlying horizon was a silty clay loam. These observations are consistent with previously measured accretion rates and peat depths [@miller_subsidence_2008].

All soil samples were immediately bagged, and sub-samples were field extracted in both 0.5 M HCl and 0.2 M sodium citrate/0.05 M sodium ascorbate solutions to measure HCl-extractable Fe (Fe^2+^ and Fe^3+^) and poorly crystalline Fe oxide pools, respectively. The HCl extraction solubilizes reactive Fe^3+^ minerals and absorbed/solid Fe^2+^, and is used to quantify oxidized (Fe^3+^) and reduced (Fe^2+^) Fe pools. The low pH of HCl prevents any Fe^2+^ oxidation after the samples are collected [@hall_reducing_2015]. Roughly 3 g of sample (dry mass equivalent) were immediately placed into the HCl solution, and once back in the lab, samples were vortexed, shaken for 1 hour, and centrifuged for 10 minutes at 1000 rcf. Concentrations of Fe^2+^ and Fe^3+^ were measured colorimetrically using the ferrozine assay [@viollier_ferrozine_2000]. Citrate-ascorbate extractions were used to quantify poorly crystalline Fe oxides that are reducible by soil microbes [@hyacinthe_reactive_2006]. Roughly 1.5 g of sample (dry mass) were immediately placed into the citrate-ascorbate solutions, and in the lab, samples were vortexed, shaken for 16 hours, and then centrifuged for 20 minutes at 1000 rcf. Poorly crystalline Fe concentrations were quantified using an inductively coupled plasma optical emission spectrometer (ICP-OES; Perkin Elmer Optima 5300 DV, Waltham, Massachusetts, USA). We then air-dried soil samples at room temperature for total percent C (%C) and N (%N) analysis. Dried subsamples were sieved to 2 mm and all major visible roots were removed by hand. These samples were then ground to a fine powder and analyzed in duplicate for total C and N using an elemental analyzer (CE Elantech, Lakewood, New Jersey, USA).

## Wavelet-Information Theory Analysis
We used a combination of wavelet time series decomposition and information theory to (1) isolate major time scales of variation within the continuous CH~4~ flux time series, and (2) identify scale-emergent interactions between CH~4~ fluxes and a number of biophysical drivers. This technique allowed us to isolate controls on CH~4~ fluxes operating at hourly, diel, and multiday timescales, and has been used to identify scale-emergent controls of CH~4~ flux at the two peat restored wetlands  [@sturtevant_identifying_2016]. Relative mutual information (*I^R^*), and information theory metric, was used to identify relationships between variables. Mutual information is derived from Shannon entropy (*H*), a measure of uncertainty [@shannon1998mathematical], and *I^R^* quantifies the amount of information shared between two variables. Mutual information and other information theory metrics, such as transfer entropy, are particularly useful for identifying relationships in complex systems because they do not assume linearity or other functional relationships and are capable of identifying asynchronous relationships [@ruddell_applying_2013]. All wavelet decomposition and entropy calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software), and a more detailed description of the approach can be found in the Supplemental Materials and @sturtevant_identifying_2016.

### Statistical Analyses and Data Processing
All additional data processing, statistical analysis, and visualization was conducted in R 3.4.1 [@r_citation] using the 'tidyverse' [@tidyverse], 'zoo' [@zoo], 'gridExtra' [@gridExtra], 'ggpubr' [@ggpubr], 'lubridate' [@lubridate], and 'scales' [@scales] packages. We used parametric statistics, such as ANOVA, Welch's t-tests, and Tukey's HSD, to assess differences in soil properties across/within sites because properties were normally distributed within soil horizons at each site. Kolmogorov-Smirnov tests were used to determine whether CH~4~ fluxes across temporary tower sites originated from the same distribution, and Welch's t-tests were used to determine significant differences in flux magnitude because half-hourly CH~4~ fluxes were normally distributed within each site. Annual GHG budgets were calculated using the 45x 100-year sustained-flux global warming potential for CH~4~ presented in @neubauer_moving_2015. This manuscript is reproducible via R Markdown, and its code can be found at https://bitbucket.org/sdchamberlain/wetlandcomparison. The wavelet-information theory analysis was conducted using MATLAB and is not directly reproducible in R; however, the raw to processed datasets and MATLAB source code are available online with this manuscript.

# Results
```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat19_processing.R") #19yr old peat site
source("R/peat6_processing.R") #6yr old peat site
source("R/alluvium_processing.R") #alluvium site

#merge daily sums to single dataset
all_daily <- rbind(peat19_daily, peat6_daily); all_daily <- rbind(all_daily, alluvium_daily)
all_daily$date <- as.Date(all_daily$datetime) #data plotting variable

#merge annual budgets to single dataset
all_annual <- rbind(peat19_yearly, peat6_yearly); all_annual <- rbind(all_annual, alluvium_yearly)

#subset recently restored sites
new_wetlands <- subset(all_daily, site != "Old Peat") #remove old peat site

#create a year since restoration column
new_wetlands$rest_year <- ifelse(new_wetlands$site == "Alluvium",  
                         new_wetlands$year - 2014, new_wetlands$year - 2011)

```

## Post-restoration Flux Trajectories 
Methane flux trajectories from the peat and alluvium wetlands were quite different following initial restoration, where flux magnitudes were considerably lower from the alluvium site for the first three years following restoration and then converged with the peat site (Fig. 2a). Alluvium wetland fluxes increased with each year, whereas CH~4~ fluxes from the peat wetland peaked during year one and decreased with subsequent years (Figs. 2a). Differences in daily fluxes across wetlands were largest during year one when alluvium wetland fluxes were increasing and peat wetland emissions were peaking (Fig. 3a). Annual CH~4~ budgets for the peat site were over two times larger than the alluvium site during year zero and one. Year zero budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2014]` $\pm$ `r alluvium_yearly$ciCH4[alluvium_yearly$year == 2014]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2011]` $\pm$ `r peat6_yearly$ciCH4[peat6_yearly$year == 2011]` g CH~4~-C m^-2^ yr^-1^ (mean $\pm$ 95% CI) for the alluvium and peat wetland, respectively, and year one budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2015]` $\pm$ `r alluvium_yearly$ciCH4[alluvium_yearly$year == 2015]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2012]` $\pm$ `r peat6_yearly$ciCH4[peat6_yearly$year == 2012]` g CH~4~-C m^-2^ yr^-1^ for the alluvium and peat wetland, respectively. This gap began to close during year two, and by year three fluxes were quite similar across sites (Fig. 2a), when annual fluxes were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2017]` $\pm$ `r alluvium_yearly$ciCH4[alluvium_yearly$year == 2017]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2014]` $\pm$ `r peat6_yearly$ciCH4[peat6_yearly$year == 2014]` g CH~4~-C m^-2^ yr^-1^ from the alluvium and peat wetland, respectively.

Differences in *NEE* were less notable across the two recently restored wetlands and did not follow similar patterns to CH~4~ flux. Daily *NEE* was similar across the two wetlands during year zero and one (Fig. 2b), and annual budgets did not differ across the two sites during these years. Year zero budgets were `r alluvium_yearly$tNEE[alluvium_yearly$year == 2014]` $\pm$ `r alluvium_yearly$ciNEE[alluvium_yearly$year == 2014]` and `r peat6_yearly$tNEE[peat6_yearly$year == 2011]` $\pm$ `r peat6_yearly$ciNEE[peat6_yearly$year == 2011]` g C m^-2^ yr^-1^, and year one budgets were `r alluvium_yearly$tNEE[alluvium_yearly$year == 2015]` $\pm$ `r alluvium_yearly$ciNEE[alluvium_yearly$year == 2015]` and `r peat6_yearly$tNEE[peat6_yearly$year == 2012]` $\pm$ `r peat6_yearly$ciNEE[peat6_yearly$year == 2012]` g CO~2~-C m^-2^ yr^-1^ for the alluvium and peat wetland, respectively. During year two, the peat wetland sequestered less CO~2~ despite emitting more CH~4~ compared to the alluvium wetland (Fig. 2b). Here, annual *NEE* was significantly less from the peat wetland (`r peat6_yearly$tNEE[peat6_yearly$year == 2013]` $\pm$ `r peat6_yearly$ciNEE[peat6_yearly$year == 2013]` g CO~2~-C m^-2^ yr^-1^) compared to the alluvium wetland (`r alluvium_yearly$tNEE[alluvium_yearly$year == 2016]` $\pm$ `r alluvium_yearly$ciNEE[alluvium_yearly$year == 2016]` g CO~2~-C m^-2^ yr^-1^). Similar to CH~4~ fluxes during year three, daily *NEE* were roughly equivalent between the two recently restored sites, though the onset of the growing season varied (Fig. 2b).

Other dominant biophysical drivers of GHG fluxes (*ET*, *PAR*, *T~a~*, *WTD*) were similar across the wetlands, despite measurements occurring across different time periods (2011 - 2014 for the peat site; 2014 - 2017 for the alluvium site). Daily *ET* exhibited a similar seasonality and magnitude across both both sites, with the exception of year one at the peat wetland where particularly large *ET* rates were observed (Fig. 2c). Daily mean *PAR* and *T~a~* were also similar across the wetlands, where values generally overlapped in range (Fig. 2d,e). Additionally, the water table was always above surface at either wetland for the first three years following restoration and is therefore not responsible for reduced CH~4~ fluxes at the alluvium site (Fig. 2f).

```{r fig2, fig.cap="Daily (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), (c) evapotranspiration (*ET*), as well as mean (d) photosynthetically active radiation (*PAR*), (e) air temperature (*T~a~*), and (f) water table depth (*WTD*) at the peat and alluvium wetland from time since initial restoration. Day zero is 2011-01-01 for the peat wetland and 2014-01-01 for the alluvium wetland.", fig.width=6, fig.height=9, dpi=600}
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#compare first three years only (we only have 3.5 yrs of alluvium site)
trajectory <- subset(new_wetlands, rest_year < 4)

#Early restoration time series
a <- ggplot(trajectory, aes(x=(dday-365)/365, y=mgCH4, color=site)) +
  geom_point(aes(fill=site), colour="black", pch=21, size=0.9) +
  ylab(expression(F[CH4]~" "~"("~mg~C~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=4, y=Inf, label="(a)", vjust=1, hjust=1) +
  theme_bw() + time_series +
  theme(legend.title=element_blank(), legend.margin = margin(0, 0, 0, 0),
        legend.position=c(.15, .85),
        legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8),
        legend.background = element_rect(fill = "transparent"))

b <- ggplot(trajectory, aes(x=(dday-365)/365, y=gCO2, color=site)) +
  geom_point(aes(fill=site), colour="black", pch=21, size=0.9) +
  ylab(expression("NEE "~"("~g~C~m^{-2}~d^{-1}~")")) + 
  xlab("Wetland age (years)") +  
  annotate("text", x=4, y=Inf, label="(b)", vjust=1, hjust=1) +
  theme_bw() + time_series + theme(legend.position = "none")

c <- ggplot(trajectory, aes(x=(dday-365)/365, y=mET, color=site)) +
  geom_point(aes(fill=site), colour="black", pch=21, size=0.9) +
  ylab(expression("ET ("~mm~d^{-1}~")")) +
  annotate("text", x=4, y=Inf, label="(c)", vjust=1, hjust=1) +
  theme_bw() + time_series + theme(legend.position = "none")

d <- ggplot(trajectory, aes(x=(dday-365)/365, y=PAR, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  ylab(expression(PAR~" ("~mu~mol~m^{-2}~s^{-1}~")")) + 
  annotate("text", x=4, y=Inf, label="(d)", vjust=1, hjust=1) +
  theme_bw() + time_series + theme(legend.position = "none")

e <- ggplot(trajectory, aes(x=(dday-365)/365, y=Tair, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  ylab(expression(T[a]~" ("~degree~C~")")) +
  annotate("text", x=4, y=Inf, label="(e)", vjust=1, hjust=1) +
  theme_bw() + time_series + theme(legend.position = "none")

f <- ggplot(trajectory, aes(x=(dday-365)/365, y=WTD, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  ylab("WTD (cm)") + xlab("Year since restoration") +
  geom_hline(yintercept=0, linetype="dashed") +
  annotate("text", x=4, y=Inf, label="(f)", vjust=1, hjust=1) +
  theme_bw() + 
  theme(legend.position = "none", panel.grid.minor=element_blank(), 
        panel.grid.major.y = element_blank(), axis.title.y = element_text(size=9))

grid.arrange(a, b, c, d, e, f, ncol=1)

```

## Soil Properties
```{r soils, echo=FALSE, include=FALSE}
source("R/soils.R") #process soils data into summary dataframe

#Computation of signficant differences among sites. These output are reported by hand in the following section, where significance is reported at 0.05, 0.001, and 0.0001 levels for all values where P < 0.05. In all cases, the maximum P value is reported when multiple Tukey HSD comparisons are cited.  Data were normally distrubuted within each site/horizon combination, so ANOVA and Tukey's were used.

#Accreted comparisons
accreted <- subset(soils, horizon == "Accreted")

hclfe_A <- aov(accreted$FeT_mg.g ~ accreted$site) #hcl fe
summary(hclfe_A); TukeyHSD(hclfe_A)

feII_A <- aov(accreted$FeII_mg.g ~ accreted$site) #feII
summary(feII_A); TukeyHSD(feII_A)

pcfe_A <- aov(accreted$FePC_mg.g ~ accreted$site) #fe PC
summary(pcfe_A); TukeyHSD(pcfe_A)

oxfe_A <- aov(accreted$percent_ox ~ accreted$site) #percent oxidized
summary(oxfe_A); TukeyHSD(oxfe_A)

pH_A <- aov(accreted$pH ~ accreted$site) #pH
summary(pH_A); TukeyHSD(pH_A)

percentC_A <- aov(accreted$C ~ accreted$site) #percent C
summary(percentC_A); TukeyHSD(percentC_A)

## Parent comparisons
parent <- subset(soils, horizon == "Parent")

hclfe_P <- aov(parent$FeT_mg.g ~ parent$site) #hcl fe
summary(hclfe_P); TukeyHSD(hclfe_P)

feII_P <- aov(parent$FeII_mg.g ~ parent$site) #feII
summary(feII_P); TukeyHSD(feII_P)

pcfe_P <- aov(parent$FePC_mg.g ~ parent$site) #fe PC
summary(pcfe_P); TukeyHSD(pcfe_P)

oxfe_P <- aov(parent$percent_ox ~ parent$site) #percent oxidized
summary(oxfe_P); TukeyHSD(oxfe_P)

pH_P <- aov(parent$pH ~ parent$site) #pH
summary(pH_P); TukeyHSD(pH_P)

percentC_P <- aov(parent$C ~ parent$site) #percent C
summary(percentC_P); TukeyHSD(percentC_P)

percentN_P <- aov(parent$N ~ parent$site) #percent C
summary(percentN_P); TukeyHSD(percentN_P)
```

Extractable Fe concentrations were significantly higher at the alluvium wetland compared in the recently accreted (Tukey's HSD, *P* < 0.0001) and parent soil horizons (Tukey's HSD, *P* < 0.001) (Fig. 3a). Extractable Fe concentrations trended lower in the accreted compared to parent horizon at both wetlands, but these differences were not significant (Fig. 3a). Appreciable quantities of Fe^3+^ were present in all site horizons, where `r mean(soils$percent_ox[which(soils$site == "Alluvium")])`% and `r mean(soils$percent_ox[which(soils$site == "Peat")])`% of extractable Fe was in Fe^3+^ form at the alluvium and peat sites, respectively. Most Fe was in the reduced Fe^2+^ form across all sites, and statistical trends in Fe^2+^ concentrations were similar to total extractable Fe (Fig. 3b). While total extractable Fe concentrations did not vary between horizons at the alluvium site, Fe^2+^ concentrations were significantly higher in the accreted horizon compared to its underlying parent soil (Welch's t, *P* < 0.001).

Poorly crystalline Fe oxide concentrations were roughly two orders of magnitude lower than HCl-extractable Fe concentrations across both sites (Fig. 3c), and trends were similar to HCl-extractable Fe. Poorly crystalline Fe concentrations were significantly higher at the alluvium site both soil horizons (Tukey's HSD, *P* < 0.0001) (Fig. 3c). Poorly crystalline Fe concentrations did not vary between horizons at either wetland (Fig. 3c). Mean poorly crystalline Fe concentrations were roughly an order of magnitude higher in alluvium soils (`r mean(soils$FePC_mg.g[which(soils$site == "Alluvium")])` $\pm$ `r sd(soils$FePC_mg.g[which(soils$site == "Alluvium")])` mg Fe g^-1^ soil; *n* = `r sum(!is.na(soils$FePC_mg.g[which(soils$site == "Alluvium")]))`) compared to the peat soils (`r mean(soils$FePC_mg.g[which(soils$site == "Peat")])` $\pm$ `r sd(soils$FePC_mg.g[which(soils$site == "Peat")])` mg Fe g^-1^ soil; *n* = `r sum(!is.na(soils$FePC_mg.g[which(soils$site == "Peat")]))`) (Fig. 3c).

Soil pH also varied across the wetlands (ANOVA, *P* < 0.05), where alluvium soils were acidic (`r mean(soils$pH[which(soils$site == "Alluvium")])` $\pm$ `r sd(soils$pH[which(soils$site == "Alluvium")])`; *n* = `r sum(!is.na(soils$pH[which(soils$site == "Alluvium")]))`) and peat soils were near-neutral (`r mean(soils$pH[which(soils$site == "Peat")], na.rm=T)` $\pm$ `r sd(soils$pH[which(soils$site == "Peat")], na.rm=T)`; *n* = `r sum(!is.na(soils$pH[which(soils$site == "Peat")]))`) (Fig. 3d). pH did not vary by depth at the peat wetland; however, the parent horizon was significantly more acidic at the alluvium wetland (Welch's t, *P* < 0.0001) (Fig. 3d).

Parent horizon %C did not vary between the peat and alluvium wetlands (Tukey's HSD, *P* = 0.50) (Fig. 3e). Significant differences in %C were observed in the accreted horizon (Tukey's HSD, *P* < 0.0001), where %C was higher at the peat site (`r mean(soils$C[which(soils$site == "Peat")], na.rm=T)` $\pm$ `r sd(soils$C[which(soils$site == "Peat")], na.rm=T)`%; *n* = `r sum(!is.na(soils$C[which(soils$site == "Peat" & soils$horizon == "Accreted")]))`) than the alluvium site (`r mean(soils$C[which(soils$site == "Alluvium")])` $\pm$ `r sd(soils$C[which(soils$site == "Alluvium")])`%; *n* = `r sum(!is.na(soils$C[which(soils$site == "Alluvium" & soils$horizon == "Accreted")]))`). Soil %C was higher in the recently accreted horizon at the peat site (Welch's t, *P* < 0.001), but no differences in C content were observed between horizons at the alluvium wetland (Fig. 3e). Similar patterns were observed for soil %N across sites; however, parent horizon %N was significantly lower at the alluvium site (Tukey's HSD, *P* = 0.01) (Fig. 3f).

```{r fig2, fig.cap="Soil (a) HCl-extractable iron (Fe), (b) ferrous iron (Fe^2+^), (c) poorly crystalline Fe oxides (Fe~ca~), (d) pH, and (e) percent carbon (%C) in accreted and parent horizons across the three core wetland sites (*n* $\\geq$ 12 per depth). Panel (f) shows average daily methane fluxes (*F~CH4~*) for the early growing season (April-July) first year post-restoration and 2017 to represent *F~CH4~* when only the parent horizon was present and after accreted soil had developed, respectively. Solid horizontal lines are medians, boxes are the interquartile range, whiskers are the 95% confidence interval, and points denote measured outliers. Significance levels are shown for pairwise comparisons between soil horizons at each site (Welch's t-test), where ns denotes no significant differences and increasing asterisks denotes significance levels from *P* < 0.05, 0.01, 0.001, to 0.0001.", fig.width=8, fig.height=4, dpi=600}
source("R/soils.R") #process soils data into summary dataframe
#theme
boxplot_theme <- theme(axis.text.x = element_text(size=8),
                 axis.title.y = element_text(size=9),
                 axis.title.x = element_blank())

#add panel showing change from 1st year to 2017 alluvium fluxes
new_wetlands$horizon <- ifelse(new_wetlands$rest_year == 0, "Parent",
                            ifelse(new_wetlands$year == 2017, "Accreted",
                            NA))

new_wetlands2 <- new_wetlands[!is.na(new_wetlands$horizon),]

a <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=FeT_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Extractable Fe (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,32)) +
  annotate("text",x=0.6,y=32,label="(a)") +
  boxplot_theme + theme(legend.position="none")

b <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=FeII_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Fe2+ (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,25)) +
  annotate("text",x=0.6,y=25,label="(b)") +
  boxplot_theme + theme(legend.position="none")

c <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=FePC_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression(Fe[ca]~" (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,0.5)) +
  annotate("text",x=0.6,y=0.5,label="(c)") +
  boxplot_theme + theme(legend.position="none")

d <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=pH, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(4.5,7.6)) +
  ylab(expression("pH")) +
  annotate("text",x=0.6,y=7.6,label="(d)") +
  boxplot_theme +
  theme(legend.position=c(.79,.2), legend.title = element_blank(), 
        legend.margin = margin(0, 0, 0, 0), 
        legend.background = element_rect(fill = "white"))

e <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=C, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression("%C")) +
  scale_y_continuous(limits=c(10,30)) +
  annotate("text",x=0.6,y=30,label="(e)") +
  boxplot_theme + theme(legend.position="none")

f <- ggplot(subset(soils, site != "Old Peat"), aes(x=site, y=N, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression("%N")) +
  #scale_y_continuous(limits=c(10,30)) +
  annotate("text",x=0.6,y=1.6,label="(f)") +
  boxplot_theme + theme(legend.position="none")

# f <- ggplot(subset(new_wetlands2, month < 8 & month > 5), aes(x=site, y=mgCH4, fill=horizon)) +
#   geom_boxplot() +
#   stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
#   ylab(expression(F[CH4]~" ("~mg~CH[4]-C~m^{-2}~d^{-1}~")")) + 
#   scale_y_continuous(limits=c(0,400)) +
#   annotate("text",x=0.6,y=400,label="(f)") +
#   boxplot_theme + theme(legend.position="none")

grid.arrange(a, b, c, d, e, f, ncol=3)
```




## Biophysical drivers of CH~4~ flux
During year one post-restoration, when differences in CH~4~ flux magnitudes between the peat and alluvium wetland were largest (Fig. 3a), CH~4~ flux patterns and their relations to biophysical drivers were notably different across sites (Fig. 4). At the peat wetland in year one (2012), CH~4~ fluxes peaked between 10:00 to 12:00 local time (LT) coinciding with the period of peak *NEE* (Fig. S3). Similar diel CH~4~ flux patterns were also observed from the old peat wetland, and these mid-day CH~4~ flux peaks were consistent across years (Fig. S3). Methane fluxes from the peat wetland were strongly coupled to plant processes across multiple time scales during year one (Fig. 4a,c,e). At the hourly time scale, CH~4~ flux shared most information with *ET* (Fig. 4a), and at the diel scale, dominant interactions were synchronous and related to plant processes, such as *ET*, *GEP*, and *NEE*, as well as wind direction (*WD*) (Fig. 4c). At the multiday time scale, CH~4~ flux only shared significant levels of information with *ET* and *WD* (Fig. 4e). In total, 51.5% of total CH~4~ flux variability occurred at the diel scale, 33.4% occurred the multiday scale, and 15.1% occurred at the hourly time scale (Fig. 4a,c,e).

![Mutual information (*I^R^*) shared between methane flux (*F~CH4~*) and biophysical variables at multiple time scales for the peat and alluvium wetland (a-f) one-year post-restoration and (g-l) during the 2016 growing season. Black bars indicate *I^R^* at zero time lag, and colored extension indicates maximum *I^R^* observed at the time lag illustrated by the color scale. Vertical grey bars represent 95% confidence intervals generated from 1000 Monte Carlo random walks. Italicized percentage inset on each panel indicates the percent of total wavelet variance occurring at each time scale. Biophysical variables include net ecosystem exchange (*NEE*), gross ecosystem photosynthesis (*GEP*), evapotranspiration (*ET*), air temperature (*T~a~*), air pressure (*PA*), friction velocity (*u$^\ast$*), wind direction (*WD*), soil temperature (*T~s~*), and water temperature (*T~w~*).](fig4.png)

Peat-alluvium wetland CH~4~ flux patterns and interactions one-year post-restoration (2015) were notably different from the peat wetland. Peat-alluvium wetland CH~4~ fluxes were dominantly coupled to physical processes at diel to multiday time scales, such as temperature and atmospheric pressure fluctuations (Fig. 4b,d,f), in contrast to the peat site where fluxes were primarily coupled to plant processes (Fig. 4a,c,e). At the hourly scale, CH~4~ fluxes also shared most information with *ET* (Fig. 4b). At the diel scale, dominant interactions were observed with *PA* and *T~w~* at no time lag, and with *T~a~* at a three hour lag (Fig. 4d). Here, wetland maximum CH~4~ fluxes were observed in the late afternoon during atmospheric pressure lows (17:00 to 19:00 LT), and this relationship displayed strong hysteresis (Fig. S4). This dominant diel *PA* interaction was not observed at the peat site (Fig. 4c). Strong interactions were also observed between CH~4~ flux and *GEP*/*NEE*/*ET*, but these interactions occurred at a four to five hour time lag (Fig. 4d). These interactions can be intuitively seen through the timing of fluxes, where peak *NEE* and *ET* were observed mid-day, while peak CH~4~ flux occurred many hours later in the afternoon (Fig. S3). At the multiday scale, alluvium CH~4~ fluxes only shared significant information with temperature variables (*T~a~* and *T~w~*) (Fig. 4f). Hourly scale CH~4~ flux variability was more dominant at the alluvium wetland in year one, as 57% of total CH~4~ flux signal variability occurred at the diel scale, 30.5% occurred at the hourly scale, and only 12.5% occurred at the multiday time scale (Fig. 4b,d,f).

By the 2016 growing season, CH~4~ flux patterns and interactions had converged across the recently restored wetland sites. Peak CH~4~ fluxes were observed at roughly 15:00 LT at both sites by 2016. This shift to an afternoon peak flux occurred in 2013 at the peat wetland, and patterns in flux timing were similar for each year thereafter (Fig. S3). Both alluvium and peat sites were quite different in terms of flux timing compared to the old peat site, where peak CH~4~ flux occurred around noon most years (Fig. S3). 

Scale-emergent controls of CH~4~ flux were also very similar between the alluvium and peat site during the 2016 growing season (Fig. 4). For both sites, hourly CH~4~ fluxes exhibited dominant synchronous couplings to *ET* and *u$^\ast$* (Fig. 4g,h), and at the diel scale, dominant interactions with CH~4~ flux included *T~a~*, *ET*, *GEP*, and *NEE* at both wetlands (Fig. 4i,j). By 2016, dominant synchronous *PA* interactions observed at the alluvium site (Fig. 4d) were no longer apparent, though *PA* shared similar levels of information with CH~4~ flux at a 5-6 hour time lag (Fig. 4j). Multiday scale interactions were also similar across sites, where significant interactions with temperature variables (*T~a~*, *T~w~*, and *T~s~*), as well as *ET*, were observed (Fig. 4k,l). There were some notable differences between sites at the multiday scale, as CH~4~ flux also shared information with *NEE* at the peat site (Fig. 4k), and CH~4~ flux shared information with *PA* at the alluvium site (Fig. 4l). At the peat wetland, most variability still occurred at the diel scale (48.1%), followed by the multiday (29.2%) and hourly (22.7%) time scales (Fig. 4g,i,k). Most variability at the alluvium wetland still occurred at the diel scale (44.6%), followed closely by the hourly scale (41.4%), and much less variability occurred at the multiday scale (14%) (Fig. 4h,j,l).

## What causes increased alluvium CH~4~ fluxes with time?
```{r model_fit_prediction}

#subset alluvium and peat datasets for model fitting; first three years post-restoration
alluvium_train <- alluvium_daily %>%
  filter(year < 2017) %>%
  ungroup() %>%
  select(mgCH4, PAR, Tair, mVPD, PA, u., WTD, gCO2, mET, gER, gGEP)

peat_train <- peat6_daily %>%
  filter(year < 2014) %>%
  ungroup() %>%
  select(mgCH4, PAR, Tair, mVPD, PA, u., WTD, gCO2, mET, gER, gGEP)

#cross validated model training where optimal model parameters are chosen, missing values are knn imputed
glmnet_alluvium <- train(mgCH4 ~ ., data = alluvium_train, method = "glmnet",
                  preProcess = c("knnImpute"),
                  trControl = trainControl(method = "cv"),
                  na.action = na.pass)

glmnet_peat <- train(mgCH4 ~ ., data = peat_train, method = "glmnet",
                  preProcess = c("knnImpute"),
                  trControl = trainControl(method = "cv"),
                  na.action = na.pass)

# extract prediction datasets
alluvium <- alluvium_daily %>%
  select(mgCH4, PAR, Tair, mVPD, PA, u., WTD, gCO2, mET, gER, gGEP, year)

peat <- peat6_daily %>%
  select(mgCH4, PAR, Tair, mVPD, PA, u., WTD, gCO2, mET, gER, gGEP, year)

# generate predictions and residuals; create column for training v. test
alluvium <- alluvium %>%
  na.omit() %>%
  filter(year < 2018) %>%
  mutate(period = factor(ifelse(year < 2017, "Training (< 3 yrs old)", "Test (> 3 yrs old)"),
                         levels=c("Training (< 3 yrs old)", "Test (> 3 yrs old)"))) %>%  
  add_residuals(glmnet_alluvium) %>%
  add_predictions(glmnet_alluvium)

peat <- peat %>%
  na.omit() %>%
  filter(year < 2017) %>% # prediction on 2017 are biased due to flushing disturbance
  mutate(period = factor(ifelse(year < 2014, "Training (< 3 yrs old)", "Test (> 3 yrs old)"),
                         levels=c("Training (< 3 yrs old)", "Test (> 3 yrs old)"))) %>%
  add_residuals(glmnet_peat) %>%
  add_predictions(glmnet_peat)
```

To determine whether long-term changes in CH~4~ flux with time at the alluvium wetland could be explained by common biophysical drivers, we constructed empirical models based on ten common biophysical drivers identified in @sturtevant_identifying_2016 and this study, including *PAR*, *T~a~*, *PA*, *u$^\ast$*, *WTD*, *NEE*, *GEP*, *ER*, and vapor pressure deficit. Cross-validated elastic net regularized linear models were used to prevent overfitting, and models were trained on the first three years of data at both sites. Model predictions were then compared to new data, when wetlands were over three years old, to assess whether variation in these common biophysical drivers could explain more recent observed patterns in CH~4~ flux. Tuned elastic net models explained `r  max(glmnet_alluvium$results$Rsquared)*100`% and `r  max(glmnet_alluvium$results$Rsquared)*100`% (*r^2^*) of the training data variance based on 10-fold cross-validation (alluvium tuning parameters - $$\alpha$$ = `r glmnet_alluvium$bestTune$alpha` & $$\lambda$$ = `r glmnet_alluvium$bestTune$lambda`; peat tuning parameters - $$\alpha$$ = `r glmnet_peat$bestTune$alpha` & $$\lambda$$ = `r glmnet_peat$bestTune$lambda`).

For the model training period (first three years post-restoration), residuals were evenly distrubuted around zero and displayed little positive or negative bias across both wetlands (Fig. 5). The peat wetland model performed well on the test data (> three years post-restoration) where model residuals were evenly distributed around zero (Fig. 5). The alluvium model (trained on 2013-2016 fluxes) did not perform well on 2017 data (test set). Modeled fluxes were substantially underestimated, as demonstrated by positively skewed residuals (`r median(alluvium$resid[alluvium$period == "Test (> 3 yrs old)"])` $\pm$ `r IQR(alluvium$resid[alluvium$period == "Test (> 3 yrs old)"])`; median $\pm$ interquartile range) in the test set (Fig. 5). These findings suggest that dominant biophysical drivers alone cannot account for recent increases in CH~4~ flux at the alluvium wetland.

Though our soil data was collected during a single sampling in summer 2017, we can subsitute depth (parent v. accretted horizons) for time to understand what changes in soil properties might explain these increases in CH~4~ flux that cannot be explained by dominant environmental drivers alone (Fig. 5). If we assume the parent horizon is indicative of soil properties at the time of restoration and the accreted horizon is indicative of the current state of wetland soils, we observe a signficantly decrease in soil acidity (Fig. 3d) and shift towards more reduced soil conditions over time, as seen by the significant shift towards more Fe in its reduced form (Fe^2+^) within the accreted soil horizon (Fig. 3b).

```{r fig5, fig.cap="Residual CH~4~ flux (*F~CH4~*; mg C m^-2^ d^-1^) after accounting for dominant environmental drivers at the recently restored peat and alluvium wetlands. Residuals are given for the model training period (first three years post-restoration) and test period (> three years post-restoration). Training periods were 2011-2013 and 2014-2016 for the peat and alluvium wetlands, respectively, and testing periods were 2014-2016 and 2017 for the peat and alluvium wetlands, respectively. 2017 was omitted from analysis at the peat site because an extended freshwater flushing disturbance occurred that cannot be accounted for by our measurement campaign. Models were fitted using elastic net regularized linear models usinf ten-fold cross validation.", fig.width=7, fig.height=3, dpi=600}
a <- ggplot(alluvium, aes(x=period, y=resid)) +
  geom_ref_line(h = 0) +
  geom_boxplot(outlier.shape = NA) +
  xlab("Alluvium Wetland") + ylab(expression("Residual "~F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) +
  scale_y_continuous(limits = c(-100,200))

b <- ggplot(peat, aes(y=resid, x=period)) +
  geom_ref_line(h = 0) +
  geom_boxplot(outlier.shape = NA) +
  xlab("Peat Wetland") +
  scale_y_continuous(limits = c(-100,200)) +
  theme(axis.title.y = element_blank())

grid.arrange(a, b, nrow=1)
```

## Implications for restored wetland GHG emissions
We observed a steady increase in annual CH~4~ flux from the alluvium wetland, and by 2017 annual CH~4~ fluxes from the alluvium wetland exceeded those from the recently restored peat wetland and an additional peat wetland ('old peat' wetland) restored in 1999 (Fig. 6a). For periods when the wetland was fully vegetated (2015+ or year 1+ on Fig. 2), this initially corresponded to neutral GHG emissions from the alluvium wetland (Fig. 6b); however, by 2017 annual GHG budgets were larger than both peat sites yet budgets from the peat and alluvium sites overlapped (Fig. 6b). These wetlands experienced similar climatic and hydrologic conditions over this time period (Fig. SX), and a more detailed summary of meteorologic conditions can be found in Eichelmann et al. (*in review*). 

Annual CH~4~ budgets from the peat and old peat wetland covaried and were not statistically different from 2014 to 2017 (Fig. 6a). Reduced CH~4~ fluxes from the old peat wetland were observed during 2013 when multiple water table drawdowns were observed (data not shown). Both peat wetlands were GHG sources from 2014 to 2016, and these sites were GHG neutral during 2017 due to reduced CH~4~ fluxes (Fig. 6) and high *NEE* (Fig. SX). Soil iron content was similar across the two peat wetlands, though soil %C was significantly higher at the old peat wetland compared to the peat site ()


```{r fig6, fig.cap="Annual CH~4~ flux (*F~CH4~*) and greenhouse gas budget (*GHG*) across the two recently restored peat and alluvium wetlands, as well as an additional 19 year old wetland restored on peat soils. Annual GHG balance is computed from annual *F~CH4~* and *NEE* budgets using the sustained global warming potential for CH~4~ [@neubauer_moving_2015].", fig.width=5, fig.height=4, dpi=600}
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.x = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

pd <- position_dodge(width = 0.0) #dodge parameter

all_annual$site <- factor(all_annual$site, levels=c("Alluvium", "Peat", "Old Peat"))
all_pst14 <- subset(all_annual, year > 2013)

# only show years when wetlands are fully vegetated (developed fluxes)
all_pst14$tCH4 <- ifelse(all_pst14$site == "Alluvium" & all_pst14$year == 2014, NA, all_pst14$tCH4)
all_pst14$GHGbalance <- ifelse(all_pst14$site == "Alluvium" & all_pst14$year == 2014, NA, all_pst14$GHGbalance)


a <- ggplot(all_pst14, aes(x=year, y=tCH4, color=site)) +
        geom_point(position = pd) + geom_line() +
        geom_errorbar(aes(ymin = tCH4 - ciCH4, ymax = tCH4 + ciCH4), position = pd, width = 0) +
        scale_x_continuous(breaks = c(2014, 2015, 2016, 2017)) +
        ylab(expression(F[CH4]~"("~g~C~m^{-2}~yr^{-1}~")")) + 
        time_series + theme(legend.title=element_blank(), legend.margin = margin(0, 0, 0, 0),
                            legend.position=c(.112, .45), legend.key.size =  unit(0.17, "in"), 
                            legend.text = element_text(size=8), 
                            legend.background = element_rect(fill = "transparent"))

b <- ggplot(all_pst14, aes(x=year, y=GHGbalance, color=site)) +
        geom_point(position = pd) + geom_line() +
        geom_errorbar(aes(ymin = GHGbalance - ciGHG, ymax = GHGbalance + ciGHG), 
                      position = pd, width = 0) +
        scale_x_continuous(breaks = c(2014, 2015, 2016, 2017)) +
        ylab(expression("GHG (g"~CO[2-eq]~m^{-2}~yr^{-1}~")")) + xlab("Year") +
        time_series + theme(legend.position = "none")

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth)

grid.arrange(iA, iB, ncol=1)
```

Annual CH~4~ fluxes from the alluvium wetland were significantly lower than either peat wetland for the two years when all sites were fully vegetated (2015 and 2016), and in both years, annual CH~4~ fluxes did not vary between the two peat wetlands (Table 1). In 2015, alluvium CH~4~ budgets were 48-50% the magnitude of either peat site (Table 1), and differences between sites were largest during the growing season months (Fig. 4a). Differences in daily CH~4~ flux between the alluvium and peat sites became less pronounced in 2016 (Fig. 4a), when the alluvium CH~4~ budget was 75-78% of either peat site (Table 1). These wetlands experienced similar climatic and hydrologic conditions over this time period, though *T~w~* was reduced at the old peat wetland. A more detailed summary of meteorologic conditions can be found in Eichelmann et al. (*in review*).

Methane fluxes from the alluvium site were the same magnitude or larger than either peat site by 2017 (Fig. 4a). Young peat CH~4~ fluxes were low in 2017, likely due to freshwater flushing and an insect outbreak that occurred. If we consider alluvium growing season fluxes from 2015 as indicative of CH~4~ emissions from the underlying parent soil only and fluxes from 2017 as indicative of CH~4~ emissions from the combined accreted and parent soil profile, we observe a significant increase in average daily CH~4~ flux between these time periods (Welch's t, *P* < 0.0001) (Fig. 2f). Increased alluvium CH4 flux across this time period also correspond to significant increases in Fe^2+^ (Welch's t, *P* < 0.05) and decreases in acidity (Welch's t, *P* < 0.0001) within the accreted alluvium soil horizon.

This pattern of initially lower, and later converged, CH~4~ flux from the alluvium wetland was not observed in ecosystem CO~2~ fluxes. The alluvium wetland was a significantly larger CO~2~ sink than the old peat in 2015 and the peat wetland in 2016 (Table 1), and exhibited similar levels of CO~2~ uptake throughout each growing season (Fig. 4b). Annual *ER* budgets were not significantly different across all three sites in 2015 and 2016 (Table 1), and annual *GEP* from the alluvium trended higher in 2015 and was significantly higher than the peat wetland in 2016 (Table 1). Daily *GEP* and *ER* for the alluvium wetland were similar to both peat sites in 2015 when differences in CH~4~ flux were largest (Fig. 4c,d). In the 2016 growing season, daily *NEE*, *GEP*, and *ER* from the peat wetland was considerably lower than the alluvium wetland despite maintaining larger CH~4~ fluxes over this same time period (Fig. 4b-d).

Differences in annual CH~4~ flux between the alluvium and two peat sites were reflected in annual GHG budgets. In both 2015 and 2016, the alluvium wetland was GHG neutral, while both peat wetlands were significant GHG sources (Table 1). Peat-alluvium GHG emissions were significantly lower than either peat site in both years, and GHG emissions did not vary between peat wetlands in 2015 (Table 1). In 2016, GHG emissions from the peat wetland were significantly higher than emissions from the other sites (Table 1), likely due to reduced growing season *GEP* (Fig. 4c). These reductions coincided with a strong increase in salinity at the peat wetland (data not shown).


# Discussion
Our results suggest that soil Fe content, a legacy of the pre-restoration landscape, influences ecosystem-scale CH~4~ emissions from restored wetlands. We observed substantially lower CH~4~ emissions from wetlands restored on alluvium soils compared to those restored on peat soils; however, CH~4~ flux magnitudes converged across wetland type three years post-restoration. These differences were not driven by variable climate or hydrologic forcing, as these wetlands experienced similar meteorologic conditions and all remained inundated year-round (Fig. 5). High soil Fe content in alluvium wetland soils likely explains the reduced CH~4~ fluxes relative to peat wetlands. Concentrations of multiple Fe species were higher in alluvium soils compared to peat soils, and Fe concentrations did not vary across peat sites (Fig. 2). These patterns in soil Fe content mirror differences in annual CH~4~ flux, as 2015 and 2016 annual CH~4~ fluxes from the alluvium wetland were significantly lower than either peat site, and did not vary between peat wetlands (Table 1). Annual *NEE* did not vary in a similar pattern across wetland sites (Table 1). Peat-alluvium CO~2~ fluxes were often similar to, or larger than, CO~2~ fluxes from the other peat sites (Table 1; Fig. 4), demonstrating that differences in annual CH~4~ flux were not due to covariation with other factors broadly affecting wetland GHG exchange. These differences in annual CH~4~ flux are relevant from a greenhouse gas management perspective, as the alluvium wetland was GHG neutral in both years, while the two peat wetlands were significant sources of CO~2~ equivalent emissions (Table 1). 

Peat-alluvium soil Fe concentrations were high, comparable to tropical forest soils where microbial Fe reduction is known to inhibit CH~4~ production. @yang_high_2015 observed mean Fe concentrations of 23.6 mg Fe g^-1^ soil in forest soils where @teh_suppression_2008 had previously documented suppression of methanogenesis by microbial Fe reduction. Peat-alluvium wetland Fe concentrations were similar (`r mean(soils$FeT_mg.g[which(soils$site == "Alluvium")])` $\pm$ `r sd(soils$FeT_mg.g[which(soils$site == "Alluvium")])` mg Fe g^-1^ soil; *n* = `r sum(!is.na(soils$FeT_mg.g[which(soils$site == "Alluvium")]))`), demonstrating that Fe concentrations were within the range where microbial Fe reduction inhibits methanogenesis. In contrast to upland humid tropical forests where most Fe is found in poorly crystalline form [@dubinsky_tropical_2010; @hall_reducing_2015; @yang_high_2015], poorly crystalline Fe concentrations were roughly two orders of magnitude lower than the HCl-extractable Fe pool across all three wetland sites. Poorly crystalline Fe pools are readily reducible by soil microbes [@hyacinthe_reactive_2006; @hall_reducing_2015], and depletion of these pools suggests high activity by microbial Fe reducers in wetland soils. Depleted reducible Fe pools in wetlands relative to upland systems is expected, as stable anoxic wetland conditions promotes extended Fe reducer activity and utilization of poorly crystalline Fe pools [@teh_suppression_2008; @dubinsky_tropical_2010]. Our observations of high Fe concentrations, low poorly crystalline Fe pools, and low CH~4~ fluxes at the alluvium wetland all suggest that Fe reduction is a dominant form of anaerobic respiration capable of inhibiting CH~4~ flux at this site. Further incubation-based experiments would allow for the direct identification of this mechanism.

Differences in wetland biophysical CH~4~ drivers one-year post-restoration also suggests that alluvium CH~4~ fluxes were inhibited by Fe presence. Peat-alluvium CH~4~ fluxes were decoupled from plant processes across multiple scales compared to the peat site. Peat CH~4~ fluxes were more dominantly coupled to plant processes, such as *GEP* and *ET* (Fig. 7), while alluvium CH~4~ fluxes were more dominantly coupled to physical transport (pressure pumping; Fig. S4) and temperature, which dictates CH~4~ production rates in bulk soil [@yvon-durocher_methane_2014]. This decoupling from plant-derived substrates (*GEP*) and transport pathways (*ET*) suggests that less CH~4~ was derived from plant exudates in the rhizosphere. Oxygen is released into wetland soils through plant roots, and the rhizosphere is typically an area of high CH~4~ oxidation [@van_der_nat_seasonal_1998; @laanbroek_methane_2010]. The wetland rhizosphere is also a hotspot of microbial Fe oxidation and reduction given the co-occurrence of oxic-anoxic conditions [@weiss_enumeration_2003]. For sites with higher Fe, such as the alluvium wetland, CH~4~ fluxes might be more decoupled from plant substrates (*GEP*) and transport pathways (*ET*) where rhizosphere methanogenesis is further inhibited by active Fe redox cycling [@laanbroek_methane_2010]. Such inhibition of methanogenesis has been observed in microcosm and incubation studies, where oxygen input via plant roots re-oxidizes ferrous Fe and further suppresses CH~4~ production in the root zone [@roden_organic_1996, @frenzel_rice_1999]. Anaerobic microbial re-oxidation of Fe coupled to NO~3~^-^ reduction is also known to occur in wetland sediments [@weber_anaerobic_2006], and may be relevant to Fe cycling in these wetlands if significant NO~3~^-^ enters the system from upslope agriculture.

Conversely, we might expect to see stronger couplings to plant processes for sites with lower soil Fe concentrations, as we observe for the peat wetland (Fig. 7). @sturtevant_identifying_2016 also demonstrated strong diel CH~4~ couplings to *ET* and *GEP* at the old peat wetland where soil Fe concentrations are low relative to the alluvium site (Fig. 7). @hatala_gross_2012 and @knox_biophysical_2016 found that diel patterns in CH~4~ fluxes from Delta rice were driven by *GEP*, rather than temperature, because peak CH~4~ flux lagged *GEP* by ~1-2 hrs and led maximum soil temperature. We saw a very different dynamic at the alluvium wetland where peak CH~4~ flux occurred late in the afternoon, many hours after peak *GEP*, *ET*, and *T~a~*, and was instead synchronously coupled to *PA* and *T~w~* fluctuations (Fig. 7). These dynamics further suggest decoupling from recent plant-derived C substrates and transport pathways at the alluvium wetland, with more dominant couplings to physical transport and drivers of bulk soil methanogenesis, as expected if rhizosphere CH~4~ production was inhibited by active Fe redox.

Soil C concentrations also differed between Delta peat and alluvium soils [@soilsurvey], but we can rule out C concentrations as the dominant driver of CH~4~ flux differences for multiple reasons. First, soil C concentrations did not vary between parent soil horizons at the peat and alluvium wetlands (Fig. 2), yet we observed over two-fold higher CH~4~ fluxes from the peat wetland the first years following restoration (Fig. 3). Accreted soil horizons had not developed at this stage, and since parent soil C concentrations did not vary across sites it cannot explain CH~4~ flux differences. Secondly, we might expect to see reduced *ER* fluxes if C substrates were limited in alluvium soils, but instead, *ER* fluxes from the alluvium wetland were often similar to or larger than those from either peat wetland (Table 1; Fig. 4d). @mcnicol_separate_2014 demonstrated that heterotrophic respiration from these soils was reduced by 50% under flooded, anaerobic conditions compared to oxic conditions. We should therefore have been able to observe changes in the heterotrophic component of wetland *ER* flux because heterotrophic respiration was not fully suppressed under anoxic conditions. Similarly, *GEP* from the alluvium wetland was similar to or larger than *GEP* from either peat site (Table 1; Fig. 4c), demonstrating that methanogens were not limited by photosynthates at the alluvium wetland. Finally, CH~4~ fluxes did not scale with C concentrations at the peat sites. Recent studies suggest that CH~4~ production may be inhibited by humic substances in high peat soils [@minderlein_humic-rich_2010; @bridgham_methane_2013; @ye_soil_2016]; however, soil C concentrations were significantly higher in both horizons of the old peat site, but daily (Figs. 4a) to annual (Table 1) CH~4~ fluxes were equivalent between young and old peat wetlands. These combined observations demonstrate that soil %C neither limits or enhances ecosystem CH~4~ fluxes across these three wetland sites.

Wetland age is an important factor in that the CH~4~ flux differences between peat and alluvium wetlands appears to be transient, and observed differences become less apparent multiple years post-restoration once new wetland soil has accumulated (Figs. 3 and 4). This convergence is also observed in the biophysical drivers of CH~4~ flux, as the two sites had notably different controls at diel to multiday time scales one-year post-restoration but became more similar by the 2016 growing season (Fig. 7).

The question remains: Why did CH~4~ flux differences between the peat and alluvium wetlands disappear with time? We expect this is related to development of accreted soil horizons at the alluvium wetland, but further incubation studies are warranted to identify differences in CH~4~ production and consumption between parent and accreted soil horizons. We observed increases in reduced ferrous iron (Fe^2+^) and a shift toward less acidic conditions in alluvium accreted soil (Fig. 2). Increases in Fe^2+^ suggest more reduced conditions favorable to methanogenesis in the accreted soils, and acidic soils are known to inhibit methanogenesis [@dunfield_methane_1993; @ye_ph_2012] and alter methanogen community structure [@kotsyurbenko_shift_2007]. The interaction between less acidic and more reduced conditions could enhance CH~4~ production rates in the newly accreted soil. Contrary to our expectations, soil C and total Fe concentrations did not vary between parent and accreted horizons at the alluvium wetland (Fig. 2). This suggests that the accreted horizon was primarily a re-suspension of parent sediments. Over longer timescales we do expect to see reductions in Fe and increases of C within the accreted horizon, as we observed this trend in old peat wetland that where a deep peat layer has developed over 20 years. It may take over ten years for Fe concentrations to drop significantly, as only C concentrations varied signficantly between parent and accreted soil horizons at the seven year old, peat wetland (Fig. 2). Regardless, the increased alluvium CH~4~ flux is not likely due to substantial peat accumulation but is rather driven by less acidic, more reduced conditions present in the accreted soil horizon. 

Convergence of CH~4~ flux biophysical drivers also suggests that as soils develop with time CH~4~ fluxes become more similar across wetland types. The hierarchy of dominant biophysical drivers and lagged effects were quite similar across the peat and alluvium wetlands in 2016 (Fig. 7g-l) compared to the first year when flux-driver relationships were very different across the two systems (Fig. 7a-f). Here, the most notable shift occurred at the alluvium wetland, where CH~4~ fluxes were strongly coupled to physical processes in year one (Fig. 7b,d,f) and by 2016 fluxes were also coupled to plant processes across multiple scales, such as *NEE*, *GEP*, and *ET* (Fig. 7h,j,l). Differences in controls across scale are discussed in more detail in @sturtevant_identifying_2016, but we find short-term (hourly) variability in CH~4~ flux is influenced by transport mechanisms (*ET* and *u$^\ast$*), diel variation is more dictated by plant processes and temperature (*NEE*, *GEP*, *ET*, and *T~a~*), while weekly to monthly variation is driven by temperature oscillations. Surprisingly, much more flux variability occurred at the multiday scale for the peat wetland than the alluvium wetland, and this pattern was consistent over time (Fig. 7). While the driver of this inter-site variation is not clear, it demonstrates that eddy covariance is particularly well-suited to ecosystem CH~4~ flux measurements. Non-automated chamber measurement campaigns could easily under-sample important modes of CH~4~ flux variation, particularly if dominant modes of variation change across wetlands distributed over small spatial scales.

Overall, our findings demonstrate that site history and Fe content impacts ecosystem-scale wetland CH~4~ fluxes. It is well understood from numerous small-scale studies that microbial Fe reduction inhibits methanogenesis [@jackel_suppression_2000; @yao_effect_2000; @van_bodegom_direct_2004; @teh_suppression_2008; @jerman_wetland_2009; @zhou_methanogenesis_2014; @miller_methane_2015], but this work provides the first observational evidence of ecosystem-scale attenuation of CH~4~ flux by soil Fe. Our regional flux tower network approach allowed us to investigate the influence of spatially variable soil properties on ecosystem-scale emissions, while controlling for other common confounding variables, such as climate, hydrologic drivers, ecosystem productivity, plant community composition, and ecosystem age. These results demonstrate the power of observational network approaches to investigating process-level controls to ecosystem-scale greenhouse gas dynamics. Further incubation studies are warranted to identify the direct mechanisms affecting CH~4~ fluxes. These findings are likely widely relevant throughout wetlands in the Sacramento River drainage, as the Sacramento River drains the northern Sierra Nevada range where Fe concentrations are particularly high compared to the central or south Sierra Nevada [@graham_soil_2010]. Concentrations of Fe observed in alluvium wetland soils were similar to those in northern Sierra soils, further supporting fluvial deposition as the likely source of Fe [@graham_soil_2010]. 

We found that differences in CH~4~ flux between alluvium and peat wetlands were pronounced for the first few years following restoration, causing significant reductions in net GHG emissions from wetlands restored on alluvium soils. However, these differences disappeared with time, likely due to the development of more reduced, less acidic conditions favorable to CH~4~ production within accreted sediments. Wetland restoration projects have a lifetime of multiple decades (over 20 years for the old peat wetland), so these transient reductions in CH~4~ flux are likely of minor importance from a policy or management perspective because any GHG benefit of restoring wetlands on alluvium soil are lost within a few years of restoration. Regardless, this work provides evidence of ecosystem-scale attenuation of wetland CH~4~ flux by soil Fe content, which may be widely relevant to our understanding of wetland CH~4~ emissions from regions where soil Fe concentrations are high, such as the tropics and subtropics.

# Acknowledgements
This research was supported in part by the U.S. Department of Energys Office of Science, and its funding of Ameriflux core sites (Ameriflux contract 7079856), and the California Division of Fish and Wildlife, through a contract of the California Department of Water Resources (Award 4600011240).

# References

