---
title: "Manscript draft"
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Sara H Knox^4^, Jaclyn
  Hatala Matthes^5^, Daphne J Szutu^1^, Joseph G Verfaillie^1^, Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  pdf_document: default
  word_document: default
bibliography: library.bib
csl: global-change-biology.csl
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, Hayward, California, USA

^3^National Ecological Observatory Network, Boulder, Colorado, USA

^4^U.S. Geological Survey, Menlo Park, California, USA

^5^Department of Biological Sciences, Wellesley College, Wellesley, Massachusetts, USA

**Running Head**:

**Keywords**: test

**Type of Paper**: Primary research article

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
```

\newpage
# Abstract

# Introduction
Wetlands plays a fundamental role in regulating global climate by storing ~18% of Earth's soil carbon (C) [@lal_carbon_2008], emitting ~30% of global methane (CH~4~) emissions [@kirschke_three_2013; @saunois_global_2016], while only occupying 5-8% of the terrestrial land surface [@mitsch_wetlands_2013]. Most wetlands are greenhouse gas (GHG) sinks over multi-century time scales due to carbon dioxide (CO~2~) sequestration within anoxic soils, though they are commonly greenhouse gas (GHG) sources over shorter time scales due to the high global warming potential of emitted CH~4~ [@whiting_greenhouse_2001; @mitsch_wetlands_2013; @petrescu_uncertain_2015]. Wetland C sequestration and CH~4~ emissions are inextricably linked, as anoxic soil conditions inhibiting ecosystem respiration (*ER*) also activates archaeal CH~4~ production. The balance between these opposing fluxes primarily determines wetland GHG balances [@petrescu_uncertain_2015], and small differences in ecosystem CH~4~ emissions can induce large changes in wetland GHG exchange due to the > 28-fold radiative forcing of CH~4~ compared to CO~2~ over decadal to centennial time horizons [@pachauri_climate_2015]. Recent rises in atmospheric CH~4~ have been attributed to a global increase in wetland CH~4~ fluxes [@nisbet_rising_2016], and wetlands are currently the most uncertain component of the global CH~4~ budget [@kirschke_three_2013]. A key driver of this uncertainty is a lack of ecosystem-scale flux measurements to better constrain models and improve our understanding of how process-level controls scale to whole ecosystem CH~4~ emissions [@bridgham_methane_2013; @saunois_global_2016]. 
<!-- It is therefore important to target ecosystem flux measurement campaigns across environmental gradients known to influence CH~4~ fluxes from small-scale field and laboratory studies. -->

Methane fluxes from wetlands are governed by a number of interacting biophysical drivers including temperature, C inputs, alternative electron acceptor pools, and ecosystem inundation [@bridgham_methane_2013]. How these governing processes interact and scale to whole ecosystem fluxes if often highly uncertain due to sparse CH~4~ flux measurements across time and space [@petrescu_uncertain_2015], and current ecosystem measurement campaigns may under-sample relevant environmental gradients. Chamber flux syntheses have demonstrated that water table depth, temperature, vegetation, disturbance, and wetland type are important modulators of wetland CH~4~ flux [@turetsky_synthesis_2014]. A similar understanding of CH~4~ flux controls is emerging from eddy covariance studies, where temperature [@rinne_annual_2007; @olson_interannual_2013; @chu_net_2014], recent C inputs [@hatala_gross_2012; @morin_environmental_2014], wetland structure [@matthes_parsing_2014], and water table depth [@brown_evidence_2014; @goodrich_overriding_2015; @chamberlain_underlying_2015; @chamberlain_influence_2016; @sturtevant_identifying_2016; @chamberlain_impact_2017] have been identified as dominant controls across many ecosystems. Smaller-scale field and laboratory studies have also demonstrated the influence of soil C content [@bridgham_methane_2013; @ye_soil_2016] and alternative electron acceptor pools, such as sulfate and ferric iron (Fe^3+^), on wetland CH~4~ fluxes [@laanbroek_methane_2010; @poffenbarger_salinity_2011; @bridgham_methane_2013]; however, the importance of these factors has not been consistently demonstrated at ecosystem scale. Recently, @holm_ecosystem_2016 and @krauss_component_2016 observed attenuation of ecosystem CH~4~ fluxes with increasing salinity (i.e. sulfate concentrations), verifying the ecosystem-scale relevance of previous chamber-based observations [@poffenbarger_salinity_2011]. Iron (Fe) presence is also expected to inhibit ecosystem CH~4~ fluxes [@laanbroek_methane_2010], as observed in many laboratory studies [@jackel_suppression_2000; @van_bodegom_direct_2004; @jerman_wetland_2009], but Fe inhibition to ecosystem fluxes has not been documented. Understanding the relevance of these factors may be particularly important to constraining wetland CH~4~ emissions from the tropics, the largest source of the global wetland CH~4~ emissions [@kirschke_three_2013], where high soil iron concentrations are known to influence organic matter decomposition and CH~4~ production rates in forest soils [@dubinsky_tropical_2010; @hall_iron_2013].

Wetland restoration and management programs have increasingly been proposed and implemented as a means to mitigate climate change, and restoration strategies that minimize CH~4~ emissions and maximize CO~2~ uptake will provide the optimum climate benefit. These programs are particularly appealing in coastal ecosystems where CO~2~ sequestration rates are high and CH~4~ flux rates are low [@crooks_mitigating_2011; @conservation_international_blue; @poffenbarger_salinity_2011], and in drained peatlands where large CO~2~ emissions from oxidizing peat can be attenuated by re-flooding the landscape [@hatala_greenhouse_2012; @knox_agricultural_2015; @wilson_multiyear_2016]. A strong knowledge of CH~4~ flux drivers is essential to these programs because the magnitude of CH~4~ fluxes may determine if a restored wetland is a GHG source or sink [@knox_agricultural_2015], and CH~4~ emissions could affect the funding of wetland restoration programs if financed through carbon markets [@oikawa_evaluation_2017]. 

The Sacramento-San Joaquin Delta region of California (hereafter referred to as the Delta) is an area of extensive wetland restoration where financing through California's Cap and Trade program is being considered. The Delta is comprised of a network of artificially drained islands reclaimed for agriculture, where the water table is maintained below sea level by active pumping and levees surrounding each island. Drainage began in the 1850's to facilitate agricultural development, and exposure of the highly organic soils has led to substantial land surface subsidence [@deverel_historic_2010] and CO~2~ emissions [@hatala_greenhouse_2012; @knox_agricultural_2015]. Prior to drainage, this region was a 5 billion m^3^ tidal marsh where soil mineral and C content varied substantially. A major driver of this soil variation was fluvial deposition from major rivers, which carried mineral alluvium from the Sierra Nevada mountain range to the Delta marshes. Alluvium marshes frequently developed adjacent to peat marsh soils that were spatially removed from the main river pathways. Alluvium soils are more common on islands adjacent to the Sacramento River compared to the central and eastern Delta region where fluvial deposition was less pronounced [@deverel_historic_2010]. These differences in pre-drainage wetland geomophology give rise to a contemporary agricultural landscape where large changes in soil carbon and mineral content occur over small spatial scales [@soilsurvey], but how this edaphic variation influences current GHG fluxes from restored wetlands is unknown. Understanding the importance of these legacy effects is highly desirable, as it would allow for more informed wetland restoration strategies where restoration sites could be chosen to minimize GHG emissions.

The objectives of this study were to (1) determine how wetland CH~4~ and GHG budgets varied between peat and alluvium restored wetlands, (2) determine how soil properties varied across wetland types and which properties most likely caused observed flux differences, (3) identify if relationships between biophysical drivers and CH~4~ fluxes varied across wetland types, and (4) determine whether observed flux differences were economically relevant in a carbon market. We measured ecosystem-scale CH~4~ and CO~2~ fluxes by eddy covariance from three restored wetlands for 3+ years, where two wetlands were restored on peat soils and one wetland was restored on alluvium soil. We then used a combination of wavelet analysis and information theory to identify scale-emergent biophysical drivers of wetland CH~4~ fluxes and how they varied across the three wetlands, a technique previously applied by @sturtevant_identifying_2016 to identify scale-emergent CH~4~ flux controls for the two peat wetlands. We then measured multiple soil properties within the tower footprints, including soil C, nitrogen (N), ferrous iron (Fe^2+^), Fe^3+^, and poorly crystalline Fe, to identify potential drivers of observed flux differences. These tower sites are an ideal network to isolate these controlling factors, as all wetlands sites have a similarly managed hydroperiod, are within 13 km of one another, experience the same climate, and have similar plant community compositions. To further control potential covariates, we also used two temporary flux tower sites during summer 2017 to measure CH~4~ fluxes from a restored wetland complex where age is held constant but soil type varies across space. Finally, we calculated potential economic benefits of restoring wetlands on peat versus alluvium soils using the observed differences in our eddy covariance fluxes and current carbon market costs used in California's Cap and Trade Program [@carbon_dash].

# Materials and Methods
## Site Description
We measured wetland ecosystem CH~4~ and CO~2~ fluxes using a network of eddy covariance towers in the Sacramento-San Joaquin Delta region of California, USA.  The Delta is located within a Mediterranean climate that experiences hot, dry summers and cooler, rainy winters. The mean annual temperature is 15.1$^\circ$C (1998-2015 average), and the region receives 326 mm of rainfall annually [@knox_biophysical_2016]. All measurement sites were located on Sherman and Twitchell Islands in the northwest Delta. Both islands are adjacent to the Sacramento River and have typically lower carbon soils than central and eastern Delta islands within the San Joaquin River drainage [@deverel_historic_2010]. Sherman and Twitchell Islands are a mosaic of alluvium mollisols and deep histosols [@soilsurvey]. The alluvium mollisols formed via fluvial deposition from major rivers, and histosols formed densely organic soils in areas less disturbed by fluvial input [@deverel_historic_2010]. The dominant mollisol series on these islands are Gazwell and Scribner (Fig. 1), both of which are formed in mixed alluvium, belong to the soil class Cumulic Endoaquolls, and are found only on Delta islands within the Sacramento River drainage [@soilsurvey]. Rindge is the major histosol series found on both islands (Fig. 1) and is widely distributed across the Delta region. Rindge soils belong to the  Typic Haplosaprists class, and are characterized by deep, poorly drained marsh soils formed from decomposed plant organic matter [@soilsurvey]. 

Wetland ecosystem fluxes have been measured for over three years from three core wetland sites, where one site is restored on Scribner series and two sites are restored on Rindge series (Fig. 1). The Rindge series site on Sherman Island (Ameriflux site Us-Myb; N 38.0498, W 121.7650), hereafter referred to as the young peat site, was restored from a former pasture in October 2010, and ecosystem flux measurement began at this time. The Rindge series site on Twitchell Island (Ameriflux site Us-Tw1; N 38.1074, W 121.6469), hereafter referred to as the old peat site, was restored from a degraded agricultural site in 1997 and ecosystem flux measurements began in July 2012. The Scribner series site on Twitchell Island (Ameriflux site US-Tw4; N 38.10275, W 121.64125), hereafter referred to as the alluvium site, was restored in January 2014 on a former corn field and ecosystem flux measurements began at this time. All sites are vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*) and are actively managed to remain inundated year-round. Both the young peat and alluvium sites were constructed to have heterogeneous bathymetry, providing channels of open water and areas of closed vegetation. The old peat site is entirely closed vegetation and the ground surface is saturated tule and cattail detritus, whereas the newly restored young peat and alluvium sites have standing water under the plant canopy as less peat has accumulated compared to the old peat wetland. For a more detailed description the young and old peat sites see @miller_carbon_2011, @matthes_parsing_2014, and @knox_agricultural_2015.

Two additional temporary tower sites were installed within the same restored wetland complex as the alluvium wetland site, where wetland age is held constant and soil type varies across space. One temporary tower was installed in a Rindge soil (N 38.10607, W 121.64172) and the second tower was installed more centrally within the Scribner soil region (N 38.09876, W 121.64217) where the core alluvium tower is also found (Fig. 1). Soil characteristics indicative of peat versus alluvium soils were verified at these additional sites from observed differences in soil %C measured across 117 sample locations in this wetland complex prior to re-flooding in December 2013 (data not shown). Both temporary towers and the alluvium tower are installed within a single wetland complex that was restored in 2014. These sites are also vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*), are actively managed to remain inundated year-round, and are similar in structure to the young peat and alluvium core sites.

![Soil type and tower flux footprints for (a) Sherman and (b) Twitchell Islands in the Sacramento-San Joaquin Delta of California. Tower footprints include wetland area only, and footprint rings, from largest to smallest, correspond to the 90, 85, 80, 70, and 50% flux footprints at each site. Tower footprints were calculated using a two-dimensional analytical model described in @detto_soil_2006. The white tower footprint rings represent the core tower sites, and the footprint on (a) in Rindge series is the young peat site, the footprint on (b) in Rindge series is the old peat site, and the footprint on (b) in Scribner series is the alluvium site. The yellow footprints represent the temporary tower sites within the wetland complex that includes the core alluvium wetland site.](fig1.png)

## Ecosystem Flux and Environmental Measurements
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat were measured from the three core and two temporary wetland sites by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA). High-frequency data from the alluvium, old peat, young peat, and temporary peat site were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval, and high-frequency data from the temporary alluvium site were logged to CR1000 datalogger (Campbell Scientific, Logan, Utah, USA) at 10 Hz interval. Measurements took place at a height of 5.44m, 4.64m, 5.36m, 4.70m, and 3.88m for the young peat, old peat, alluvium, temporary alluvium, and temporary peat site, respectively. Corrections were applied to vertical wind speed data prior to flux processing for WindMaster Pro's affected by the 'w bug' that causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto_scaling_2010; @hatala_greenhouse_2012; @knox_agricultural_2015]. Flux corrections and quality control were applied as described in detail in @knox_agricultural_2015, @chamberlain_evaluation_2017, and Eichelmann et al. (*in review*), and included high-frequency data despiking, 2-D coordinate rotations, density corrections, and site-specific friction velocity (*u$^\ast$*) filtering. At the old peat site, we reject fluxes from wind directions 290$^\circ$-240$^\circ$ because fluxes from these directions were from other wetland types; however, we did not apply wind direction filtering to the other sites where flux footprints were more homogeneous. Flux footprints at all sites were calculated using a two-dimensional analytical model [@detto_soil_2006; @hsieh_approximate_2000]. We then filled all gaps in the half-hourly flux time series using artificial neural networks (ANNs) as described in detail in @knox_biophysical_2016 and Eichelmann et al. (*in review*). Separate ANNs were trained for daytime and nighttime CO~2~ fluxes, and the nighttime CO~2~ flux ANN was used to model *ER* at all times. Gross ecosystem photosynthesis (*GEP*) was then estimated as the difference between the gap-filled CO~2~ flux and modeled *ER* [@baldocchi_does_2015]. More detail on this partitioning strategy can be found in @baldocchi_does_2015 and @knox_biophysical_2016. We did not gap fill or partition fluxes for the temporary tower sites and only present observed fluxes from these sites.

We also measured a number of auxiliary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using shielded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*PA*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component and double-sided net-radiometers (CNR1 or NR Lite; Kipp and Zonen, Delft, Netherlands). We also measured soil (*T~s~*) and water (*T~w~*) temperature at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). An in depth description of the measurements made and instruments used at each core tower site can be found in Eichelmann et al. (*in review*).

## Soil Analyses
We measured soil C, N, and Fe content at the young peat, old peat, and alluvium sites in the recently accreted wetland O horizon, hereafter referred to as accreted horizon, and the underlying pre-restoration parent soils, hereafter referred to as the parent horizon. Fifteen samples of both accreted and parent soil were collected at each site across three transects with each sampling location at least 3m apart. The transects were within each tower's flux footprint (Fig. 1), all samples were collected from fully inundated locations, and the accreted sample was collected directly above the underlying parent soil sample. At the young peat and alluvium sites (both < 8 years old), the differentiation between these horizons was clear, as recently accreted horizon was loose, mucky, and heavily comprised of poorly-decomposed plant matter, whereas the underlying parent horizon was hardpack agricultural soil. At these sites, we collected the accreted horizon by hand and the top 15cm of parent horizon using a sediment core. Our sampling strategy was different at the 20 year old peat site because more than 0.5m of O horizon has accumulated since initial restoration. Here, the accreted horizon has accumulated to the water surface, and we collected the top 2 cm of this horizon to represent the most recently accreted peat. We then bore holes through the 0.5-0.7m saturated peat layer to collect the underlying parent soil. Differentiation between these layers was also clear, as the top 0.5-0.7m was comprised of poorly-decomposed plant matter and the underlying horizon was silty soil. All soil samples were immediately bagged, and sub-samples were field extracted in 0.5M HCl and 0.2M sodium citrate-0.05M sodium ascorbate solutions to measure HCl-extractable Fe (Fe^2+^ and Fe^3+^) and poorly crystalline Fe, respectively.


## Data Analysis
### Wavelet and Information Theory Analysis
We used a combination of wavelet decomposition and information theory to (1) isolate major time scales of variation within the continuous CH~4~ flux time series, and (2) identify scale-emergent interactions between CH~4~ fluxes and a number of biophysical drivers. Here, flux and environmental variable time series' were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [@percival_estimation_1995], where we isolate variation occurring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales corresponding to 2^n^ measurement intervals. The hourly time scale was reconstructed from dyadic scales 1-2 (1-2h), diel from dyadic scales 3-6 (4h to 1.3 days), and multiday from dyadic scales 7-10 (2.7-21.3 days). This approach is explained in greater detail in @sturtevant_identifying_2016, and our analysis used the same approach and processing software described therein.

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information is derived from Shannon entropy (*H*), a measure of uncertainty [@shannon1998mathematical], and *I* quantifies the amount of information shared between two variables. Mutual information and other information theory metrics, such as transfer entropy, are particularly useful for identifying relationships in complex systems because they do not assume linearity or other functional relationships and are capable of identifying asynchronous relationships [@ruddell_applying_2013]. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies for each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables:
$$H_x=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{x,y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Here, we bin our continuous data into discrete states to calculate *H~x~*, *H~y~*, and *H~x,y~*. We follow the same processing methodology in @sturtevant_identifying_2016 for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. Analyses for each year were confined to the growing season (DOY 100-300), and time series border periods were removed to avoid wavelet-induced distortions. We then calculated joint and marginal Shannon entropies using 10 fixed-interval bins for the probability distributions, and calculated *I* for a number of time lags and leads of *X*. We calculated *I* at lags and leads up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases, CH~4~ flux (*F~CH4~*) was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps described above. A more detailed description of this processing methodology can be found in @sturtevant_identifying_2016, and all wavelet decomposition and entropy calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R^*) where *I* is divided by the Shannon entropy of the Y variable (*F~CH4~*). 

# Results

```{r load_process}
#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat19_processing.R") #19yr old peat site
source("R/peat6_processing.R") #6yr old peat site
source("R/alluvium_processing.R") #alluvium site

#merge daily sums to single dataset for core sites
all_daily <- rbind(peat19_daily, peat6_daily); all_daily <- rbind(all_daily, alluvium_daily)
all_daily$date <- as.Date(all_daily$datetime) #data plotting variable

#load temporary tower sites
load("data/mobile_alluvium.Rdata")
load("data/mobile_peat.Rdata")
```

## Wetland Weather and Hydrology
The three wetland sites on Sherman and Twitchell Islands experienced similar meteorologic and hydrologic conditions for major biophysical drivers of CO~2~ and CH~4~ flux (Fig. 2). Incoming *PAR* and *T~a~* exhibited strong seasonality, where peak values were observed in summer months, and daily mean values were similar across all three wetland sites (Fig. 2a,b). Water temperature exhibited a similar seasonality across all three sites; however, *T~w~* was lower at the old peat compared to the two young wetland sites (Fig. 2c). Wetlands were similarly managed to maintain year-round inundation, where wetland *WTD* was maintained above the land surface across all years (Fig. 2d). In some cases, wetland water control structures would fail causing periods when the *WTD* would drop below the land surface. These events were common at the old peat wetland in summer 2013, when *WTD* dropped below the land surface 24 total days (Fig. 2d). A more detailed summary of meteorologic conditions across sites can be found in Eichelmann et al. (in review).

```{r fig2, fig.cap="Daily mean (a) photosynthetically active radiation (*PAR*), (b) air temperature (*T~a~*), (c) water temperature (*T~w~*), and (d) water table depth (*WTD*) for the alluvium, young peat, and old peat wetland sites. The dashed line on panel d represents the ground surface elevation at the water table well location, and the solid lines represents daily mean *WTD*.", warning=FALSE, fig.width=7, fig.height=5}
#themes
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#create plot objects
a <- ggplot(all_daily, aes(x=date, y=PAR, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(PAR~" ("~mu~mol~m^{-2}~s^{-1}~")")) + 
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

b <- ggplot(all_daily, aes(x=date, y=Tair, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[a]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")


c <- ggplot(all_daily, aes(x=date, y=Tw, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[w]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

d <- ggplot(all_daily, aes(x=date, y=WTD, color=site)) +
  geom_line(size=0.5) + theme_bw() + 
  geom_hline(yintercept=0, linetype="dashed") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab("WTD (cm)") +
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  time_series + theme(legend.title=element_blank(), legend.position=c(.09, .45),
                      legend.margin = margin(0, 0, 0, 0), 
                      legend.background = element_rect(fill = "transparent"))

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

## Soil Properties
```{r soils, echo=FALSE, include=FALSE}
source("R/soils.R") #process soils data into summary dataframe

#Computation of signficant differences among sites. These output are reported by hand in the following section, where significance is reported at 0.05, 0.001, and 0.0001 levels for all values where P < 0.05. In all cases, the maximum P value is reported when multiple Tukey HSD comparisons are cited. 

hcl_fe <- aov(soils_df$FeT_mg.g ~ soils_df$site*soils_df$horizon) #hcl fe
summary(hcl_fe); TukeyHSD(hcl_fe)

pc_fe <- aov(soils_df$FePC_mg.g ~ soils_df$site*soils_df$horizon) #poorly crystalline fe
summary(pc_fe); TukeyHSD(pc_fe)

ox_fe <- aov(soils_df$percent_ox ~ soils_df$site*soils_df$horizon) #percent oxidized
summary(ox_fe); TukeyHSD(ox_fe)

pH_aov <- aov(soils_df$pH ~ soils_df$site*soils_df$horizon) #percent oxidized
summary(pH_aov); TukeyHSD(pH_aov)
```

Extractable Fe concentrations were significantly higher at the alluvium site compared to both peat sites in the recently accreted (Tukey's HSD, *P* < 0.0001) and parent soil horizons (Tukey's HSD, *P* < 0.001) (Fig. 3a). At the alluvium site, extractable Fe concentrations did not vary with depth (Tukey's HSD, *P* = 0.6); however, extractable Fe concentrations were lower in the accreted horizon at both young peat (Tukey's HSD, *P* = 0.051) and old peat sites (Tukey's HSD, *P* < 0.001). Mean Fe concentrations were `r mean(soils_df$FeT_mg.g[which(soils_df$site == "Young Peat" & soils_df$horizon == "Parent")]) - mean(soils_df$FeT_mg.g[which(soils_df$site == "Young Peat" & soils_df$horizon == "Accreted")])` and `r mean(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Parent")]) - mean(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Accreted")])` mg Fe g^-1^ soil lower in accreted compared to parent horizons at the young and old peat sites, respectively (Fig. 3a). Extractable Fe concentrations were close to zero in the accreted horizon at the old peat site (median = `r median(soils_df$FeT_mg.g[which(soils_df$site == "Old Peat" & soils_df$horizon == "Accreted")])` mg Fe g^-1^), where the layer was predominantly poorly decomposed organic matter. Appreciable quantities of oxidized iron (Fe^3+^) were present in all other site horizons, and `r mean(soils_df$percent_ox[which(soils_df$site == "Alluvium")])`%, `r mean(soils_df$percent_ox[which(soils_df$site == "Young Peat")])`%, and `r mean(soils_df$percent_ox[which(soils_df$site == "Old Peat" & soils_df$horizon == "Parent")])`% of extractable Fe was in Fe^3+^ form at the alluvium site, young peat site, and parent horizon of the old peat site, respectively. The percentage of Fe in Fe^3+^ form did not vary between horizons at the young peat and alluvium sites (Tukey's HSD, *P* > 0.58).

Poorly crystalline Fe concentrations were roughly two orders of magnitude lower than HCl-extractable Fe concentrations across all three sites (Fig. 3b), and trends in poorly crystalline Fe were similar to extractable Fe. Poorly crystalline Fe concentrations were significantly higher in alluvium soils compared to young and old peat wetland soils (Tukey's HSD, *P* < 0.0001) and did not vary between peat sites (Tukey's HSD, *P* > 0.8). Poorly crystalline Fe concentrations trended lower in accreted horizons across all sites, but none of these differences were significant (Tukey's HSD, *P* > 0.08). Mean poorly crystalline Fe concentrations were over an order of magnitude higher in alluvium soils (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Alluvium")])` mg Fe g^-1^ soil) compared to the young (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Young Peat")])` mg Fe g^-1^ soil) and old (`r mean(soils_df$FePC_mg.g[which(soils_df$site == "Old Peat")])` mg Fe g^-1^ soil) peat soils (Fig. 3b).

Soil pH also varied across wetland sites (ANOVA, *P* < 0.05), where pH was lowest at the alluvium site (mean = `r mean(soils_df$pH[which(soils_df$site == "Alluvium")])`), intermediate at the old peat site (mean = `r mean(soils_df$pH[which(soils_df$site == "Old Peat")])`), and near-neutral at the young peat site (mean  = `r mean(soils_df$pH[which(soils_df$site == "Young Peat")], na.rm=T)`) (Fig. 3c). pH did not vary by depth at either the old or young peat wetlands (Tukey's HSD, *P* > 0.49); however, pH varied significantly with depth at the alluvium wetland (Tukey's HSD, *P* < 0.0001) where the parent horizon was more acidic (Fig. 3c).

```{r fig3, fig.cap="Soil (a) HCl-extractable and (b) poorly crystalline iron (Fe), as well as soil (c) pH and (d) percent carbon (%C) in accreted and parent sediments across the three wetland sites (*n* $\\geq$ 14 per depth). Solid horizontal lines are medians, boxes are the interquartile range (IQR), and whiskers are the 95% confidence interval. Points denote measured outliers.", warning=FALSE, fig.width=8, fig.asp = .3}
source("R/soils.R") #process soils data into summary dataframe
#theme
boxplot_theme <- theme(axis.text.x = element_text(size=9),
                 axis.title.y = element_text(size=10),
                 axis.title.x = element_blank())

a <- ggplot(soils_df, aes(x=site, y=FeT_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("HCl-extractable Fe (mg "~g^{-1}~"soil)")) +
  boxplot_theme +
  theme(legend.position=c(.82,.90), legend.title = element_blank(), 
        legend.margin = margin(0, 0, 0, 0), 
        legend.background = element_rect(fill = "white"))

b <- ggplot(soils_df, aes(x=site, y=FePC_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Poorly crystalline Fe (mg "~g^{-1}~"soil)")) +
  boxplot_theme + theme(legend.position="none")

c <- ggplot(soils_df, aes(x=site, y=pH, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("pH")) +
  boxplot_theme + theme(legend.position="none", axis.title.x = element_blank())

grid.arrange(a, b, c, nrow=1)
```

## Wetland CO~2~ and CH~4~ Flux Magnitudes
```{r combine_annual, echo=FALSE}
#Add age since restoration variable
peat19_yearly$age <- peat19_yearly$year - 1997
peat6_yearly$age <- peat6_yearly$year - 2011
alluvium_yearly$age <- alluvium_yearly$year - 2014

peat19_s <- subset(peat19_yearly, year != 2013) # remove year with draw down disturbance at WP

#merge to single dataset
all_sites <- rbind(peat19_s, peat6_yearly); all_sites <- rbind(all_sites, alluvium_yearly)

#remove incomplete years
all_sites <- subset(all_sites, Days >= 365)
```

```{r annual_stats, echo=FALSE, include=FALSE}
#Computation of significant differences among sites. These output are reported by hand in the following section, where significance is reported at 0.05, 0.001, and 0.0001 levels for all values where P < 0.05. In all cases, the maximum P value is reported when multiple Tukey HSD comparisons are cited. 
NEE <- aov(all_sites$tNEE ~ all_sites$site); summary(NEE) #NEE - ANOVA

GPP <- aov(all_sites$tGEP ~ all_sites$site); summary(GPP) #GPP - ANOVA

ER <- aov(all_sites$tER ~ all_sites$site); summary(ER) #ER - ANOVA

CH4 <- aov(all_sites$tCH4 ~ all_sites$site); summary(CH4) #FCH4 ANOVA
TukeyHSD(CH4) #Tukey HSD comparisons

GHG <- aov(all_sites$GHGbalance ~ all_sites$site); summary(GHG) #CO2-eq budget ANOVA
```

```{r fig4, fig.cap="Average annual fluxes of (a) net ecosystem exchange (*NEE*), (b) gross ecosystem photosynthesis (*GEP*),(c) ecosystem respiration (*ER*), and (d) methane flux (*F~CH4~*). Letters denote significant differences. Solid horizontal lines are medians, boxes are the interquartile range (IQR), and whiskers are the 95% confidence interval. Points denote measured outliers.", echo=FALSE, fig.width=3, fig.height=7}
#theme
box_theme <- theme(axis.title.x = element_blank(), 
                   axis.title.y = element_text(size=9),
                   axis.text.x = element_text(size=8))

a <- ggplot(all_sites, aes(x=site, y=tNEE)) +
  ylab(expression(NEE~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) + 
  geom_boxplot() + box_theme

b <- ggplot(all_sites, aes(x=site, y=abs(tGEP))) +
  ylab(expression(GEP~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) + 
  geom_boxplot() + theme(axis.title.x = element_blank(), axis.title.y = element_text(size=10)) + box_theme

c <- ggplot(all_sites, aes(x=site, y=tER)) +
  ylab(expression(ER~" ("~g~CO[2]-C~m^{-2}~yr^{-1}~")")) + 
  geom_boxplot() + box_theme

d <- ggplot(all_sites, aes(x=site, y=tCH4)) +
  ylab(expression(F[CH4]~" ("~g~CH[4]-C~m^{-2}~yr^{-1}~")")) + 
  geom_boxplot() + box_theme


grid.arrange(a, b, c, d, ncol=1)

```

Mean annual *NEE* budgets were not significantly different across the three wetland sites (ANOVA, *P* = 0.7), though all sites were CO~2~ sinks on average (Fig. 4a). Mean annual *NEE* was `r mean(alluvium_yearly$tNEE)` $\pm$ `r sd(alluvium_yearly$tNEE)` (*n* = `r length(alluvium_yearly$tNEE)`), `r mean(peat19_s$tNEE)` $\pm$ `r sd(peat19_s$tNEE)` (*n* = `r length(peat19_yearly$tNEE)`), and `r mean(peat6_yearly$tNEE)` $\pm$ `r sd(peat6_yearly$tNEE)` (*n* = `r length(peat6_yearly$tNEE)`) g CO~2~-C m^-2^ yr^-1^ for the alluvium, old peat, and young peat wetlands, respectively (mean $\pm$ SD). Similarly, mean annual *GEP* (ANOVA, *P* = 0.97) and *ER* (ANOVA, *P* = 0.9) did not vary across the three wetland sites (Fig. 4b,c). These averages comprise three complete years from the alluvium wetland, four years from the old peat wetland, and six years from the young peat wetland (Fig. 5). Ecosystem CO~2~ fluxes (*NEE*, *GEP*, and *ER*) followed similar seasonal patterns across all three wetlands, where peak daily fluxes were observed during summer and the lowest fluxes were observed January to February of each year (Fig. 5). Daily *NEE* for the three wetlands overlapped in range, particularly in 2015, though the young peat wetland had highly variable fluxes across years (Fig. 5a). At this site, large *GEP* and *ER* fluxes were observed one year post-restoration when the vegetation was completely established, and fluxes were attenuated mid-growing season in 2014 and 2016 due to insect outbreaks and salinization, respectively (Fig. 5b,c). Young peat CO~2~ fluxes were similarly reduced in the 2017 growing season coincident with another insect outbreak and large-scale flushing to reduce wetland salinity (Fig. 5a,b,c). CO~2~ fluxes from the old peat wetland were much more consistent across time (Fig. 5a,b,c), and once the alluvium wetland was fully vegetated (2015 onward), *NEE*, *GEP*, and *ER* fluxes from this system were very similar to the old peat wetland site (Fig. 5a,b,c). Additionally, CO~2~ flux patterns from the young peat and alluvium wetland were similar during the initial flooding phase when wetland vegetation cover was not fully established. Here, peak CO~2~ fluxes were observed late in the growing season when plant cover established within the flux footprints (Fig. 5a,b,c).

Mean annual CH~4~ fluxes varied significantly across sites (ANOVA, *P* < 0.01). Annual CH~4~ fluxes from the alluvium wetland were significantly lower than those from either peat site (Tukey's HSD, *P* < 0.05), and did not vary between young and old peat wetlands (Tukey's HSD, *P* = 0.92) (Fig. 4d). Annual CH~4~ fluxes were `r mean(alluvium_yearly$tCH4)` $\pm$ `r sd(alluvium_yearly$tCH4)` (*n* = `r length(alluvium_yearly$tCH4)`), `r mean(peat19_s$tCH4)` $\pm$ `r sd(peat19_s$tCH4)` (*n* = `r length(peat19_yearly$tCH4)`), and `r mean(peat6_yearly$tCH4)` $\pm$ `r sd(peat6_yearly$tCH4)` (*n* = `r length(peat6_yearly$tCH4)`) g CH~4~-C m^-2^ yr^-1^ for the alluvium, old peat, and young peat wetlands, respectively (mean $\pm$ SD). Methane fluxes from the old peat wetland were highly variable in 2013 due to multiple periods when *WTD* dropped below the land surface (24 days total), and CH~4~ fluxes reduced during these drained periods (Fig. 5d). Old peat wetland fluxes from 2013 were therefore omitted from the annual flux analysis above because they were not representative of stable state wetland fluxes. One similar draining event occurred in 2014, but CH~4~ fluxes recovered from this event and late growing season fluxes were similar to the young peat wetland. Methane fluxes from the two peat sites were very similar for 2015 and 2016; however, CH~4~ fluxes from the alluvium wetland were much lower than either peat site (Fig. 5d). Consequently, annual GHG equivalent budgets were higher from the old (`r mean(peat19_s$GHGbalance)` $\pm$ `r sd(peat19_s$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^) and young peat sites (`r mean(peat6_yearly$GHGbalance)` $\pm$ `r sd(peat6_yearly$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^) compared to the alluvium site (`r mean(alluvium_yearly$GHGbalance)` $\pm$ `r sd(alluvium_yearly$GHGbalance)` g CO~2~eq. m^-2^ yr^-1^); however, these differences were not statistically significant (ANOVA, *P* = 0.15).

```{r fig5, fig.cap="Daily (a) net ecosystem exchange (*NEE*), (b) gross ecosystem photosynthesis (*GEP*), (c) ecosystem respiration (*ER*), and (d) methane (*F~CH4~*) flux from the alluvium, old peat, and young peat wetlands. Flux measurements from the young peat and alluvium wetland begin at the point of restoration.", warning=FALSE, fig.width=7, fig.height=5.2}
#themes
times_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#plot objects, complete years only
a <- ggplot(all_daily, aes(x=date, y=gCO2, color=site)) +
  geom_line() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(NEE~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.position = "none")

b <- ggplot(all_daily, aes(x=date, y=gGEP, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(GEP~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.position = "none")

c <- ggplot(all_daily, aes(x=date, y=gER, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  scale_y_continuous(limits=c(0,15)) +
  ylab(expression(ER~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  times_series + theme(axis.text.x = element_blank(), legend.title=element_blank(),
                       legend.position=c(.08, .68), 
                       legend.margin = margin(0, 0, 0, 0),
                       legend.background = element_rect(fill = "transparent"))

d <- ggplot(all_daily, aes(x=date, y=mgCH4, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2017-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  times_series + theme(legend.position = "none")

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

Methane flux trajectories from the young peat and alluvium wetlands were quite different following initial restoration (Fig. 6). General flux patterns were similar across sites, where CH~4~ fluxes peaked late-season during year zero (days 0-365), and in the following growing seasons, peak CH~4~ fluxes occurred mid-summer (Fig. 6). Flux magnitudes were considerably lower from the alluvium site for the first three complete years following restoration (Fig. 6). Alluvium wetland fluxes increased with each year, whereas CH~4~ fluxes from the young peat wetland peaked during year one and decreased with subsequent years (Figs. 5d and 6). Differences in daily fluxes across wetlands were largest during year one when alluvium wetland fluxes were increasing and young peat wetland emissions were peaking (Fig. 6). Annual CH~4~ budgets for the young peat site and one were over two times larger than the alluvium site during year 0-1. Year zero budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2014]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2011]` g CH~4~ m^-2^ yr^-1^ for the alluvium and young peat wetland, respectively, and year one budgets were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2015]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2012]` g CH~4~ m^-2^ yr^-1^ for the alluvium and young peat wetland, respectively. This gap began to close during year two, when annual fluxes from the alluvium wetland were `r (alluvium_yearly$tCH4[alluvium_yearly$year == 2016])/(peat6_yearly$tCH4[peat6_yearly$year == 2013])*100`% the peat site, and were `r alluvium_yearly$tCH4[alluvium_yearly$year == 2016]` and `r peat6_yearly$tCH4[peat6_yearly$year == 2013]` g CH~4~ m^-2^ yr^-1^ from the alluvium and young peat wetland, respectively. During year three, fluxes from the young peat and alluvium wetlands were roughly equivalent, and the differences observed in the previous years were no longer apparent (Fig. 6). Similarly, CH~4~ fluxes from the old peat and alluvium wetlands were roughly equivalent during the 2017 growing season (Fig. 5d). Methane fluxes from the young peat wetland were lower and unstable for most of 2017, but reached a similar magnitude as the other sites by mid-summer (Fig. 5d). These temporarily reduced CH~4~ fluxes in 2017 were likely due to multiple sources of disturbance at the young peat site, namely insect outbreak and freshwater flushing to reduce salinity, also observed in the CO~2~ flux time series (Fig. 5).

```{r fig6, fig.cap="Methane fluxes (*F~CH4~*) from the young peat and alluvium wetland from time since initial restoration. Day zero is 2011-01-01 for the young peat wetland and 2014-01-01 for the alluvium wetland.", fig.width=3.5, fig.height=2, warning=FALSE}
#subset and index data
trajectory <- subset(all_daily, site != "Old Peat") #remove old peat site

#create a year since restoration column
trajectory$rest_year <- ifelse(trajectory$site == "Alluvium",  
                         trajectory$year - 2013, trajectory$year - 2010)

#compare first three years only (we only have 3.5 yrs of alluvium site)
trajectory <- subset(trajectory, rest_year < 5)

#Early restoration time series
ggplot(trajectory, aes(x=(dday-365), y=mgCH4, color=site)) +
  geom_point(aes(fill=site), colour="black", pch=21, size=0.9) +
  ylab(expression(F[CH4]~" "~"("~mg~C~m^{-2}~day^{-1}~")")) + 
  xlab("Day since restoration") +  theme_bw() + 
  scale_x_continuous(breaks=c(0, 365, 730, 1095)) +
  theme(legend.title=element_blank(), legend.position=c(.18, .82),
        legend.margin = margin(0, 0, 0, 0), legend.background = element_rect(fill = "transparent"),
        axis.title = element_text(size=10), panel.grid.minor=element_blank(), 
        panel.grid.major.y = element_blank())

```

```{r mobile_towers, include=FALSE}
#subset all Twitchell Island four year old sites and combine to single data set
alluvium4 <- select(alluvium_all, wm, datetime) #core alluvium site
alluvium4$site <- "Alluvium (core site)"
alluvium4 <- subset(alluvium4, datetime > "2017-06-30" & datetime <= "2017-08-15")

peat4 <- select(mobile_peat, wm, datetime) #mobile peat site
peat4$site <- "Peat (temporary)"
peat4 <- subset(peat4, datetime > "2017-06-30" & datetime <= "2017-08-15")

alluvium2_4 <- select(mobile_alluvium, wm, datetime)#mobile alluvium site
alluvium2_4$site <- "Alluvium (temporary)"
alluvium2_4 <- subset(alluvium2_4, datetime > "2017-06-30" & datetime <= "2017-08-15")

#combine to single dataframe
twitchell <- rbind(alluvium2_4, peat4); twitchell <- rbind(twitchell, alluvium4)

## summary statistics for comparisons across site. P reported at the 0.0001, 0.001, 0.01, 0.05 in text
#differences in flux magnitudes
wilcox.test(alluvium4$wm, alluvium2_4$wm, alternative="less")
wilcox.test(alluvium4$wm, peat4$wm, alternative = "less")

ks.test(alluvium4$wm, alluvium2_4$wm)
ks.test(alluvium4$wm, peat4$wm)

```

During the 2017 growing season, we also measured CH~4~ fluxes from two additional wetland locations within the same wetland complex as the alluvium site, where age was held constant (four years old) but the underlying soil properties varied (Fig. 1). Methane fluxes from the core and temporary alluvium site generally overlapped in range, and mean flux rates were `r mean(alluvium4$wm, na.rm=T)` $\pm$ `r sd(alluvium4$wm, na.rm=T)` and `r mean(alluvium2_4$wm, na.rm=T)` $\pm$ `r sd(alluvium2_4$wm, na.rm=T)` nmol CH~4~ m^-2^ s^-1^ (mean $\pm$ SD) from the core and temporary alluvium site, respectively (Fig. 7). Despite overlapping ranges, CH~4~ fluxes magnitudes were larger (Mann-Whitney U, *P* < 0.0001) and more broadly distributed (Kolmogorov-Smirnov, *P* < 0.0001) at the temporary alluvium site compared to the core alluvium site (Fig. 7). Mean CH~4~ fluxes from the temporary peat site were `r mean(peat4$wm, na.rm=T)` $\pm$ `r sd(peat4$wm, na.rm=T)` nmol CH~4~ m^-2^ s^-1^, much larger than fluxes from the core alluvium site (Mann-Whitney U, *P* < 0.0001), and their distributions exhibited little overlap in range (Kolmogorov-Smirnov, *P* < 0.0001) (Fig. 7). In general, CH~4~ fluxes from the core alluvium site were more similar to the temporary alluvium site than the temporary peat site (Fig. 7).

```{r fig7, fig.cap="Distribution of methane fluxes (*F~CH4~*) from the sites on Twitchell Island where wetland age is held constant and parent soil type varies. All measurements were made during the 2017 growing season between July 1 and August 15. All sites were restored in 2014. The temporary peat site measures fluxes from a Rindge peat soil series, and the temporary alluvium tower measures fluxes from a Scribner alluvium soil series similar to the alluvium core wetland site. Both temporary sites are within 500m of the alluvium core tower site.", fig.height=3, fig.width=3.5}
ggplot(data=twitchell, aes(wm, color=site, fill=site)) +
  geom_density(alpha=0.1) +
  xlab(expression(F[CH4]~" "~"("~nmol~CH[4]~m^{-2}~s^{-1}~")")) + 
  theme_bw() + 
  theme(legend.title=element_blank(), legend.position=c(.75, .88),
        legend.margin = margin(0, 0, 0, 0), legend.background = element_rect(fill = "transparent"),
        axis.title = element_text(size=10), 
        panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8))
```

## Methane flux patterns and biophysical drivers
During the first year post-restoration, when differences in CH~4~ flux magnitudes between the two newly restored wetlands were largest (Fig. 6), CH~4~ flux patterns and their relations to biophysical drivers were notably different across the two wetlands. At the young peat wetland in 2012, CH~4~ fluxes peaked between 10:00 to 12:00 local time (LT) coinciding with the period of peak *NEE* (Fig. S1). Similar diel CH~4~ flux patterns were also observed from the old peat wetland, and these mid-day CH~4~ flux peaks were consistent across years at the old peat site (Fig. S1). Methane fluxes from the young peat wetland were strongly coupled to plant processes (*NEE*, *GEP*, and *ET*) across multiple time scales during year one (Fig. 8a,c,e). At the hourly time scale, CH~4~ flux shared most information with *ET* (Fig. 8a), and at the diel scale, dominant interactions were synchronous and related to multiple plant processes, such as *ET*, *GEP*, and *NEE*, as well as wind direction (*WD*) (Fig. 8c). At the multiday time scale, CH~4~ flux only shared significant levels of information with *ET* and *WD* (Fig. 8e). 51.5% of total CH~4~ flux variability occurred at the diel scale, 33.4% occurred the multiday scale, and 15.1% occurred at the hourly time scale (Fig. 8a,c,e).

<!-- # ```{r fig8, fig.height = 5, fig.width = 3.5, fig.cap="Diel patterns of (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), and (c) evapotranspiration (*ET*) for the first year following restoration and 2016 for both recently restored wetlands. One year post-restoration was 2012 and 2015 for the young peat and alluvium wetland, respectively, when the wetlands were fully colonized by plants. By 2016 the young peat wetland was six years old and the alluvium wetland was three years old. Solid lines and shaded regions represent means $\\pm$ 1 standard deviation.", echo=FALSE} -->
<!-- # source("R/Diel_patterns.R") #process the data -->
<!-- #  -->
<!-- # #Compare first year fluxes and 2016. Indexing and subsetting (young peat and alluvium only) -->
<!-- # diel_s <- subset(diel_all, site == "Young Peat" | site == "Alluvium") -->
<!-- # diel_s$index <- ifelse(diel_s$site == "Young Peat" & diel_s$year == 2012, "Year 1", -->
<!-- #                        ifelse(diel_s$site == "Alluvium" & diel_s$year == 2015, "Year 1", -->
<!-- #                        diel_s$year)) -->
<!-- #  -->
<!-- # diel_s <- subset(diel_s, index == "Year 1" | index== "2016") # comparing most recent and 1st year -->
<!-- #  -->
<!-- # #turn index into factor for plotting -->
<!-- # diel_s$index = factor(diel_s$index, levels=c('Year 1', '2016')) -->
<!-- #  -->
<!-- # ##plotting -->
<!-- # #themes -->
<!-- # time_series <- theme_bw() + -->
<!-- #   theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(), -->
<!-- #         axis.title = element_text(size=10)) -->
<!-- #  -->
<!-- # #plot objects -->
<!-- # a <- ggplot(diel_s, aes(x=time, y=CH4, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   geom_ribbon(aes(x=time, ymin=CH4-sdCH4, ymax=CH4+sdCH4, group=site, fill=site), alpha=0.1) + -->
<!-- #   ylab(expression(F[CH4]~" ("~nmol^{-2}~s^{-1}~")")) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(legend.position = "none", axis.title.x = element_blank()) -->
<!-- #  -->
<!-- # b <- ggplot(diel_s, aes(x=time, y=NEE, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   ylab(expression(NEE~" ("~mu~mol^{-2}~s^{-1}~")")) +  -->
<!-- #   geom_ribbon(aes(x=time, ymin=NEE-sdNEE, ymax=NEE+sdNEE, group=site, fill=site), alpha=0.1) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(legend.position = "none", axis.title.x = element_blank(), -->
<!-- #                       strip.background = element_blank(), strip.text = element_blank()) -->
<!-- #  -->
<!-- # c <- ggplot(diel_s, aes(x=time, y=ET, color=site)) + -->
<!-- #   geom_line(size=0.5) + facet_grid(.~index) + -->
<!-- #   ylab(expression(ET~" ("~mmol^{-2}~s^{-1}~")")) + xlab("Time of day") + -->
<!-- #   geom_ribbon(aes(x=time, ymin=ET-sdET, ymax=ET+sdET, group=site, fill=site), alpha=0.1) + -->
<!-- #   scale_x_continuous(breaks=seq(0,24,5)) + -->
<!-- #   time_series + theme(strip.background = element_blank(), strip.text = element_blank(), -->
<!-- #                       legend.position = c(0.65,.85), legend.title = element_blank(), -->
<!-- #                       legend.margin = margin(0, 0, 0, 0), legend.key.size =  unit(0.15, "in"), -->
<!-- #                        legend.background = element_rect(fill = "transparent"),  -->
<!-- #                       legend.text = element_text(size=7)) -->
<!-- #  -->
<!-- # grid.arrange(a, b, c, ncol=1) -->
<!-- # ``` -->

Alluvium wetland CH~4~ flux patterns and interactions were notably different compared to the young peat wetland one year post-restoration. While CH~4~ fluxes from the young peat site were primarily coupled to plant processes across time scales, alluvium wetland CH~4~ fluxes were dominantly coupled to physical processes, such as temperature and atmospheric pressure fluctuations (Fig. 8b,d,f). At the hourly scale, CH~4~ fluxes also shared most information with *ET* (Fig. 8b). At the diel scale, dominant synchronous interactions were observed with *PA* and *T~w~*, and with *T~a~* at a three hour lag (Fig. 8d). Diel *PA* interactions were notably not observed across other sites (Fig. 8), and here we observe dominant non-linear interaction where peak CH~4~ flux occurs during late afternoon pressure lows (Fig. S2). This interaction helps explain why peak CH~4~ flux is observed so late in the day (17:00 to 19:00 LT) compared to the young peat site in year one (Fig. S1). Strong interactions were also observed between CH~4~ flux and *GEP*/*NEE*/*ET*, but these interactions occurred at a four to five hour time lag (Fig. 8d). This interaction can be intuitively seen in the timing of fluxes, where peak *NEE* and *ET* were observed mid-day, while peak CH~4~ flux occurred many hours later in the afternoon (Fig. S1). At the multiday scale, the alluvium site only shared significant information with temperature variables (*T~a~* and *T~w~*) (Fig. 8f). Hourly scale CH~4~ flux variability was more dominant at the alluvium wetland in year one, as 57% of total CH~4~ flux signal variability occurred at the diel scale, 30.5% occurred at the hourly scale, and only 12.5% occurred at the multiday time scale (Fig. 8).

![Mutual information (*I^R^*) shared between methane flux (*F~CH4~*) and biophysical variables at multiple time scales for the young peat and alluvium wetland one year post-restoration. Black bars indicate *I^R^* at zero time lag, and colored extension indicates maximum *I^R^* observed at time lag illustrated by the color scale. Vertical grey bars represent 95% confidence interval generated from 1000 Monte Carlo random walks. Italicized percentage inset on each panel indicates the percent of total wavelet variance occurring at each time scale. Biophysical variables include net ecosystem exchange (*NEE*), gross ecosystem photosynthesis (*GEP*), evapotranspiration (*ET*), air temperature (*T~a~*), air pressure (*PA*), friction velocity (*u$^\ast$*), wind direction (*WD*), soil temperature (*T~s~*), and water temperature (*T~w~*).](fig8.png)

By the 2016 growing season, CH~4~ flux patterns and interactions had converged across the recently restored wetland sites. Peak CH~4~ flux was observed at roughly 15:00 LT at both sites by 2016. This shift to an afternoon peak flux occurred in 2013 at the young peat wetland, and patterns in flux timing were similar for each year thereafter (Fig. S1). Both alluvium and young peat sites were quite different in terms of flux timing compared to the old peat site, where peak CH~4~ flux occurred around noon across most years (Fig. S1). 

Scale-emergent controls of CH~4~ flux were also very similar between the alluvium and young peat site by the 2016 growing season (Fig. 9). For both sites, hourly CH~4~ fluxes exhibited dominant synchronous couplings to *ET* and *u$^\ast$* (Fig. 9a,b), and at the diel scale, dominant interactions with CH~4~ flux included *T~a~*, *ET*, *GEP*, and *NEE* at both wetlands (Fig. 9c,d). By this time, strong synchronous *PA* interactions observed at the alluvium site in 2014 (Fig. 8d) were not apparent, and *PA* only shared similar levels of information at a 5-6 hour time lag (Fig. 10d). Interactions across sites were also similar at the multiday time scale, where significant interactions with temperature variables (*T~a~*, *T~w~*, and *T~s~*), as well as *ET*, were observed across both recently restored wetlands (Fig. 9e,f). There were some notable differences between sites at the multiday scale, as CH~4~ flux shared information with *NEE* at the young peat site (Fig. 9e), and CH~4~ flux shared information with *PA* at the alluvium site (Fig. 9f). At the young peat wetland, most variability still occurred at the diel scale (48.1%), followed by the multiday (29.2%), and hourly (22.7%) time scales. Most variability at the alluvium wetland still occurred at the diel scale (44.6%), followed closely by the hourly scale (41.4%), and much less variability occurred at the multiday scale (14%) (Fig. 9).

![Mutual information (*I^R^*) shared between methane flux (*F~CH4~*) and biophysical variables at multiple time scales for the young peat and alluvium wetland during the 2016 growing season. Black bars indicate *I^R^* at zero time lag, and colored extension indicates maximum *I^R^* observed at time lag illustrated by the color. Biophysical variable abbreviations are the same used in Fig. 8. Vertical grey bars represent 95% confidence interval generated from 1000 Monte Carlo random walks. Italicized percentage inset on each panel indicates the percent of total wavelet variance occurring at each time scale.](fig9.png)

# Discussion


# References

