---
title: "Manscript draft"
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Sara H Knox^4^, Jaclyn
  Hatala Matthes^5^, Daphne J Szutu^1^, Joseph G Verfaillie^1^, Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  pdf_document: default
  word_document: default
bibliography: library.bib
csl: global-change-biology.csl
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, Hayward, California, USA

^3^National Ecological Observatory Network, Boulder, Colorado, USA

^4^U.S. Geological Survey, Menlo Park, California, USA

^5^Department of Biological Sciences, Wellesley College, Wellesley, Massachusetts, USA

**Running Head**:

**Keywords**:

**Type of Paper**: Primary research article

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
```

\newpage
# Abstract

# Introduction
Wetlands plays a fundamental role in regulating Earth's climate by storing ~18% of Earth's 2500 Pg soil carbon (C) pool [@lal_carbon_2008], emitting ~30% of global CH~4~ emissions [@kirschke_three_2013; @saunois_global_2016], while only occupying 5-8% of Earth's land surface [@mitsch_wetlands_2013]. Most wetlands are greenhouse gas (GHG) sinks over multi-century time scales due to CO~2~ sequestration within anoxic soils, though they are commonly greenhouse gas (GHG) sources over shorter time scales due to the high global warming potential of emitted CH~4~ [@whiting_greenhouse_2001; @mitsch_wetlands_2013; @petrescu_uncertain_2015]. Wetland C sequestration and CH~4~ emissions are inextricably linked, as ecosystem respiration (*ER*) is inhibited by anoxic soil conditions that also activates archaeal CH~4~ production. The balance between these opposing fluxes primarily determines wetland GHG balances [@petrescu_uncertain_2015], and small differences in ecosystem CH~4~ emissions can induce large changes in wetland GHG exchange due to the > 28-fold radiative forcing of CH~4~ compared to CO~2~ over decadal to centennial time horizons [@pachauri_climate_2015]. Recent rises in atmospheric CH~4~ have been attributed to a global increase in wetland CH~4~ fluxes [@nisbet_rising_2016], and wetlands are currently the most uncertain component of the global CH~4~ budget [@kirschke_three_2013]. A key driver of this uncertainty is a lack of ecosystem-scale flux measurements to better constrain models and improve our understanding of how process-level controls scale to whole ecosystem CH~4~ emissions [@bridgham_methane_2013; @saunois_global_2016]. 
<!-- It is therefore important to target ecosystem flux measurement campaigns across environmental gradients known to influence CH~4~ fluxes from small-scale field and laboratory studies. -->

Methane fluxes from wetlands are governed by a number of interacting biophysical drivers including temperature, C inputs, alternative electron acceptor pools, and ecosystem inundation [@bridgham_methane_2013]. How these governing processes interact and scale to whole ecosystem fluxes if often highly uncertain due to sparse CH~4~ flux measurements across time and space [@petrescu_uncertain_2015], and current ecosystem measurement campaigns may under-sample relevant environmental gradients. Chamber flux syntheses have demonstrated that water table depth, temperature, vegetation, and wetland type are important modulators of wetland CH~4~ flux [@turetsky_synthesis_2014]. A similar understanding of CH~4~ flux controls is emerging from eddy covariance studies, where temperature [@rinne_annual_2007; @olson_interannual_2013; @chu_net_2014], recent C inputs [@hatala_gross_2012; @morin_environmental_2014], wetland structure [@matthes_parsing_2014], and water table depth [@brown_evidence_2014; @goodrich_overriding_2015; @chamberlain_underlying_2015; @chamberlain_influence_2016; @sturtevant_identifying_2016; @chamberlain_impact_2017] have been identified as dominant controls across many ecosystems. Chamber and laboratory studies have also demonstrated the influence of soil C content [@bridgham_methane_2013; @ye_soil_2016] and alternative electron acceptor pools, such as sulfate and ferric iron (Fe(III)), on wetland CH~4~ fluxes [@laanbroek_methane_2010; @poffenbarger_salinity_2011; @bridgham_methane_2013], but the importance of these factors has not been consistently demonstrated at ecosystem scale. Recently, @holm_ecosystem_2016 and @krauss_component_2016 observed attenuation of ecosystem CH~4~ fluxes with increasing salinity (i.e. sulfate), verifying the ecosystem-scale relevance of previous chamber studies [@poffenbarger_salinity_2011]. Iron (Fe) presence is also expected to inhibit ecosystem CH~4~ fluxes [@laanbroek_methane_2010], as observed in many laboratory studies [@jackel_suppression_2000; @van_bodegom_direct_2004; @jerman_wetland_2009], but Fe inhibition to ecosystem fluxes has not been observed. Understanding the relevance of such factors may be particularly important to constraining tropical wetland CH~4~ emissions, the largest component of the global wetland CH~4~ budget [@kirschke_three_2013], where soil iron concentrations are high and are known to influence organic matter decomposition and CH~4~ production in tropical forest soils [@dubinsky_tropical_2010; @hall_iron_2013].

Wetland restoration and management programs have increasingly been proposed and implemented as a means to mitigate climate change, and restoration strategies that minimize CH~4~ emissions and maximize CO~2~ uptake will provide the optimum climate benefit. These programs are particularly appealing in coastal ecosystems where CO~2~ sequestration rates are high and CH~4~ flux rates are low [@crooks_mitigating_2011; @conservation_international_blue; @poffenbarger_salinity_2011], and in drained peatlands where large CO~2~ emissions from oxidizing peat can be attenuated by re-flooding the landscape [@hatala_greenhouse_2012; @knox_agricultural_2015; @wilson_multiyear_2016]. A strong knowledge of CH~4~ flux drivers from restored wetlands is essential to these programs because the magnitude of CH~4~ fluxes may determine if restored wetlands are a GHG source or sink [@knox_agricultural_2015], and these CH~4~ emissions could affect the funding of wetland restoration programs if financed through carbon markets [@oikawa_evaluation_2017]. 

The Sacramento-San Joaquin Delta region of California (hereafter referred to as the Delta) is an area of extensive wetland restoration where financing through California's Cap and Trade program is being considered. The Delta is comprised of a network of artificially drained islands reclaimed for agriculture, where the water table is maintained below sea level by active pumping and levees surrounding each island. Drainage began in the 1850's to facilitate agricultural development, and exposure of the highly organic soils has led to substantial CO~2~ emissions [@hatala_greenhouse_2012; @knox_agricultural_2015] and land surface subsidence [@deverel_historic_2010]. Prior to drainage, this region was a 5 billion m^3^ tidal marsh where soil mineral and C content varied substantially. A major driver of this soil variation was fluvial deposition from major rivers, which carried mineral alluvium from the Sierra Nevada mountain range to the Delta marshes. Alluvium marshes frequently developed adjacent to peat marsh soils that were spatially removed from the main river pathways. Alluvium soils are more common on islands adjacent to the Sacramento River compared to the central and eastern Delta region where fluvial deposition was less pronounced [@deverel_historic_2010]. These differences in pre-drainage wetland geomophology give rise to a contemporary agricultural landscape where large changes in soil carbon and mineral content occur over small spatial scales [@soilsurvey], but how this edaphic variation influences current GHG fluxes from restored wetlands is unknown. Understanding the importance of these legacy effects is highly desirable, as it would allow for more informed wetland restoration strategies where restoration sites could be chosen to minimize GHG emissions.

The objectives of this study were to (1) determine how wetland CH~4~ and GHG flux budgets varied between peat and alluvium wetlands, (2) determine how soil properties varied across wetland types and which property most likely caused observed flux differences, (3) identify differences in dominant biophysical drivers across wetland types, and (4) determine whether observed flux differences were economically relevant in a carbon market. We measured ecosystem-scale CH~4~ and CO~2~ fluxes by eddy covariance from three restored wetlands for 3+ years, where two wetlands were restored on peat soils and one wetland was restored on alluvium soil. We then used a combination of wavelet analysis and information theory to identify scale-emergent biophysical drivers of wetland CH~4~ fluxes and how they varied across the three wetlands, a technique previously applied by @sturtevant_identifying_2016 to identify CH~4~ flux controls for the two peat wetlands. We then measured multiple soil properties within the tower footprints, including soil C, N, ferrous iron (Fe(II)), Fe(III), and poorly crystalline Fe, to identify potential drivers of observed flux differences. These tower sites are an ideal network to isolate these controlling factors, as all wetlands sites have a similarly managed hydroperiod, are within 13 km of one another, experience the same climate, and have a similar plant community composition. Finally, we calculated potential economic benefits of restoring wetlands on peat versus alluvium soils using the observed differences in our eddy covariance fluxes and current carbon market costs used in California's Cap and Trade Program [@carbon_dash].

# Materials and Methods
## Site Description
We measured wetland ecosystem fluxes of CH~4~, CO~2~, H~2~O, and sensible heat using a network of eddy covariance towers in the Sacramento-San Joaquin Delta region of California, USA.  The region is located within a Mediterranean climate that experiences hot, dry summers and cooler, rainy winters. The mean annual temperature is 15.1$^\circ$C (1998-2015 average), and the region receives 326 mm of rainfall annually [@knox_biophysical_2016]. All measurement sites were located on Sherman and Twichell islands in the northwest Delta region. Both islands are adjacent to the Sacramento River and have typically lower carbon soils than islands within the central and eastern Delta surrounding tributaries of the San Joaquin River [@deverel_historic_2010]. Both islands are a mosaic of alluvium mollisols and deep histosols [@soilsurvey]. The alluvium mollisols  likely formed via fluvial deposition from major rivers, and histosols formed densely organic soils in areas less disturbed by fluvial input [@deverel_historic_2010]. On these islands the dominant mollisol series are Gazwell and Scribner (Fig. 1), both of which are formed in mixed alluvium, belong to the soil class Cumulic Endoaquolls, and are found only on Delta islands within the Sacramento River drainage [@soilsurvey]. Rindge is the major histosol series found on both islands (Fig. 1) and is widely distributed across the Delta region. Rindge soils belong to the  Typic Haplosaprists class, and are characterized by deep and poorly drained organic soils formed in marshes from decomposed plant organic matter [@soilsurvey]. 

Wetland ecosystem fluxes have been measured for over three years from one wetland restored on Scribner series and two wetlands restored on Rindge series (Fig. 1). The Rindge series site on Sherman Island (Ameriflux site Us-Myb; N 38.0498, W 121.7650), hereafter referred to as the young peat site, was restored from a former pasture in October 2010, and ecosystem flux measurement began at this time. The Rindge series site on Twichell Island (Ameriflux site Us-Tw1; N 38.1074, W 121.6469), hereafter referred to as the old peat site, was restored from a degraded agricultural site in 1997 and ecosystem flux measurements began in July 2012. The Scribner series site on Twitchell Island (Ameriflux site US-Tw4; N 38.10275, W 121.64125), hereafter referred to as the alluvium site, was restored in January 2014 on a former corn field and ecosystem flux measurements began at this time. All sites are vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*) and are actively managed to remain inundated year-round. Both the young peat and alluvium sites were constructed to have heterogeneous bathymetry, providing channels of open water and areas of closed vegetation. The old peat site is entirely closed vegetation and the ground surface is saturated tule and cattail detritus, whereas the newly restored young peat and alluvium sites have standing water under the plant canopy as less peat has accumulated compared to the old peat site. For a more detailed description the young and old peat sites see @miller_carbon_2011, @matthes_parsing_2014, and @knox_agricultural_2015.

![Soil type and tower flux footprints for (a) Sherman and (b) Twitchell Islands in the Sacramento-San Joaquin Delta of California. Tower footprints include wetland area only, and footprint rings, from largest to smallest, correspond to the 90, 85, 80, 70, and 50% flux footprints at each site. All tower footprints measure wetland fluxes only, and were calculated using a two-dimensional analytical model described in @detto_soil_2006 The tower on Sherman Island (a) measures fluxes from the young peat wetland site, the tower on Rindge soil on Twitchell (b) measures fluxes from the old peat site, and the tower in the Scribner series footprint is the alluvium site.](fig1.png)

## Ecosystem Flux and Environmental Measurements
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat from the restored wetlands were measured by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval, and all high-frequency data were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA). Measurements took place at a height of 5.44m for the young peat wetland, 4.64m for the old peat wetland, and 5.36m for the alluvium wetland. Corrections were applied to vertical wind speed data prior to flux processing for WindMaster Pro's affected by the 'w bug' that causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto_scaling_2010, @hatala_greenhouse_2012, @knox_agricultural_2015]. Flux corrections and quality control were applied as described in detail in @knox_agricultural_2015, @chamberlain_evaluation_2017, and Eichelmann et al. (*in review*), and included high-frequency data despiking, 2-D coordinate rotations, density corrections, and friction velocity (*u$^\ast$*) filtering. At the old peat site, we reject fluxes from wind directions 290$^\circ$-240$^\circ$ because fluxes from these directions were from other wetland types; however, we did not apply wind direction filtering to the other sites where the flux footprints were more homogeneous. Flux footprints at all sites were calculated using a two-dimensional analytical model [@detto_soil_2006, @hsieh_approximate_2000]. We then filled all gaps in the half-hourly flux time series using artificial neural networks (ANNs) as described in detail in @knox_biophysical_2016 and Eichelmann et al. (*in review*). Separate ANNs were trained for daytime and nighttime CO~2~ fluxes, and the nighttime CO~2~ flux ANN was then used to model ecosystem respiration (*ER*) at all times. Gross ecosystem photosynthesis (*GEP*) was then estimated from the difference between gap-filled CO~2~ flux and modeled *ER* (@baldocchi_does_2015). More detail on this partitioning strategy can be found in @knox_biophysical_2016

We also measured a number of auxiliary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using shielded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*p~a~*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component net-radiometers (CNR1; Kipp and Zonen, Delft, Netherlands) and double-sided net radiometers (NR Lite; Kipp and Zonen, Delft, Netherlands). Double-sided net radiometers were used at the young peat site only. We also measured soil (*T~s~*) and water (*T~w~*) temperature (*T~s~*) at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). An in depth description of the measurements made at each site individually can be found in Eichelmann et al. (*in review*).

## Soil Analyses
We measured soil C, N, and Fe content at the young peat, old peat, and alluvium sites in the recently accreted O horizon and the underlying pre-restoration parent soils. Fifteen samples of both 0 horizon and parent soils were collected at each site across three transects with each sampling location at least 5m apart. The transects at each site were within the flux footprints of each eddy flux tower (Fig. 1), all samples were collected from fully inundated locations, and in the recently accreted O horizon was collected immediately above the underlying parent soil sample. At the young peat and alluvium sites (both < 8 years old), the differentiation between these horizons was clear, as recently accreted soil was loose, mucky, and heavily comprised of non-decomposed peat, whereas the underlying parent soils were hardpack agricultural soils. At these sites, we collected the O horizon soils by hand and collected the underlying parent soils using a sediment corer. Our sampling strategy was different at the old peat site because more than 0.5m of O horizon has accumulated since initial restoration at 20 year old site. Here, the recent O horizon has accumulated to the water surface, and we collected the top 2 cm of this horizon, representing the most recent peat. We then bore a hole through the 0.5-0.7m deep saturated peat layer to the underlying agricultural soil horizon, and collected this underlying soil horizon to represent the original parent agricultural soil. All soil samples were immediately bagged, and subsamples were extracted in 0.5M HCl and 0.2M sodium citrate-0.05M sodium ascorbate solutions to measure extractable Fe and poorly crystalline Fe, respectively.


## Data Analysis
### Wavelet and Information Theory Analysis
We used a combination of wavelet decomposition and information theory to (1) isolate major time scales of variation within the continuous CH~4~ flux time series, and (2) identify scale-emergent interactions between CH~4~ fluxes and a number of biophysical drivers. Here, flux and environmental variable time series' were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [@percival_estimation_1995], where we isolate variation occurring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales that correspond to 2^n measurement intervals. As such, the hourly time scale was reconstructed from scales 1-2 (1-2h), diel from dyadic scales 3-6 (4h to 1.3 days), and multiday from dyadic scales 7-10 (2.7-21.3 days). This approach is explained in greater detail in @sturtevant_identifying_2016, and our analysis used the same approach and processing software described therein.

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information is derived from Shannon entropy (*H*), a measure of uncertainty [@shannon1998mathematical], and *I* quantifies the amount of information shared between two variables. Mutual information, and similar information theory metrics such as transfer entropy, are particularly useful for identifying relationships in complex time series because these methods do not assume linearity or other functional relationships between interacting variables [@ruddell_applying_2013]. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies for each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables:
$$H_x=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{x,y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Since we are working with continuous data, we bin the data into discrete states for all variables. We follow the same processing methodology in @sturtevant_identifying_2016 for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. To calculate *H* and *I*, we then estimated probability distributions for *H* using 10 fixed-interval bins and calculated *I* for a number of time lags and leads of *X*. We calculated *I* at lags and leads up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases *F~CH4~* was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps described above. A more detailed description of this processing framework can be found in @sturtevant_identifying_2016, and all wavelet decomposition and *I* calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R*) where *I* is divided by the Shannon entropy of the Y variable, in this case *F~CH4~*. 

# Results

# References

