---
title: "Manscript draft"
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Sara H Knox^4^, Jaclyn
  Hatala Matthes^5^, Daphne J Szutu^1^, Joseph G Verfaillie^1^, Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  pdf_document: default
  word_document: default
bibliography: library.bib
csl: global-change-biology.csl
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, Hayward, California, USA

^3^National Ecological Observatory Network, Boulder, Colorado, USA

^4^U.S. Geological Survey, Menlo Park, California, USA

^5^Department of Biological Sciences, Wellesley College, Wellesley, Massachusetts, USA

**Running Head**:

**Keywords**:

**Type of Paper**: Primary research article

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
```

\newpage
# Abstract

# Introduction
Wetlands plays a fundamental role in regulating Earth's climate by storing ~18% of Earth's 2500 Pg soil carbon (C) pool [@lal2008carbon] and emitting ~30% of global CH~4~ emissions [@kirschke2013three, @saunois2016global], while only occupying 5-8% of Earth's land surface [@mitsch2013wetlands]. Most wetlands are greenhouse gas sinks (GHG) over century time scales due to long-term CO~2~ sequestration within anoxic soils [@mitsch2013wetlands], though they may also be greenhouse gas (GHG) sources over shorter time scales due to the high global warming potential of emitted CH~4~ [@mitsch2013wetlands, @petrescu2015uncertain]. The C sequestration and CH~4~ emission potential of wetlands is inextricably linked, as the anoxic soil conditions that attentuate ecosystem respiration (*ER*) also activates CH~4~ production by *Archaea*. The balance between these opposing fluxes primarily determines wetland GHG balances [@petrescu2015uncertain], and small differences in ecosystem CH~4~ emissions can induce large changes in wetland GHG budgets given the > 28-fold radiative forcing of CH~4~ compared to CO~2~ over decadal to centential time horizons [@pachauri_climate_2015]. Understanding the drivers of CH~4~ flux variation across wetlands is thus particularly important, as wetland CH~4~ fluxes are the most uncertain component of the global CH~4~ budget [@kirschke2013three], and recent rises in atmospheric CH~4~ concentrations have been attributed to a global increase in wetland CH~4~ fluxes [@nisbet2016rising]. A key driver in these uncertainties is a lack of ecosystem-scale flux measurements to better constrain models [@bridgham2013methane, @saunois2016global] and improve our understanding of how process-level controls scale to whole ecosystem CH~4~ emissions.

Methane fluxes from wetlands are governed by a number of interacting biophysical drivers including temperature, C inputs, alternative electron acceptor pools, and ecosystem inundation [@bridgham2013methane]. How these governing processes interact and scale to whole ecosystem fluxes if often highly uncertain due to sparse CH~4~ flux measurements across time and space [@petrescu2015uncertain] and current ecosystem measurement campaigns may undersample relevant environmental gradients. Chamber-based flux measurements have demonstrated water table depth, temperature, vegetation, and wetland type are important modulators of wetland CH~4~ flux [@turetsky_synthesis_2014]. Similarly, temperature [@knox2016biophysical], recent C inputs [@hatala_gross_2012], and water table depth [@brown_evidence_2014, @goodrich_overriding_2015, @chamberlain_underlying_2015, @chamberlain_influence_2016, @sturtevant2016identifying, @chamberlain_impact_2017] have been demonstrated as important controls to whole ecosystem emissions, as evidenced through continuous eddy covariance CH~4~ flux measurements. Other important controls, such as alternative electron acceptor pools (nitrate, sulfate, ferric iron), have been observed primarily in smaller-scale field or laboratory studies (CITATIONS), so our understanding of the relevance of these drivers to regulating ecosystem CH~4~ fluxes remains limited. Nitrate is not expected to repress methanogenesis due to its rapid uptake by plants [@laanbroek_methane_2010], but attentuated CH~4~ emissions have been observed in high sulfate coastal waterways by chamber-based [@poffenbarger_salinity_2011] and eddy covariance studies [@holm_ecosystem_2016, @krauss_component_2016]. Iron suppression of methanogenesis has been observed in small-scale laboratory studies [@jackel_suppression_2000, @jerman_wetland_2009, @van_bodegom_direct_2004], but these observations have not been made at the ecosystem-scale so the relevance of iron presence to ecosystem wetland CH~4~ fluxes is uncertain. This is a potentially important driver of wetland CH~4~ fluxes in areas of the tropics or fluvial deposition where soil iron concentrations may be high (CITATION).

Wetland restoration and management programs have increasingly been proposed and implemented as a means to mitigate climate change, and restoration strategies that minimize CH~4~ emissions and maximize CO~2~ uptake will provide the optimum climate benefit. These programs are particularly appealling in coastal ecosystems where CO~2~ sequestration rates are high and CH~4~ flux rates are low [@crooks_mitigating_2011, @conservation_international_blue, @poffenbarger_salinity_2011], and in drained peatlands where large CO~2~ emissions from oxidizing peat can be attentuated by reflooding the landscape. Wetland restoration from peatland agriculture in the Sacramento-San Joaquin Delta of California leads to a large reduction in the ecosystem GHG budget, even when accounting for CH~4~ emissions [@hatala2012greenhouse, @knox2015agricultural]. Here, wetland restoration turned drained peatlands from a strong GHG source to near GHG neutral wetlands, though annual CH~4~ fluxes varied significantly across wetlands determining if the system was a GHG source or sink [@knox2015agricultural]. Understanding the sources of this CH~4~ flux variation is important to developing better informed wetland restoration strategies, as well as identifying potential sources of uncertainty in wetland CH~4~ budgets.


The Delta region is currently comprised of a network of artificially drained islands reclaimed for agriculture, where the water table is maintained below sea level through active pumping and levees surrounding each island. Drainage began in the 1850's to facilitate agricultural development, and exposure of the highly organic soils has led to substaintial CO~2~ emissions [@hatala2012greenhouse, @knox2015agricultural] and land surface subsidence [@deverel2010historic]. Prior to drainage, this region was a 5 billion m^3^ tidal marsh where soil type and organic matter varied substantially over small spatial scales [@deverel2010historic]. A major driver of this soil variation is fluvial deposition from major rivers, such as the Sacramento river, which would carry mineral alluvium from the Sierra Nevada range [@deverel2010historic]. Alluvium soils are more common in islands adjacent to the Sacramento River, as compared to the central and eastern Delta region where fluvial deposition was less pronounced  [@deverel2010historic]. As result, central and eastern Delta islands are typically less heterogeneous and higher carbon than islands adjacent to the Sacramento River drainage [@deverel2010historic].

# Materials and Methods
## Site Description
We measured wetland ecosystem fluxes of CH~4~, CO~2~, H~2~O, and sensible heat using a network of eddy covariance towers in the Sacramento-San Joaquin Delta region of California, USA.  The region is located within a Mediterranean climate that experiences hot, dry summers and cooler, rainy winters. The mean annual temperature is 15.1$^\circ$C (1998-2015 average), and the region receives 326 mm of rainfall annually [@knox2016biophysical]. All measurement sites were located on Sherman and Twichell islands in the northwest Delta region. Both islands are adjacent to the Sacramento River and have typically lower carbon soils than islands within the central and eastern Delta surrounding tributaries of the San Joaquin River [@deverel2010historic]. Both islands are a mosaic of alluvium mollisols and deep histosols [@soilsurvey]. The alluvium mollisols  likely formed via fluvial deposition from major rivers, and histosols formed densely organic soils in areas less disturbed by fluvial input [@deverel2010historic]. On these islands the dominant mollisol series are Gazwell and Scribner (Fig. 1), both of which are formed in mixed alluvium, belong to the soil class Cumulic Endoaquolls, and are found only on Delta islands within the Sacramento River drainage [@soilsurvey]. Rindge is the major histosol series found on both islands (Fig. 1) and is widely distributed across the Delta region. Rindge soils belong to the  Typic Haplosaprists class, and are characterized by deep and poorly drained organic soils formed in marshes from decomposed plant organic matter [@soilsurvey]. 

Wetland ecosystem fluxes have been measured for over three years from one wetland restored on Scribner series and two wetlands restored on Rindge series (Fig. 1). The Rindge series site on Sherman Island (Ameriflux site Us-Myb; N 38.0498, W 121.7650), hereafter referred to as the young peat site, was restored from a former pasture in October 2010, and ecosystem flux measurement began at this time. The Rindge series site on Twichell Island (Ameriflux site Us-Tw1; N 38.1074, W 121.6469), hereafter referred to as the old peat site, was restored from a degraded agricultural site in 1997 and ecosystem flux measurements began in July 2012. The Scribner series site on Twitchell Island (Ameriflux site US-Tw4; N 38.10275, W 121.64125), hereafter referred to as the alluvium site, was restored in January 2014 on a former corn field and ecosystem flux measurements began at this time. All sites are vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*) and are actively managed to remain inundated year-round. Both the young peat and alluvium sites were constructed to have heterogenous bathymetry, providing channels of open water and areas of closed vegetation. The old peat site is entirely closed vegeation and the ground surface is saturated tule and cattail detritus, whereas the newly restored young peat and alluvium sites have standing water under the plant canopy as peat accumulation is less than the old site. For a more detailed description the young and old peat sites see @miller2011carbon, @matthes2014parsing, and @knox2015agricultural.

![Soil type and tower flux footprints for (a) Sherman and (b) Twitchell Islands in the Sacramento-San Joaquin Delta of California. Tower footprints include wetland area only, and footprint rings, from largest to smallest, correspond to the 90, 85, 80, 70, and 50% flux footprints at each site. All tower footprints measure wetland fluxes only, and were calculated using a two-dimensional analytical model described in @detto2006soil. The tower on Sherman Island (a) measures fluxes from the young peat wetland site, the tower on Rindge soil on Twitchell (b) measures fluxes from the old peat site, and the tower in the Scribner series footprint is the alluvium site.](fig1.png)

## Ecosystem Flux and Environmental Measurements
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat from the restored wetlands were measured by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval, and all high-frequency data were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA). Measurements took place at a height of 5.44m for the young peat wetland, 4.64m for the old peat wetland, and 5.36m for the alluvium wetland. Corrections were applied to vertical wind speed data prior to flux processing for WindMaster Pro's affected by the 'w bug' that causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto2010scaling, @hatala2012greenhouse, @knox2015agricultural]. Flux corrections and quality control were applied as described in detail in @knox2015agricultural, @chamberlain2017evaluation, and Eichelmann et al. (*in review*), and included high-frequency data despiking, 2-D coordinate rotations, density corrections, and friction velocity (*u$^\ast$*) filtering. At the old peat site, we reject fluxes from wind directions 290$^\circ$-240$^\circ$ because fluxes from these directions were from other wetland types; however, we did not apply wind direction filtering to the other sites where the flux footprints were more homogeneous. Flux footprints at all sites were calculated using a two-dimensional analytical model [@detto2006soil, @hsieh2000approximate]. We then filled all gaps in the half-hourly flux time series using artificial neural networks (ANNs) as described in detail in @knox2016biophysical and Eichelmann et al. (*in review*). Separate ANNs were trained for daytime and nighttime CO~2~ fluxes, and the nighttime CO~2~ flux ANN was then used to model ecosystem respiration (*ER*) at all times. Gross ecosystem photosynthesis (*GEP*) was then estimated from the difference between gap-filled CO~2~ flux and modelled *ER* (@baldocchi2015does). More detail on this partitioning strategy can be found in @knox2016biophysical.

We also measured a number of auxillary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using sheilded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*p~a~*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component net-radiometers (CNR1; Kipp and Zonen, Delft, Netherlands) and double-sided net radiometers (NR Lite; Kipp and Zonen, Delft, Netherlands). Double-sided net radiometers were used at the young peat site only. We also measured soil (*T~s~*) and water (*T~w~*) temperature (*T~s~*) at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). An in depth description of the measurements made at each site individually can be found in Eichelmann et al. (*in review*).

## Soil Analyses
We measured soil carbon (C), nitrogen (N), and iron (Fe) content at the young peat, old peat, and alluvium sites in the recently accreted O horizon and the underlying pre-restoration parent soils. Fifteen samples of both 0 horizon and parent soils were collected at each site across three transects with each sampling location at least 5m apart. The transects at each site were within the flux footprints of each eddy flux tower (Fig. 1), all samples were collected from fully inundated locations, and in the recently accreted O horizon was collected immediately above the underlying parent soil sample. At the young peat and alluvium sites (both < 8 years old), the differentiation between these horizons was clear, as recently accreted soil was loose, mucky, and heavily comprised of non-decomposed peat, whereas the underlying parent soils were hardpack agricultural soils. At these sites, we collected the O horizon soils by hand and collected the underlying parent soils using a sediment corer. Our sampling strategy was different at the old peat site because more than 0.5m of O horizon has accumulated since initial restoration at 20 year old site. Here, the recent O horizon has accumulated to the water surface, and we collected the top 2 cm of this horizon, representing the most recent peat. We then bore a hole through the 0.5-0.7m deep saturated peat layer to the underlying agricultural soil horizon, and collected this underlying soil horizon to represent the original parent agricultural soil. All soil samples were immediately bagged, and subsamples were extracted in 0.5M HCl and 0.2M sodium citrate-0.05M sodium ascorbate solutions to measure extractable Fe and poorly crystalline Fe, respectively.


## Data Analysis
### Wavelet and Information Theory Analysis
We used a combination of wavelet decomposition and information theory to (1) isolate major time scales of variation within the continuous CH~4~ flux time series, and (2) identify scale-emergent interactions between CH~4~ fluxes and a number of biophysical drivers. Here, flux and environmental variable time series' were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [percival1995estimation], where we isolate variation occuring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales that correspond to 2^n measurement intervals. As such, the hourly time scale was reconstructed from scales 1-2 (1-2h), diel from dyadic scales 3-6 (4h to 1.3 days), and multiday from dyadic scales 7-10 (2.7-21.3 days). This approach is explained in greater detail in @sturtevant2016identifying, and our analysis used the same approach and processing software described therein.

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information is derived from Shannon entropy (*H*), a measure of uncertainty [@shannon1998mathematical], and *I* quantifes the amount of information shared between two variables. Mutual information, and similar information theory metrics such as transfer entropy, are particularly useful for identifying relationships in complex time series because these methods do not assume linearity or other functional relationships between interacting variables [@ruddell2013applying]. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies for each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables:
$$H_x=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{x,y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Since we are working with continuous data, we bin the data into discrete states for all variables. We follow the same processing methodology in @sturtevant2016identifying for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. To calculate *H* and *I*, we then estimated probability distributions for *H* using 10 fixed-interval bins and calculated *I* for a number of time lags and leads of *X*. We calculated *I* at lags and leads up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases *F~CH4~* was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps descibed above. A more detailed description of this processing framework can be found in @sturtevant2016identifying, and all wavelet decomposition and *I* calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R*) where *I* is divided by the Shannon entropy of the Y variable, in this case *F~CH4~*. 

# Results

# References

