---
title: 'Supplemental Materials: Soil properties and sediment accretion modulate methane
  fluxes from restored wetlands'
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Daphne J Szutu^1^, Joseph
  G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document: default
  html_document: default
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{hyperref}
link-citations: yes
csl: global-change-biology.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, California, USA

^3^National Ecological Observatory Network, Battelle, Boulder, Colorado, USA

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
```

# Methods

## Eddy Covariance Instrumentation and Processing
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat were measured from the three sites by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA). High-frequency data from all sites were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval. Measurements took place at a height of 5.44 m, 4.64 m, and 5.36 m for the peat, mature, and alluvium site, respectively. Corrections were applied to vertical wind speed data prior to flux processing for sonic anemometers affected by the 'w bug', which causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto_scaling_2010; @hatala_greenhouse_2012; @knox_agricultural_2015].

We also measured a number of auxiliary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using shielded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*PA*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component and double-sided net-radiometers (CNR1 or NR Lite; Hukseflux & Kipp and Zonen, Delft, Netherlands). We measured the canopy greenness index (*GCC*) using digital cameras. We measured soil (*T~s~*) and water (*T~w~*) temperature at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). We used 10 cm above the sediment surface *T~w~* measurements at the peat and alluvium wetland, and 8 cm *T~w~* measurements for the mature wetland. All *T~s~* measurements were 8 cm below the sediment surface.

## Wavelet-Information Theory Analysis
Flux and environmental variable time series were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [@percival_estimation_1995], where we isolate variation occurring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales corresponding to 2^n^ measurement intervals. The hourly time scale was reconstructed from dyadic scales 1-2 (1-2 h), diel scale from dyadic scales 3-6 (4 h to 1.3 days), and multiday scale from dyadic scales 7-10 (2.7-21.3 days). 

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies of each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables as follows:
$$H_X=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_Y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{X,Y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Here, we bin our continuous data into discrete states to calculate *H~X~*, *H~Y~*, and *H~X,Y~*. We follow the same processing methodology in @sturtevant_identifying_2016 for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. Analyses for each year were confined to the growing season (DOY 100-300), and time series border periods were removed to avoid wavelet-induced distortions. We then calculated joint and marginal Shannon entropies using 10 fixed-interval bins for the probability distributions of each variable, and calculated *I* for a number of time lags of *X*. We calculated *I* at lags up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases, CH~4~ flux (*F~CH4~*) was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps described above. A more detailed description of this entire processing methodology can be found in @sturtevant_identifying_2016, and all wavelet decomposition and entropy calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R^*) where *I* is divided by the Shannon entropy of the *Y* variable (*F~CH4~*). 

\newpage

# Results

```{r figS1, fig.height = 8, fig.width = 7, dpi=600, fig.cap="Diel patterns of mean (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), and (c) evapotranspiration (*ET*) by year across all sites."}
source("R/Diel_patterns.R") #process the data

a <- ggplot(diel_all, aes(x=time, y=CH4, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(F[CH4]~" ("~nmol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

b <- ggplot(diel_all, aes(x=time, y=NEE, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(NEE~" ("~mu~mol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

c <- ggplot(diel_all, aes(x=time, y=ET, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(ET~" ("~mmol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);

grid.arrange(a, b, c, ncol=1)
```

```{r figS2, fig.height = 3, fig.width = 4, dpi=600, fig.cap="Alluvium wetland mean methane flux (*F~CH4~*) and atmospheric pressure across all half-hourly periods in the 2015 growing season (DOY 100-300). Color scale for each dot denotes the time of day. Peak *F~CH4~* occurs late in the afternoon during atmospheric pressure lows, and this relationship displays hysteresis."}
ggplot(data=subset(diel_all, site == "Alluvium" & year == 2015), aes(x=PA, y=CH4, color=time)) +
  geom_point(size=1.5) +
  xlab("Atmospheric pressure (kPA)") +
  ylab(expression(F[CH4]~" ("~nmol~m^{-2}~s^{-1}~")")) +
  theme_bw()


```

```{r figS3, fig.cap="Daily mean (a) photosynthetically active radiation (*PAR*), (b) air temperature (*T~a~*), (c) water temperature (*T~w~*), and (d) water table depth (*WTD*) for the alluvium, peat, and mature wetland sites. The dashed horizontal line on panel d represents the ground surface elevation at water table well locations.", fig.width=7, fig.height=5, dpi=600}

#load and process wetland sites into half-hourly, daily, and annual sums
source("R/peat19_processing.R") #19yr old 'mature' peat site
source("R/peat6_processing.R") #6yr old peat site
source("R/alluvium_processing.R") #alluvium site

#merge daily sums to single dataset
all_daily <- rbind(alluvium_daily, peat6_daily); all_daily <- rbind(all_daily, peat19_daily)
all_daily$date <- as.Date(all_daily$datetime) #data plotting variable

#themes
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#re-order factors to group peat sites in legend
all_daily$site <- factor(all_daily$site, levels=c("Alluvium", "Peat", "Mature"))

#create plot objects
a <- ggplot(subset(all_daily, year > 2014), aes(x=date, y=PAR, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(PAR~" ("~mu~mol~m^{-2}~s^{-1}~")")) + 
  annotate("text", x=as.Date("2015-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

b <- ggplot(subset(all_daily, year > 2014), aes(x=date, y=Tair, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[a]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2015-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

c <- ggplot(subset(all_daily, year > 2014), aes(x=date, y=Tw, color=site)) +
  geom_line(size=0.5) + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(T[w]~" ("~degree~C~")")) +
  annotate("text", x=as.Date("2015-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position="none")

d <- ggplot(subset(all_daily, year > 2014), aes(x=date, y=WTD, color=site)) +
  geom_line(size=0.5) + theme_bw() + 
  geom_hline(yintercept=0, linetype="dashed") +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab("WTD (cm)") +
  annotate("text", x=as.Date("2015-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  time_series + theme(legend.title=element_blank(), legend.position=c(.09, .40),
                      legend.margin = margin(0, 0, 0, 0), 
                      legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8),
                      legend.background = element_rect(fill = "white"))

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

```{r figS4, fig.cap="Daily (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), (c) gross ecosystem photosynthesis (*GEP*), and (d) ecosystem respiration (*ER*) from the alluvium, peat, and mature wetlands when measurements were available across all three sites.", fig.width=7, fig.height=5.2, dpi=600}
#themes
time_series <- theme_bw() +
  theme(panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        axis.title.x = element_blank(), axis.title.y = element_text(size=9))

#plot objects, complete years only
a <- ggplot(subset(all_daily, year > 2013), aes(x=date, y=mgCH4, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(F[CH4]~" (mg C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2018-01-01"), y=Inf, label="a", vjust=1, hjust=1) +
  time_series + theme(legend.position = "none", axis.text.x = element_blank())

b <- ggplot(subset(all_daily, year > 2013), aes(x=date, y=gCO2, color=site)) +
  geom_line() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(NEE~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2018-01-01"), y=Inf, label="b", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position = "none")

c <- ggplot(subset(all_daily, year > 2013), aes(x=date, y=gGEP, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(GEP~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2018-01-01"), y=Inf, label="c", vjust=1, hjust=1) +
  time_series + theme(axis.text.x = element_blank(), legend.position = "none")

d <- ggplot(subset(all_daily, year > 2013), aes(x=date, y=gER, color=site)) +
  geom_line() + theme_bw() +
  scale_x_date(date_breaks = "1 year", labels = date_format("%Y-%m")) +
  ylab(expression(ER~" (g C"~m^{-2}~d^{-1}~")")) + 
  annotate("text", x=as.Date("2018-01-01"), y=Inf, label="d", vjust=1, hjust=1) +
  time_series + theme(legend.title=element_blank(),
                       legend.position=c(.76, .75), 
                       legend.margin = margin(0, 0, 0, 0),
                       legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8),
                       legend.background = element_rect(fill = "transparent"))

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)
iD <- ggplotGrob(d)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5], iD$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);
iD$widths[2:5] <- as.list(maxWidth);

grid.arrange(iA, iB, iC, iD, ncol=1)
```

```{r figS5, fig.cap="Soil (a) HCl-extractable Fe, (b) ferrous iron (Fe^2+^), (c) poorly crystalline Fe oxides (Fe~ca~), (d) pH, (e) C concentration (%), and (f) N concentration (%) in accreted and parent horizons across the three wetland sites (*n* $\\geq$ 12 per depth). Solid horizontal lines are medians, boxes are the interquartile range, whiskers are the 95% confidence interval, and points denote measured outliers. Significance levels are shown for pairwise comparisons between soil horizons at each site (Welch's t-test), where ns denotes no significant differences and increasing asterisks denotes significance levels from *P* < 0.05, 0.01, 0.001, to 0.0001.", fig.width=8, fig.height=4, dpi=600}
source("R/soils.R") #process soils data into summary dataframe
#theme
boxplot_theme <- theme(axis.text.x = element_text(size=8),
                 axis.title.y = element_text(size=9),
                 axis.title.x = element_blank())

a <- ggplot(soils, aes(x=site, y=FeT_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Extractable Fe (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,32)) +
  annotate("text",x=0.6,y=32,label="(a)") +
  boxplot_theme + theme(legend.position="none")

b <- ggplot(soils, aes(x=site, y=FeII_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Fe2+ (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,25)) +
  annotate("text",x=0.6,y=25,label="(b)") +
  boxplot_theme + theme(legend.position="none")

c <- ggplot(soils, aes(x=site, y=FePC_mg.g, fill=horizon)) +
  geom_boxplot() +
  ylab(expression(Fe[ca]~" (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(0,0.5)) +
  annotate("text",x=0.6,y=0.5,label="(c)") +
  boxplot_theme + theme(legend.position="none")

d <- ggplot(soils, aes(x=site, y=pH, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  scale_y_continuous(limits=c(4.5,7.6)) +
  ylab(expression("pH")) +
  annotate("text",x=0.6,y=7.6,label="(d)") +
  boxplot_theme +
  theme(legend.position=c(.79,.2), legend.title = element_blank(), 
        legend.margin = margin(0, 0, 0, 0), 
        legend.background = element_rect(fill = "white"))

e <- ggplot(soils, aes(x=site, y=C, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression("%C")) +
  scale_y_continuous(limits=c(10,50)) +
  annotate("text",x=0.6,y=50,label="(e)") +
  boxplot_theme + theme(legend.position="none")

f <- ggplot(soils, aes(x=site, y=N, fill=horizon)) +
  geom_boxplot() +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  ylab(expression("%N")) +
  #scale_y_continuous(limits=c(10,50)) +
  #annotate("text",x=0.6,y=50,label="(e)") +
  boxplot_theme + theme(legend.position="none")

grid.arrange(a, b, c, d, e, f, ncol=3)
```

# References

