---
title: 'Supplemental Materials: Soil iron content and sediment accretion drive methane
  fluxes in restored wetlands'
author: Samuel D Chamberlain^1^, Tyler Anthony^1^, Whendee L Silver^1^, Elke Eichelmann^1^,
  Kyle S Hemes^1^, Patricia Y Oikawa^2^, Cove Sturtevant^3^, Daphne J Szutu^1^, Joseph
  G Verfaillie^1^, and Dennis D Baldocchi^1^
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document: default
  html_document: default
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{hyperref}
link-citations: yes
csl: global-change-biology.csl
bibliography: library.bib
---

^1^Department of Environmental Science, Policy, and Management, University of California, Berkeley, California, USA

^2^Department of Earth and Environmental Sciences, California State University, East Bay, California, USA

^3^National Ecological Observatory Network, Battelle, Boulder, Colorado, USA

\newpage

```{r setup, include=FALSE}
#setwd to main package directory
knitr::opts_knit$set(root.dir = normalizePath("../"))
options(digits=2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```

```{r packages, message=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(scales)
library(zoo)
library(ggpubr)
```

# Methods
## Site Descriptions and Tower Locations
Wetland ecosystem fluxes have been measured for over three years from three core wetland sites, where one site is restored on Scribner series and two sites are restored on Rindge series (Fig. 1). The Rindge series site on Sherman Island (Ameriflux site Us-Myb; N 38.0498, W 121.7650), referred to as the young peat site, was restored from a former pasture in October 2010, and ecosystem flux measurement began at this time. The Rindge series site on Twitchell Island (Ameriflux site Us-Tw1; N 38.1074, W 121.6469), referred to as the old peat site, was restored from a degraded agricultural site in 1997 and ecosystem flux measurements began in July 2012. The Scribner series site on Twitchell Island (Ameriflux site US-Tw4; N 38.10275, W 121.64125), referred to as the peat-alluvium site, was restored in January 2014 on a former corn field and ecosystem flux measurements began at this time. All sites are vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*) and are actively managed to remain inundated year-round. Both the young peat and peat-alluvium sites were constructed to have heterogeneous bathymetry, providing channels of open water and areas of closed vegetation. The old peat site is entirely closed vegetation and the ground surface is saturated tule and cattail detritus, whereas the recently restored young peat and peat-alluvium sites have standing water under the plant canopy. For a more detailed description of the young and old peat sites see @miller_carbon_2011, @matthes_parsing_2014, and @knox_agricultural_2015.

Two temporary towers were installed within the same restored wetland complex as the core peat-alluvium site, where wetland age is held constant and soil type varies. One temporary tower was installed in a peat Rindge soil (N 38.10607, W 121.64172), and the second tower was installed within the same alluvium Scribner soil (N 38.09876, W 121.64217) as the core peat-alluvium site (Fig. 1). We verified that soil properties at these additional sites were characteristic of peat versus peat-alluvium soils using a soil C dataset measured across 117 locations in this wetland complex from December 2013, prior to re-flooding (data not shown). The temporary towers and the core peat-alluvium tower are installed within the same wetland complex that was restored in January 2014. Temporary sites are also vegetated with a mix of cattails (*Typha* spp.) and tules (*Schoenoplectus acutus*), are actively managed to remain inundated year-round, and are similar in structure to the young peat and peat-alluvium core sites.

## Eddy Covariance Instrumentation and Processing
Fluxes of CH~4~, CO~2~, H~2~O, and sensible heat were measured from the three core and two temporary wetland sites by eddy covariance. We measured high-frequency three-dimensional wind speed (*u*,*v*,*w*) and temperature with sonic anemometers (WindMaster; Gill Instruments, Hampshire, UK) and CH~4~, CO~2~, and H~2~O concentrations with open-path infrared gas analyzers (LI-7500A and LI-7700; LI-COR Inc., Lincoln, Nebraska, USA). High-frequency data from the peat-alluvium, old peat, young peat, and temporary peat site were logged to LI-7550 dataloggers (LI-COR Inc., Lincoln, Nebraska, USA) at 20 Hz interval, and high-frequency data from the temporary peat-alluvium site were logged to a CR1000 datalogger (Campbell Scientific, Logan, Utah, USA) at 10 Hz interval. Measurements took place at a height of 5.44 m, 4.64 m, 5.36 m, 4.70 m, and 3.88 m for the young peat, old peat, peat-alluvium, temporary peat-alluvium, and temporary peat site, respectively. Corrections were applied to vertical wind speed data prior to flux processing for sonic anemometers affected by the 'w bug', which causes an underestimate in vertical wind speeds. Half-hourly fluxes were calculated from the covariance between vertical wind speed (*w*) and gas concentrations using in-house MATLAB software [@detto_scaling_2010; @hatala_greenhouse_2012; @knox_agricultural_2015]. Flux corrections and quality control were applied as described in detail in @knox_agricultural_2015, @chamberlain_evaluation_2017, and Eichelmann et al. (*in review*), and included high-frequency data despiking, 2-D coordinate rotations, density corrections, and site-specific friction velocity (*u$^\ast$*) filtering. At the old peat site, we reject fluxes from wind directions 290$^\circ$-240$^\circ$ because fluxes from these directions were from other wetland types; however, we did not apply wind direction filtering to the other sites where flux footprints were more homogeneous. Footprints at all sites were calculated using a two-dimensional analytical model [@detto_soil_2006; @hsieh_approximate_2000]. We then filled all gaps in the half-hourly flux time series using artificial neural networks (ANNs) as described in detail in @knox_biophysical_2016 and Eichelmann et al. (*in review*). Separate ANNs were trained for daytime and nighttime CO~2~ fluxes, and the nighttime CO~2~ flux ANN was used to model *ER* at all times. Gross ecosystem photosynthesis (*GEP*) was then estimated as the difference between the gap-filled CO~2~ flux and modeled *ER* [@baldocchi_does_2015]. More detail on this partitioning strategy can be found in @baldocchi_does_2015 and @knox_biophysical_2016, and our partitioning method performs well against independent verification methods in agricultural systems [@oikawa_revisiting_2017]. We did not gap fill or partition fluxes for the temporary tower sites and only present their observed fluxes.

We also measured a number of auxiliary atmospheric and soil environmental variables averaged at half-hourly intervals and logged to CR1000 dataloggers (Campbell Scientific, Logan, Utah, USA). Atmospheric variables included air temperature (*T~a~*) and relative humidity (*RH*) using shielded and aspirated thermistor and capacitance sensors (HMP45C or HMP60; Viasala, Vantaa, Finland), atmospheric pressure (*PA*) using capacitance sensors (PTB110; Viasala, Vantaa, Finland), photosynthetically active radiation (*PAR*) using quantum sensors (PAR-LITE or PQS 1; Kipp and Zonen, Delft, Netherlands), and net radiation (*R~net~*) using four-component and double-sided net-radiometers (CNR1 or NR Lite; Hukseflux & Kipp and Zonen, Delft, Netherlands). We measured soil (*T~s~*) and water (*T~w~*) temperature at various depths using copper constant thermocouples, and water table depth (*WTD*) using pressure transducers (CS450 or CS451; Campbell Scientific, Logan, Utah, USA). We used 10 cm above the sediment surface *T~w~* measurements at the peat-alluvium and young peat wetland, and 8 cm *T~w~* measurements for the old peat wetland. All *T~s~* measurements were 8 cm below the sediment surface. A more detailed description of the measurements made and instruments used at each core tower site can be found in Eichelmann et al. (*in review*).

## Wavelet-Information Theory Analysis
Flux and environmental variable time series were decomposed into major time scales of interest using maximal-overlap discrete wavelet transforms [@percival_estimation_1995], where we isolate variation occurring at the hourly (1-2 h), diel (4 h to 1.3 days), and multiday time scales (2.7-21.3 days). We used the Least Symmetric 8 wavelet filter to decompose the original time series for all variables, and each time scale was reconstructed from the wavelet detail at multiple dyadic scales corresponding to 2^n^ measurement intervals. The hourly time scale was reconstructed from dyadic scales 1-2 (1-2 h), diel scale from dyadic scales 3-6 (4 h to 1.3 days), and multiday scale from dyadic scales 7-10 (2.7-21.3 days). 

Once the time series were decomposed into the major scales of variation, we used mutual information (*I*) to identify relationships between variables. Mutual information between an *X* and *Y* variable can be described using the following equation:
$$I_{X,Y}=H_X+H_Y-H_{X,Y}$$
where *H~X~* and *H~Y~* are marginal Shannon entropies of each variable and *H~Y,X~* is the joint Shannon entropy of *X* and *Y*. These entropies are calculated from the marginal and joint probability distributions of the *X* and *Y* variables as follows:
$$H_X=-\sum_{x_t}p(x_t)log_2p(x_t)$$
$$H_Y=-\sum_{y_t}p(y_t)log_2p(y_t)$$

$$H_{X,Y}=-\sum_{x_t}\sum_{y_t}p(x_t,y_t)log_2p(x_t,y_t)$$
where *x~t~* and *y~t~* are states within the overall distribution of *X* and *Y*. Here, we bin our continuous data into discrete states to calculate *H~X~*, *H~Y~*, and *H~X,Y~*. We follow the same processing methodology in @sturtevant_identifying_2016 for calculating *I*. Briefly, we first wavelet decomposed variables of interest into hourly, diel, and multiday scales, and following wavelet decomposition, original data gaps were reinserted into the wavelet detail reconstructions. Analyses for each year were confined to the growing season (DOY 100-300), and time series border periods were removed to avoid wavelet-induced distortions. We then calculated joint and marginal Shannon entropies using 10 fixed-interval bins for the probability distributions of each variable, and calculated *I* for a number of time lags of *X*. We calculated *I* at lags up to 5 hours for the hourly and diel time scales, and up to 8 days for the multiday time scale. In all cases, CH~4~ flux (*F~CH4~*) was treated as the *Y* variable, as we were interested in the response of *F~CH4~* to external drivers. Statistical significance was then calculated at a 95% threshold using 1000 Monte Carlo random walks that underwent the same processing steps described above. A more detailed description of this entire processing methodology can be found in @sturtevant_identifying_2016, and all wavelet decomposition and entropy calculations were conducted using the ProcessNetwork Software (http://www.mathworks.com/matlabcentral/fileexchange/41515-processnetwork-processnetwork-software). Here, we present relative mutual information (*I^R^*) where *I* is divided by the Shannon entropy of the *Y* variable (*F~CH4~*). 

\newpage

# Results

```{r figS1, fig.height = 2.5, fig.width = 3.5, dpi=600, fig.cap="Soil percent nitrogen (%N) in accreted and parent horizons across the three core wetland sites (*n* $\\geq$ 12 per depth). Significance levels are shown for pairwise comparisons between soil horizons at each site (Welch's t-test), where ns denotes no significant differences and increasing asterisks denotes significance levels from *P* < 0.05, 0.01, 0.001, to 0.0001."}
source("R/soils.R") #process soils data into summary dataframe
boxplot_theme <- theme(axis.text.x = element_text(size=8),
                 axis.title.y = element_text(size=9),
                 axis.title.x = element_blank())

ggplot(soils, aes(x=site, y=N, fill=horizon)) +
  geom_boxplot() +
  ylab(expression("Extractable Fe (mg "~g^{-1}~"soil)")) +
  stat_compare_means(aes(group = horizon), label = "p.signif", method="t.test") +
  boxplot_theme + theme(legend.position="none")
```


```{r figS2, fig.cap="Distribution of net ecosystem exchange (*NEE*) from the three sites on Twitchell Island where wetland age is held constant and parent soil type varies. All measurements were made during the 2017 growing season between July 1 and August 15. Measurements were only used if simultaneous half-hourly fluxes were available across all three sites (n = 991). All sites were restored in 2014. The temporary peat tower measured fluxes from a Rindge peat soil series, and the temporary peat-alluvium tower measured fluxes from the same Scribner alluvium soil as the core peat-alluvium site. Both temporary towers were located within 500m of the core peat-alluvium tower.", fig.height=3.2, fig.width=3.5, dpi=600}
#load datasets
source("R/alluvium_processing.R") #peat-alluvium site
load("data/mobile_alluvium.Rdata")
load("data/mobile_peat.Rdata")

#subset all Twitchell Island four year old sites and combine to single data set
alluvium4 <- select(alluvium_all, wc, datetime) #core peat-alluvium site
alluvium4$datetime <- round_date(alluvium4$datetime, unit="minutes")
alluvium4 <- subset(alluvium4, datetime > "2017-06-30" & datetime <= "2017-08-15")

peat4 <- select(mobile_peat, wc, datetime) #mobile peat site
peat4$datetime <- round_date(peat4$datetime, unit="minutes")
peat4 <- subset(peat4, datetime > "2017-06-30" & datetime <= "2017-08-15")

alluvium2_4 <- select(mobile_alluvium, wc, datetime)#mobile alluvium site
alluvium2_4$datetime <- round_date(alluvium2_4$datetime, unit="minutes")
alluvium2_4 <- subset(alluvium2_4, datetime > "2017-06-30" & datetime <= "2017-08-15")

#merge by time and omit all points where measurements were not available at all three sites
twitchell <- merge(alluvium2_4, alluvium4, by="datetime")
twitchell <- merge(twitchell, peat4, by="datetime")
twitchell <- na.omit(twitchell)
twitchell <- gather(select(twitchell, datetime, wc.x, wc.y, wc), site, CH4, -datetime)

#fix site names
twitchell$site <- ifelse(twitchell$site == "wc.x", "Peat-Alluvium (temp)",
                    ifelse(twitchell$site == "wc.y", "Peat-Alluvium (core)",
                           "Peat (temp)"))


ggplot(data=twitchell, aes(CH4, color=site, fill=site)) +
  geom_density(alpha=0.1) +
  xlab(expression(NEE~" "~"("~nmol~CO[2]~m^{-2}~s^{-1}~")")) + 
  theme_bw() + 
  theme(legend.title=element_blank(), legend.position=c(.25, .90),
        legend.margin = margin(0, 0, 0, 0), legend.background = element_rect(fill = "transparent"),
        axis.title = element_text(size=10), 
        panel.grid.minor=element_blank(), panel.grid.major.y = element_blank(),
        legend.key.size =  unit(0.17, "in"), legend.text = element_text(size=8))
```

```{r figS3, fig.height = 8, fig.width = 7, dpi=600, fig.cap="Diel patterns of mean (a) methane flux (*F~CH4~*), (b) net ecosystem exchange (*NEE*), and (c) evapotranspiration (*ET*) by year."}
source("R/Diel_patterns.R") #process the data

a <- ggplot(diel_all, aes(x=time, y=CH4, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(F[CH4]~" ("~nmol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

b <- ggplot(diel_all, aes(x=time, y=NEE, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(NEE~" ("~mu~mol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

c <- ggplot(diel_all, aes(x=time, y=ET, color=factor(year))) +
  geom_line(size=0.5) + facet_grid(.~site) +
  ylab(expression(ET~" ("~mmol~m^{-2}~s^{-1}~")")) +
  xlab("Time of day (hr)") + labs(color='Year')  +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank())

#Multiple plots aligned
iA <- ggplotGrob(a)
iB <- ggplotGrob(b)
iC <- ggplotGrob(c)

maxWidth = grid::unit.pmax(iA$widths[2:5], iB$widths[2:5], iC$widths[2:5])
iA$widths[2:5] <- as.list(maxWidth); iB$widths[2:5] <- as.list(maxWidth); iC$widths[2:5] <- as.list(maxWidth);

grid.arrange(a, b, c, ncol=1)
```

```{r figS4, fig.height = 3, fig.width = 4, dpi=600, fig.cap="Mean methane flux (*F~CH4~*) and atmospheric pressure across all half-hourly periods in the 2015 growing season (DOY 100-300). Color scale on each dot denotes the time of day for each paired datapoint. Peak *F~CH4~* occurs late in the afternoon during atmopsheric pressure lows, and this relationship displays notable hysteresis."}
ggplot(data=subset(diel_all, site == "Peat-Alluvium" & year == 2015), aes(x=PA, y=CH4, color=time)) +
  geom_point(size=1.5) +
  xlab("Atmospheric pressure (kPA)") +
  ylab(expression(F[CH4]~" ("~nmol~m^{-2}~s^{-1}~")")) +
  theme_bw()


```

# References

